### 2024-06-18 docs/ManuscriptGrp10.qmd

---
title: "Honey Bee Colonies Impacted by Varroa, American Fouldbrood and Global Warming"
subtitle: ""
author: Edward Cruz, Jr, Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

Group Ten is pursuing the historical data collected by various public agencies to determine if the impact of Varroa, American Foulbrood, and global warming has demonstrated an increase in hive losses across the United States and Texas. The research will be accomplished using data sets derived from the agencies National Agricultural Statistics Service, Agricultural Statistics Board, and United States Department of Agriculture (USDA). Data sets containing several years of hive losses based on varroa and bacterium losses including recent possible environmental thermal global warming. A visualization of outcomes using R demonstrating and validating possible detrimental effects on honeybee colonies in the United States and Texas wrought by the negative impact of mites, bacterium, and global warming that will affect honey production and inevitably impact food production.

{{< pagebreak >}}

# Introduction

Honeybees have been nature's pollinators for centuries, with documented relationships with humans dating back to ancient Egyptians and Hindus. Historically, humans have maintained beehives, using honey as medicine in cultures such as the Egyptians, Assyrians, Chinese, Greeks, and Romans. The natural antibacterial properties of honey made it a valuable treatment for wounds, preventing infection, a practice used by Romans and Russians during World War I. Honeybees and other pollinators are critical for food production and nutritional security, yet bees face a variety of survival challenges. Currently, Varroa mites impact bee colonies and this mite infestation, a tiny red-brown parasite that can live on adult honeybees and reproduce on larvae and pupae in the developing brood. Another major threat is American Foulbrood Disease (AFB), caused by the bacterium Paenibacillus larvae. A disease that is fatal to honeybee larvae and found worldwide. The only effective control measure is to incinerate and destroy infected hives and live bees mitigating the infectious spread to other colonies. In addition, the exploration of climate change impacting honeybee colony losses has only recently been researched. While there are correlations between higher winter temperatures and greater colony losses, the effects of warmer autumn and winter temperatures on colony population dynamics and age structure as potential causes of reduced colony survival have not yet been fully investigated.Index Catalog // USDA Economics, Statistics and Market Information System. (n.d.-b). Index Catalog // USDA Economics, Statistics and Market Information System. (n.d.-a). , USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports. (n.d.).\
<https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en> <https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/>

## General Background Information

## Description of data and data source

Bee colonies maintained by beekeepers are considered livestock by the USDA due to their ability to produce honey, a consumable food item, and their essential role in assisting farmers with pollination crop seasons. Given the importance of bee colonies in agriculture, it was logical to source data from the following two authoritative websites: 1. USDA National Agricultural Statistics Service (NASS): This site provides comprehensive agricultural data, including statistics on honey production and colony health. 2. Bee Informed Partnership: This site offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations. Index Catalog // USDA Economics, Statistics and Market Information System. (n.d.-a). https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports. (n.d.). https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/

## Questions/Hypotheses to be addressed

Hypothesis: "The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

We got our data from the United States Department of Agriculture (USDA).

## Data import and cleaning

We decided to clean out our data from a few different datasets. We had to remove blank spaces and columns that were not pertinent to our analysis. We then filtered out other observations that did not directly deal with the data we are exploring. We are looking for cause of death to bee colonies and how they are affected by mites and climate change so we wanted to single out data that represented the losses so we can explore the different states by year and determine how the colonies were affected.

```{r}
library(readxl)
library(tidyverse)

```

```{r}

data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/DatabyState.csv")

View(data)
```

```{r}

# Select all columns except 3, 6, and 9
Data_Clean <- dplyr::select(data, -c(3, 6, 9))

```

```{r}

# Output cleaned data file to a csv file.
write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")

view(Data_Clean)

```

```{r}
library(readxl)
library(tidyverse)

```

```{r}

data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/DatabyState.csv")

```

```{r}

# Select all columns except 3, 6, and 9
Data_Clean <- dplyr::select(data, -c(3, 6, 9))

```

```{r}

# Output cleaned data file to a csv file.
write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")

```

```{r}

# Assuming your data frame is named "data"

# Filter for rows where "Loss" or "Deadout" is present in any column (case-insensitive)
library(stringr)  # Load stringr package for regular expressions
data_filtered <- Data_Clean[rowSums(sapply(data, grepl, pattern = c("Loss"), ignore.case = TRUE)) > 0, ]

```

```{r}

# Output cleaned data file to a csv file.
write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Filtered.csv")

```

```{r}
ggplot(data_filtered, aes( Data.Item, State.ANSI)) + geom_boxplot()
```

```{r}
ggplot(data_filtered, aes(Year)) + geom_histogram()
```

```{r}
ggplot(data_filtered, aes(x = Year)) + 
  geom_histogram(bins = 10, fill = "blue", color = "black", alpha = 0.7) +  # Adding fill color, border color, and transparency
  labs(title = "Histogram of Year", x = "Year", y = "Count") +  # Adding labels
  theme_minimal() 
```

```{r}
library(readxl)
library(tidyverse)
library(dplyr)

```

```{r}
# Output cleaned data file to a csv file.
hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)

#view(hcny_data)
```

```{r}
# Read the CSV file
hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)

```

```{r}
# Merge the first two rows to create a proper header
header <- hcny_data[1:2, ]
header <- sapply(header, function(x) paste(na.omit(x), collapse = " "))
colnames(hcny_data) <- header

# Remove the first two rows (header rows) and any completely blank rows
data_clean2 <- hcny_data[-c(1, 2), ]

# Remove rows that are completely blank
data_clean2 <- data_clean2[rowSums(is.na(data_clean2) | data_clean2 == "") != ncol(data_clean2), ]

# Convert columns to numeric where applicable (if they contain percentage values)
hcnydata_cleaned <- data_clean2 %>%
  mutate(across(.cols = -1, .fns = ~ as.numeric(gsub("[^0-9.-]", "", .))))

# Assuming hcny_data_cleaned is your cleaned data frame
colnames(hcnydata_cleaned) <- c("state", "varroa_mites", "other_pests", "disease", "pesticides", "other", "unknown")
```

```{r}
# Write the cleaned data to a new CSV file
write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcnydata_cleaned.csv", row.names = FALSE)

# View the updated data frame
view(hcnydata_cleaned)
```

```{r}
ggplot(hcnydata_cleaned, aes(varroa_mites)) + geom_histogram()
```

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2024-06-21 docs/coding-exercise/coding-exerciseECruzkml614.qmd
Leo,
3rd time is a charm, I believe that my file coding-exerciseECruzkml614.qmd can be found in my docs/coding-exercise folder

---
title: "R Coding Exercise"
author: "Edward Cruz kml614"
date: "June 18, 2024"
execute: 
  warning: false
  error: true
format:
  html:
    theme: ""
---

## Loading and Checking Data

### Import libraries

```{r}
update.packages('dslabs')
library(dslabs) # to import dataset
library(tidyverse) # data aggregation
library(dplyr) # data aggregation? 
data(gapminder)
```

### Help Function

```{r}
help("gapminder")
```

### Data Structure Overview

```{r}
str(gapminder)
```

### Data Summary

```{r}
summary(gapminder)
```

### Class Functions - Object Determination

```{r}
class(gapminder)
```

## Processing Data

### Assign only African countries to a new object/variable.

```{r}
africadata = filter(gapminder, continent == 'Africa')
head(africadata)
```

### Africa Data Structure Overview

```{r}
str(africadata)
```

### Africa Summary

```{r}
summary(africadata)
```

### African Infant Mortality Object

```{r}
infant_survival_rate = africadata %>%
  select(infant_mortality, life_expectancy)
head(infant_survival_rate)
```

### African Infant Mortality Data Structure

```{r}
str(infant_survival_rate)
```

### African Infant Mortality Data Summary

```{r}
summary(infant_survival_rate)
```

### African Mortality Object

```{r}
african_mortality = africadata %>%
  select(population, life_expectancy)
head(african_mortality)
```

### African Mortality Data Structure

```{r}
str(african_mortality)
```

### African Mortality Data Summary

```{r}
summary(african_mortality)
```

## Plotting

### Life Expectancy as a Function of Infant Mortality

```{r}
regression = lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality)
rsq <- function(x, y) summary(lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],3)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  infant_survival_rate$infant_mortality,
  infant_survival_rate$life_expectancy,
  main='African Life Expectancy \n Infant Mortality Effect',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
   col='aquamarine', 
   bg='cornflowerblue', 
   pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(170, 70, paste0(rsq_str, r_square), cex=0.75)
abline(
  lm(
    infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality),
  col='darkblue',
  lwd=1.75)
```

### Life Expectancy as a Function of Population

```{r}
regression = lm(african_mortality$life_expectancy ~ african_mortality$population)
rsq <- function(x, y) summary(lm(african_mortality$life_expectancy ~ african_mortality$population))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],9)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  african_mortality$population,
  african_mortality$life_expectancy,
  main='African Life Expectancy \n Population Effect',
  xlab='Population',
  ylab='Life Expectancy',
  col='aquamarine', 
  bg='cornflowerblue', 
  pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(155000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(155000000, 70, paste0(rsq_str, r_square), cex=0.75)
abline(
lm(
  african_mortality$life_expectancy ~ african_mortality$population),
col='darkblue',
lwd=1.75)
```

\*\* STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION\*\*

## More Data Processing

### Missing Years For Infant Mortality Data

```{r}
miss_inf_mort_rows = africadata %>%
  filter(is.na(infant_mortality))
head(miss_inf_mort_rows)
```

### Perform Year Groupby to Identify Distinct Missing Years

```{r}
miss_inf_mort_yrs = miss_inf_mort_rows %>%
  group_by(year) %>%
  count(infant_mortality)
miss_inf_mort_yrs
```

### Extract Only Year 2000 From africadata As New Object

```{r}
africadata2000 = filter(africadata, year == 2000)
africadata2000
```

### africadata2000 Data Structure

```{r}
str(africadata2000)
```

### africadata2000 Data Summary

```{r}
summary(africadata2000)
```

## More Plotting

### Create The Same Plots For africadata2000 Object

#### Filter New Object to Create Graph Datasets (1/2)

```{r}
african_mortality2000 = africadata2000 %>%
  select(population, life_expectancy)
head(african_mortality2000)
```

#### Filter New Object to Create Graph Datasets (2/2)

```{r}
infant_survival_rate2000 = africadata2000 %>%
  select(infant_mortality, life_expectancy)
head(african_mortality2000)
```

#### Infant Mortality Effect

```{r}
regression = lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality)
rsq <- function(x, y) summary(lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],3)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  infant_survival_rate2000$infant_mortality,
  infant_survival_rate2000$life_expectancy,
  main='African Life Expectancy \n Infant Mortality Effect',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
   col='aquamarine', 
   bg='cornflowerblue', 
   pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(130, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(130, 73, paste0(rsq_str, r_square), cex=0.75)
abline(
  lm(
    infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality),
  col='darkblue',
  lwd=1.75)
```

#### Population Effect

```{r}
regression = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
rsq <- function(x, y) summary(lm(african_mortality2000$life_expectancy ~ african_mortality2000$population))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],9)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  african_mortality2000$population,
  african_mortality2000$life_expectancy,
  main='African Life Expectancy \n Population Effect',
  xlab='Population',
  ylab='Life Expectancy',
  col='aquamarine', 
  bg='cornflowerblue', 
  pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(110000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(110000000, 73, paste0(rsq_str, r_square), cex=0.75)
abline(
lm(
  african_mortality2000$life_expectancy ~ african_mortality2000$population),
col='darkblue',
lwd=1.75)
```

## Simple Model Fits

### Fit Life Expectancy to Infant Mortality

```{r}
fit1 = lm(life_expectancy ~ log(infant_mortality), data=infant_survival_rate2000)
summary(fit1)
```

```{r}
# Linear regression
fit1 = lm(life_expectancy ~ infant_mortality, data=infant_survival_rate2000)
summary(fit1)

# Polynomial regression of degree 2
fit2 = lm(life_expectancy ~ poly(infant_mortality, 2), data=infant_survival_rate2000)
summary(fit2)

# Plotting
plot(
  infant_survival_rate2000$infant_mortality,
  infant_survival_rate2000$life_expectancy,
  main='Infant Mortality vs Life Expectancy',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
  col='aquamarine',
  pch=19
)

# Add linear regression line
abline(fit1, col='darkblue')

# Add polynomial regression line
new_data <- data.frame(infant_mortality = seq(min(infant_survival_rate2000$infant_mortality), max(infant_survival_rate2000$infant_mortality), length.out=100))
lines(new_data$infant_mortality, predict(fit2, new_data), col='cornflowerblue')

```

### Fit Life Expectancy to Population

```{r}
# Logarithmic regression
fit2 = lm(life_expectancy ~ log(population), data=african_mortality2000)
summary(fit2)

```

```{r}
# Linear regression
fit1 = lm(life_expectancy ~ population, data=african_mortality2000)
summary(fit1)

# Polynomial regression of degree 2
fit2 = lm(life_expectancy ~ poly(population, 2), data=african_mortality2000)
summary(fit2)

# Plotting
plot(
  african_mortality2000$population,
  african_mortality2000$life_expectancy,
  main='Population vs Life Expectancy',
  xlab='Population',
  ylab='Life Expectancy',
  col='aquamarine',
  pch=19
)

# Add linear regression line
abline(fit1, col='darkblue')

# Add polynomial regression line
new_data <- data.frame(population = seq(min(african_mortality2000$population), 
                                        max(african_mortality2000$population), length.out=100))
lines(new_data$population, predict(fit2, new_data), col='cornflowerblue')

# Add a legend to differentiate the lines
legend("topright", legend=c("Linear Fit", "Polynomial Fit (degree 2)"), col=c("darkblue", "cornflowerblue"), lwd=2)
```

#### Conclusion

### 2024-06-21 coding-exerciseECruzkml614.qmd

---
title: "R Coding Exercise"
author: "Edward Cruz kml614"
date: "June 18, 2024"
execute: 
  warning: false
  error: true
format:
  html:
    theme: ""
---

## Loading and Checking Data
### Import libraries
```{r}
update.packages('dslabs')
library(dslabs) # to import dataset
library(tidyverse) # data aggregation
library(dplyr) # data aggregation? 
data(gapminder)
```

### Help Function
```{r}
help("gapminder")
```

### Data Structure Overview
```{r}
str(gapminder)
```
### Data Summary
```{r}
summary(gapminder)
```
### Class Functions - Object Determination
```{r}
class(gapminder)
```

## Processing Data
### Assign only African countries to a new object/variable.
```{r}
africadata = filter(gapminder, continent == 'Africa')
head(africadata)
```

### Africa Data Structure Overview
```{r}
str(africadata)
```
### Africa Summary
```{r}
summary(africadata)
```
### African Infant Mortality Object
```{r}
infant_survival_rate = africadata %>%
  select(infant_mortality, life_expectancy)
head(infant_survival_rate)
```

### African Infant Mortality Data Structure
```{r}
str(infant_survival_rate)
```

### African Infant Mortality Data Summary
```{r}
summary(infant_survival_rate)
```
### African Mortality Object
```{r}
african_mortality = africadata %>%
  select(population, life_expectancy)
head(african_mortality)
```

### African Mortality Data Structure
```{r}
str(african_mortality)
```

### African Mortality Data Summary
```{r}
summary(african_mortality)
```
## Plotting
### Life Expectancy as a Function of Infant Mortality
```{r}
regression = lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality)
rsq <- function(x, y) summary(lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],3)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  infant_survival_rate$infant_mortality,
  infant_survival_rate$life_expectancy,
  main='African Life Expectancy \n Infant Mortality Effect',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
   col='aquamarine', 
   bg='cornflowerblue', 
   pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(170, 70, paste0(rsq_str, r_square), cex=0.75)
abline(
  lm(
    infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality),
  col='darkblue',
  lwd=1.75)
```

### Life Expectancy as a Function of Population
```{r}
regression = lm(african_mortality$life_expectancy ~ african_mortality$population)
rsq <- function(x, y) summary(lm(african_mortality$life_expectancy ~ african_mortality$population))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],9)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  african_mortality$population,
  african_mortality$life_expectancy,
  main='African Life Expectancy \n Population Effect',
  xlab='Population',
  ylab='Life Expectancy',
  col='aquamarine', 
  bg='cornflowerblue', 
  pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(155000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(155000000, 70, paste0(rsq_str, r_square), cex=0.75)
abline(
lm(
  african_mortality$life_expectancy ~ african_mortality$population),
col='darkblue',
lwd=1.75)
```

** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**

## More Data Processing
### Missing Years For Infant Mortality Data
```{r}
miss_inf_mort_rows = africadata %>%
  filter(is.na(infant_mortality))
head(miss_inf_mort_rows)
```

### Perform Year Groupby to Identify Distinct Missing Years
```{r}
miss_inf_mort_yrs = miss_inf_mort_rows %>%
  group_by(year) %>%
  count(infant_mortality)
miss_inf_mort_yrs
```

### Extract Only Year 2000 From africadata As New Object
```{r}
africadata2000 = filter(africadata, year == 2000)
africadata2000
```
### africadata2000 Data Structure
```{r}
str(africadata2000)
```
### africadata2000 Data Summary
```{r}
summary(africadata2000)
```

## More Plotting
### Create The Same Plots For africadata2000 Object
#### Filter New Object to Create Graph Datasets (1/2)
```{r}
african_mortality2000 = africadata2000 %>%
  select(population, life_expectancy)
head(african_mortality2000)
```
#### Filter New Object to Create Graph Datasets (2/2)
```{r}
infant_survival_rate2000 = africadata2000 %>%
  select(infant_mortality, life_expectancy)
head(african_mortality2000)
```

#### Infant Mortality Effect
```{r}
regression = lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality)
rsq <- function(x, y) summary(lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],3)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  infant_survival_rate2000$infant_mortality,
  infant_survival_rate2000$life_expectancy,
  main='African Life Expectancy \n Infant Mortality Effect',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
   col='aquamarine', 
   bg='cornflowerblue', 
   pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(130, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(130, 73, paste0(rsq_str, r_square), cex=0.75)
abline(
  lm(
    infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality),
  col='darkblue',
  lwd=1.75)
```

#### Population Effect
```{r}
regression = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
rsq <- function(x, y) summary(lm(african_mortality2000$life_expectancy ~ african_mortality2000$population))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],9)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  african_mortality2000$population,
  african_mortality2000$life_expectancy,
  main='African Life Expectancy \n Population Effect',
  xlab='Population',
  ylab='Life Expectancy',
  col='aquamarine', 
  bg='cornflowerblue', 
  pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(110000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(110000000, 73, paste0(rsq_str, r_square), cex=0.75)
abline(
lm(
  african_mortality2000$life_expectancy ~ african_mortality2000$population),
col='darkblue',
lwd=1.75)
```

## Simple Model Fits
### Fit Life Expectancy to Infant Mortality

```{r}
fit1 = lm(life_expectancy ~ log(infant_mortality), data=infant_survival_rate2000)
summary(fit3)
```

```{r}
# Linear regression
fit1 = lm(life_expectancy ~ infant_mortality, data=infant_survival_rate2000)
summary(fit1)

# Polynomial regression of degree 2
fit2 = lm(life_expectancy ~ poly(infant_mortality, 2), data=infant_survival_rate2000)
summary(fit2)

# Plotting
plot(
  infant_survival_rate2000$infant_mortality,
  infant_survival_rate2000$life_expectancy,
  main='Infant Mortality vs Life Expectancy',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
  col='aquamarine',
  pch=19
)

# Add linear regression line
abline(fit1, col='darkblue')

# Add polynomial regression line
new_data <- data.frame(infant_mortality = seq(min(infant_survival_rate2000$infant_mortality), max(infant_survival_rate2000$infant_mortality), length.out=100))
lines(new_data$infant_mortality, predict(fit2, new_data), col='cornflowerblue')

```

### Fit Life Expectancy to Population
```{r}
# Logarithmic regression
fit2 = lm(life_expectancy ~ log(population), data=african_mortality2000)
summary(fit4)

```
#### Conclusion

### 2024-06-22 docs/coding-exercise/coding-exerciseElect2016ECruzkml614.qmd
Leo,
I just sent you my 2016election code

---
title: "R Coding Exercise"
author: "Edward Cruz,Jr."
date: "today"
execute: 
  warning: false
  error: true
format:
  html:
    theme: ""
---
## Loading and Checking Data
### Import libraries
```{r}
update.packages('dslabs')
library(dslabs) # to import dataset
update.packages('tidyverse')
library(tidyverse) # data aggregation
library(dplyr) # data aggregation? 
data(polls_us_election_2016)
```

### Help Function
```{r}
help("polls_us_election_2016")
```

### Data Structure Overview
```{r}
str(polls_us_election_2016)
```

### Data Summary
```{r}
summary(polls_us_election_2016)
```
### Class Functions - Object Determination
```{r}
class(polls_us_election_2016)
```

### View data set 
```{r}
view(polls_us_election_2016)
```

## Processing Data
### Assign only U.S. to a new object/variable.
```{r}
vote_data = filter(polls_us_election_2016,  state == 'U.S.')
head(vote_data)
```

### U.S. Data Structure Overview
```{r}
str(vote_data)
```

### U.S. Summary
```{r}
summary(vote_data)
```
### Adjusted Poll Object
```{r}
adjusted_polls = vote_data %>%
  select(adjpoll_clinton,adjpoll_trump )
head(adjusted_polls)
```
### Adjusted Poll Data Structure
```{r}
str(adjusted_polls)
```
### Adjusted Poll Data Summary
```{r}
summary(adjusted_polls)
```
### Raw Poll Object
```{r}
raw_polls = vote_data %>%
  select(rawpoll_clinton,rawpoll_trump )
head(raw_polls)
```
### Raw Poll Data Structure
```{r}
str(raw_polls)
```
### Raw Poll Data Summary
```{r}
summary(raw_polls)
```
## Plotting
### Adjusted Poll as a Function for Trump & Clinton

Eddies Graphs Calculation Keep and delete the one prior
```{r}
# Perform linear regression
regression = lm(adjusted_polls$adjpoll_clinton ~ adjusted_polls$adjpoll_trump)

# Function to calculate R-squared
rsq <- function() summary(regression)$r.squared
r_square = round(rsq(), 3)

# Get regression coefficients
coeff = round(summary(regression)$coefficients[2, 1], 3)
y_int = round(summary(regression)$coefficients[1, 1], 3)

# Create plot
plot(
  adjusted_polls$adjpoll_trump,
  adjusted_polls$adjpoll_clinton,
  main='Adjusted Polls Trump & Clinton',
  xlab='Adjusted Trump Poll',
  ylab='Adjusted Clinton Poll',
  col='lightgreen',
  bg='olivedrab3',
  pch = 21
)

# Add regression equation and R-squared value to the plot
rsq_str = expression(paste("R"^2, ": "))
rsq_str = "r2: "
text(47, 55, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(47, 54, bquote(.(rsq_str) * .(r_square)), cex=0.75)

# Add regression line
abline(regression, col='darkblue', lwd=1.75)
```
### Raw Polls as a Function for Trump & Clinton
Eddies graph Keep delete prior
```{r}
# Perform linear regression
regression = lm(raw_polls$rawpoll_clinton ~ raw_polls$rawpoll_trump)

# Function to calculate R-squared
rsq <- function() summary(regression)$r.squared
r_square = round(rsq(), 3)

# Get regression coefficients
coeff = round(summary(regression)$coefficients[2, 1], 9)
y_int = round(summary(regression)$coefficients[1, 1], 3)

# Create plot
plot(
  raw_polls$rawpoll_trump,
  raw_polls$rawpoll_clinton,
  main='Raw Polls Trump & Clinton',
  xlab='Raw Trump Poll',
  ylab='Raw Clinton Poll',
  col='lightgreen',
  bg='olivedrab3',
  pch = 21
)

# Add regression equation and R-squared value to the plot
rsq_str = expression(paste("R"^2, ": "))
rsq_str = "r2: "
text(47, 55, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(47, 53, bquote(.(rsq_str) * .(r_square)), cex=0.75)

# Add regression line
abline(regression, col='darkblue', lwd=1.75)
```
#Additional Data Processing to determine any missing data or NA
#None Found
```{r}
miss_adj_polls=vote_data%>%
  filter(is.na(adjpoll_clinton))
head(miss_adj_polls)
```
#Additional Data Processing to determine any missing data or NA
#None Found 
```{r}
miss_adj_polls=vote_data%>%
  filter(is.na(adjpoll_trump))
head(miss_adj_polls)
```

```{r}
miss_raw_polls=vote_data%>%
  filter(is.na(rawpoll_clinton))
head(miss_raw_polls)
```

```{r}
miss_raw_polls=vote_data%>%
  filter(is.na(rawpoll_trump))
head(miss_raw_polls)
```

Fit Adjusted Poll for Clinton and Trump
```{r}
#Log Regression
fit1 = lm(adjpoll_clinton ~ log(adjpoll_trump), data=adjusted_polls)
summary(fit1)
```

```{r}
# Linear regression
fit1 = lm(adjpoll_clinton ~ adjpoll_trump, data=adjusted_polls)
summary(fit1)

# Polynomial regression of degree 2
fit2 = lm(adjpoll_clinton ~ poly(adjpoll_trump, 2), data=adjusted_polls)
summary(fit2)

# Plot the data
plot(
  adjusted_polls$adjpoll_trump,
  adjusted_polls$adjpoll_clinton,
  main='Adjusted Polls: Linear and Polynomial Regression',
  xlab='adjpoll_trump',
  ylab='adjpoll_clinton',
  col='lightgreen',
  bg='olivedrab3',
  pch = 21
)

# Add linear regression line
abline(fit1, col='red', lwd=2)

# Add polynomial regression line
# Generate a sequence of values for adjpoll_trump for the polynomial line
new_data <- data.frame(adjpoll_trump = seq(min(adjusted_polls$adjpoll_trump), max(adjusted_polls$adjpoll_trump), length.out=100))
# Predict the values of adjpoll_clinton for the generated sequence
poly_pred <- predict(fit2, new_data)
# Add the polynomial regression line to the plot
lines(new_data$adjpoll_trump, poly_pred, col='blue', lwd=2)

# Add a legend to differentiate the lines
legend("topright", legend=c("Linear Fit", "Polynomial Fit (degree 2)"), col=c("red", "blue"), lwd=2)
```
### Fit Raw Poll 
```{r}
#Log Regression
fit2 = lm(rawpoll_clinton ~ log(rawpoll_trump), data=vote_data)
summary(fit2)

```
```{r}
# Linear regression
fit1 = lm(rawpoll_clinton ~ rawpoll_trump, data=raw_polls)
summary(fit1)

# Polynomial regression of degree 2
fit2 = lm(rawpoll_clinton ~ poly(rawpoll_trump, 2), data=raw_polls)
summary(fit2)

# Plot the data
plot(
  raw_polls$rawpoll_trump,
  raw_polls$rawpoll_clinton,
  main='Raw Polls: Linear and Polynomial Regression',
  xlab='rawpoll_trump',
  ylab='rawpoll_clinton',
  col='lightgreen',
  bg='olivedrab3',
  pch = 21
)

# Add linear regression line
abline(fit1, col='red', lwd=2)

# Add polynomial regression line
# Generate a sequence of values for rawpoll_trump for the polynomial line
new_data <- data.frame(rawpoll_trump = seq(min(raw_polls$rawpoll_trump), max(raw_polls$rawpoll_trump), length.out=100))
# Predict the values of rawpoll_clinton for the generated sequence
poly_pred <- predict(fit2, new_data)
# Add the polynomial regression line to the plot
lines(new_data$rawpoll_trump, poly_pred, col='blue', lwd=2)

# Add a legend to differentiate the lines
legend("topright", legend=c("Linear Fit", "Polynomial Fit (degree 2)"), col=c("red", "blue"), lwd=2)
```
```

### 2024-06-22 docs/coding-exercise/coding-exerciseECruzkml614.qmd
My final updated Africa R code

@@
 ```
-
-\*\* STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION\*\*
-
+** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**
+ The graph attempts to demonstrate that relationship between Population increase and life expectancy overcoming life expectancy with a low R^2 is not dependent on life expectancy but possibly resulting from the stongest will survive and improving healthcare accissbility, GDP-gross domestic product, eduction and food source successfully allowing those Africans to survive and continue.
 ## More Data Processing
@@
 
-### Fit Life Expectancy to Infant Mortality
+********************This code added by Edward Cruz************************************************
+```{r}
+fit1 = lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality)
+summary(fit1)
+```
 
@@
 ### Fit Life Expectancy to Population
+
+```{r}
+fit2 = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
+summary(fit2)
+```
 
@@
 #### Conclusion
+The African population and life expectancy analysis shows no significant relationship between population size and life expectancy in African countries according to this linear model.
+The very low R-squared value indicates that population size explains only a tiny fraction of the variation in life expectancy.
+This suggests that other factors, such as healthcare access, economic conditions, education, and other social determinants, are likely much more important in explaining differences in life expectancy across African countries.
+TO reiterate, there is tremendous suffering and mortality with the children of the populations however the resiliency of this country has been successfully documented indicating the economics, education and healthcare is improving to continue allowing the population with access to these areas to successfully prosper and survive.

### 2024-06-22 docs/coding-exercise/coding-exerciseElect2016ECruzkml614.qmd
updated exercise 3

@@
 ```
+****************************interpret the clinton and trump raw results************************************************
+
 #Additional Data Processing to determine any missing data or NA

### 2024-06-23 coding-exercise/coding-exercise.qmd
Render to personal web site

--- 2024-06-22 docs/coding-exercise/coding-exerciseECruzkml614.qmd
@@
+---
+title: "R Coding Exercise"
+---
+
+Placeholder file for the future R coding exercise.
 ---
@@
 TO reiterate, there is tremendous suffering and mortality with the children of the populations however the resiliency of this country has been successfully documented indicating the economics, education and healthcare is improving to continue allowing the population with access to these areas to successfully prosper and survive.
+
+***  This part contributed by Leonel Salazar, IGE436 ***
+
+## Loading and Checking Data
+### Import libraries
+
+```{r}
+library(dslabs) # to import dataset
+library(tidyverse) # data aggregation
+library(dplyr) # data aggregation? 
+library(renv)
+library(caret)
+data("stars")
+```
+
+### Help Function
+```{r}
+help("stars")
+```
+
+### Data Structure Overview
+```{r}
+str(stars)
+```
+### Data Summary
+```{r}
+summary(stars)
+```
+### Class Functions - Object Determination
+```{r}
+class(stars)
+```
+
+### Star Type based on Magnitude
+```{r}
+Type_Magnitude = stars %>%
+  select(type, magnitude )
+head(Type_Magnitude)
+```
+
+### Stars Type and Magnitude Structure
+```{r}
+str(Type_Magnitude)
+```
+
+```{r}
+# Convert to factor (default order is alphabetical)
+Type_Magnitude$type <- as.factor(Type_Magnitude$type)
+
+str(Type_Magnitude)
+
+```
+
+### Stars Type and Magnitude data summary
+```{r}
+summary(Type_Magnitude)
+```
+### Stars by Type and Temp
+```{r}
+Type_Temp = stars %>%
+  select(type, temp)
+head(Type_Temp)
+```
+
+### Type Temp Data Structure
+```{r}
+Type_Temp$type <- as.factor(Type_Temp$type)
+
+# Convert to numeric
+Type_Temp$temp <- as.numeric(Type_Temp$temp)
+
+str(Type_Temp)
+
+```
+
+### Type Temp Data Summary
+```{r}
+summary(Type_Temp)
+```
+```{r}
+
+Mag_Temp = stars %>%
+  select(magnitude, temp)
+head(Mag_Temp)
+
+```
+
+### Histogram of variables
+
+```{r}
+
+hist(Type_Magnitude$magnitude)
+hist(Type_Temp$temp)
+
+```
+
+### Plotting
+### Star Magnitude as a function of Star temperature
+
+```{r}
+regression = lm(Mag_Temp$magnitude ~ Mag_Temp$temp)
+rsq <- function(x, y) summary(lm(Mag_Temp$magnitude ~ Mag_Temp$temp))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  Mag_Temp$magnitude,
+  Mag_Temp$temp,
+  main='Magnitude and temperature of star',
+  xlab='Magnitude',
+  ylab='Temperature',
+   col='lightblue', 
+   bg='black', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    Mag_Temp$magnitude ~ Mag_Temp$temp),
+  col='red',
+  lwd=1.75)
+```
+
+```{r}
+
+fit1 = lm(Mag_Temp$magnitude ~ Mag_Temp$temp)
+summary(fit1)
+
+```
+*** Based on this analysis, there's a statistically significant negative linear relationship between "temp" and "magnitude" in the "Mag_Temp" data.
+The model explains about 40% of the variance in "magnitude.
+***
+
+```{r}
+regression2 = lm(Type_Magnitude$magnitude ~ Type_Magnitude$type)
+rsq <- function(x, y) summary(lm(Type_Magnitude$magnitude ~ Type_Magnitude$type))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  Type_Magnitude$magnitude,
+  Type_Magnitude$type,
+  main='Magnitude by type of star',
+  xlab='Magnitude',
+  ylab='Type',
+   col='lightblue', 
+   bg='black', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    Type_Magnitude$magnitude ~ Type_Magnitude$type),
+  col='red',
+  lwd=1.75)
+
+```
+
+```{r}
+
+fit2 = lm(Type_Magnitude$magnitude ~ Type_Magnitude$type)
+summary(fit2)
+
+```
+*** Based on this analysis, there's a statistically significant relationship between "type" and "magnitude" in the "Type_Magnitude" data. The model explains about 73% of the variance in "magnitude ***
+
+```{r}
+
+regression3 = lm(Type_Temp$temp ~ Type_Temp$type)
+rsq <- function(x, y) summary(lm(Type_Temp$temp ~ Type_Temp$type))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  Type_Temp$temp,
+  Type_Temp$type,
+  main='Temp by type of star',
+  xlab='Temp',
+  ylab='Type',
+   col='lightblue', 
+   bg='black', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    Type_Temp$temp ~ Type_Temp$type),
+  col='red',
+  lwd=1.75)
+
+```
+
+```{r}
+
+fit3 = lm(Type_Temp$temp ~ Type_Temp$type)
+summary(fit3)
+
+```
+
+*** The very high R-squared value (0.8924) suggests a strong relationship between "type" and "temp." However, it's important to be cautious with such a high value, as it could indicate overfitting ***
+
+#### Conclusion
+
+*** It might be possible to accurately predict what type of star a star may be by knowing the temperature and magnitude of a star.
+We would need better data because we can see in the histograms that the data is not linear or very normal.
+We would have to transform the data to get more accurate and reliable results.
+***

### 2024-06-23 coding-exercise/coding-exercise.qmd
fixed Error, R Code should post on personal website

@@
 Placeholder file for the future R coding exercise.
----
-title: "R Coding Exercise"
-author: "Edward Cruz kml614"
-date: "June 18, 2024"
-execute: 
-  warning: false
-  error: true
-format:
-  html:
-    theme: ""
----
 
 ## Loading and Checking Data
-
 ### Import libraries
-
 ```{r}
-update.packages('dslabs')
+#update.packages('dslabs')
 library(dslabs) # to import dataset
@@
 ### Help Function
-
 ```{r}
@@
 ### Data Structure Overview
-
 ```{r}
@@
 ```
-
 ### Data Summary
-
 ```{r}
@@
 ```
-
 ### Class Functions - Object Determination
-
 ```{r}
@@
 ## Processing Data
-
 ### Assign only African countries to a new object/variable.
-
 ```{r}
@@
 ### Africa Data Structure Overview
-
 ```{r}
@@
 ```
-
 ### Africa Summary
-
 ```{r}
@@
 ```
-
 ### African Infant Mortality Object
-
 ```{r}
@@
 ### African Infant Mortality Data Structure
-
 ```{r}
@@
 ### African Infant Mortality Data Summary
-
 ```{r}
@@
 ```
-
 ### African Mortality Object
-
 ```{r}
@@
 ### African Mortality Data Structure
-
 ```{r}
@@
 ### African Mortality Data Summary
-
 ```{r}
@@
 ```
-
 ## Plotting
-
 ### Life Expectancy as a Function of Infant Mortality
-
 ```{r}
@@
 ### Life Expectancy as a Function of Population
-
 ```{r}
@@
 ```
+
 ** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**
- The graph attempts to demonstrate that relationship between Population increase and life expectancy overcoming life expectancy with a low R^2 is not dependent on life expectancy but possibly resulting from the stongest will survive and improving healthcare accissbility, GDP-gross domestic product, eduction and food source successfully allowing those Africans to survive and continue.
+
 ## More Data Processing
-
 ### Missing Years For Infant Mortality Data
-
 ```{r}
@@
 ### Perform Year Groupby to Identify Distinct Missing Years
-
 ```{r}
@@
 ### Extract Only Year 2000 From africadata As New Object
-
 ```{r}
@@
 ```
-
 ### africadata2000 Data Structure
-
 ```{r}
@@
 ```
-
 ### africadata2000 Data Summary
-
 ```{r}
@@
 ## More Plotting
-
 ### Create The Same Plots For africadata2000 Object
-
 #### Filter New Object to Create Graph Datasets (1/2)
-
 ```{r}
@@
 ```
-
 #### Filter New Object to Create Graph Datasets (2/2)
-
 ```{r}
@@
 #### Infant Mortality Effect
-
 ```{r}
@@
 #### Population Effect
-
 ```{r}
@@
 ## Simple Model Fits
-
-********************This code added by Edward Cruz************************************************
-```{r}
-fit1 = lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality)
-summary(fit1)
-```
+### Fit Life Expectancy to Infant Mortality
 
@@
 ### Fit Life Expectancy to Population
-
-```{r}
-fit2 = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
-summary(fit2)
-```
-
 ```{r}
@@
 ```
-
-```{r}
-# Linear regression
-fit1 = lm(life_expectancy ~ population, data=african_mortality2000)
-summary(fit1)
-
-# Polynomial regression of degree 2
-fit2 = lm(life_expectancy ~ poly(population, 2), data=african_mortality2000)
-summary(fit2)
-
-# Plotting
-plot(
-  african_mortality2000$population,
-  african_mortality2000$life_expectancy,
-  main='Population vs Life Expectancy',
-  xlab='Population',
-  ylab='Life Expectancy',
-  col='aquamarine',
-  pch=19
-)
-
-# Add linear regression line
-abline(fit1, col='darkblue')
-
-# Add polynomial regression line
-new_data <- data.frame(population = seq(min(african_mortality2000$population), 
-                                        max(african_mortality2000$population), length.out=100))
-lines(new_data$population, predict(fit2, new_data), col='cornflowerblue')
-
-# Add a legend to differentiate the lines
-legend("topright", legend=c("Linear Fit", "Polynomial Fit (degree 2)"), col=c("darkblue", "cornflowerblue"), lwd=2)
-```
-
-#### Conclusion
-The African population and life expectancy analysis shows no significant relationship between population size and life expectancy in African countries according to this linear model.
-The very low R-squared value indicates that population size explains only a tiny fraction of the variation in life expectancy.
-This suggests that other factors, such as healthcare access, economic conditions, education, and other social determinants, are likely much more important in explaining differences in life expectancy across African countries.
-TO reiterate, there is tremendous suffering and mortality with the children of the populations however the resiliency of this country has been successfully documented indicating the economics, education and healthcare is improving to continue allowing the population with access to these areas to successfully prosper and survive.
 

### 2024-06-23 coding-exerciseECruzkml614_leo.qmd
fixed Error, R Code should post on personal website

--- 2024-06-23 coding-exercise/coding-exercise.qmd
@@
 title: "R Coding Exercise"
+author: "Edward Cruz kml614"
+date: "June 18, 2024"
+execute: 
+  warning: false
+  error: true
+format:
+  html:
+    theme: ""
 ---
-
-Placeholder file for the future R coding exercise.
 
@@
 ```{r}
-#update.packages('dslabs')
+update.packages('dslabs')
 library(dslabs) # to import dataset

### 2024-06-23 coding-exercise/coding-exercise.qmd
Secondary Update missing conslucions from Eddie updated 06.22.24 8:46pm

@@
 ** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**
-
+ The graph attempts to demonstrate that relationship between Population increase and life expectancy overcoming life expectancy with a low R^2 is not dependent on life expectancy but possibly resulting from the stongest will survive and improving healthcare accissbility, GDP-gross domestic product, eduction and food source successfully allowing those Africans to survive and continue.
 ## More Data Processing
@@
 ```{r}
+fit2 = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
+summary(fit2)
+```
+```{r}
+# Logarithmic regression
+fit2 = lm(life_expectancy ~ log(population), data=african_mortality2000)
+summary(fit2)
+
+```
+
+```{r}
 regression = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
@@
 ```
+#### Conclusion
+The African population and life expectancy analysis shows no significant relationship between population size and life expectancy in African countries according to this linear model.
+The very low R-squared value indicates that population size explains only a tiny fraction of the variation in life expectancy.
+This suggests that other factors, such as healthcare access, economic conditions, education, and other social determinants, are likely much more important in explaining differences in life expectancy across African countries.
+TO reiterate, there is tremendous suffering and mortality with the children of the populations however the resiliency of this country has been successfully documented indicating the economics, education and healthcare is improving to continue allowing the population with access to these areas to successfully prosper and survive.
 

### 2024-06-24 docs/ManuscriptGrp10.qmd
Docs Folder  Cleaned up: All R code and Data Sets tested and verified 06.24.24

@@
 
+---
+title: "loading/cleaning/wrangling/exploring"
+author: "Edward Cruz"
+format: docx
+---
+
 ```{r}
 library(readxl)
+update.packages("tidyverse")
 library(tidyverse)
+update.packages("dplyr")
 library(dplyr)
-
+update.packages('mgcv')
+library(mgcv)
 ```
@@
 hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)
-
-#view(hcny_data)
 ```
@@
 ```{r}
-# Read the CSV file
-hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)
+head(hcny_data)
+```
 
+```{r}
+str(hcny_data)
+```
+
+```{r}
+summary(hcny_data)
+```
+
+```{r}
+view(hcny_data)
 ```
@@
 # Merge the first two rows to create a proper header
-header <- hcny_data[1:2, ]
+header <- hcny_data[1:4, ]
 header <- sapply(header, function(x) paste(na.omit(x), collapse = " "))
 colnames(hcny_data) <- header
+head(header)
+```
 
+```{r}
 # Remove the first two rows (header rows) and any completely blank rows
-data_clean2 <- hcny_data[-c(1, 2), ]
+hcnydata_cleaned <- hcny_data[-c(1,2), ]
+head(hcnydata_cleaned)
+```
 
-# Remove rows that are completely blank
-data_clean2 <- data_clean2[rowSums(is.na(data_clean2) | data_clean2 == "") != ncol(data_clean2), ]
+```{r}
+# Assuming hcny_data_cleaned is your cleaned data frame
+colnames(hcnydata_cleaned) <- c("State", "Varroa_Mites", "Other_Parasites_1/", "Disease_2/", "Pesticides", "Other_3/", "Unknown")
+```
 
-# Convert columns to numeric where applicable (if they contain percentage values)
-hcnydata_cleaned <- data_clean2 %>%
-  mutate(across(.cols = -1, .fns = ~ as.numeric(gsub("[^0-9.-]", "", .))))
+```{r}
+head(hcnydata_cleaned)
+```
 
-# Assuming hcny_data_cleaned is your cleaned data frame
-colnames(hcnydata_cleaned) <- c("state", "varroa_mites", "other_pests", "disease", "pesticides", "other", "unknown")
+```{r}
+# Convert specified columns to numeric
+hcnydata_cleaned <- hcnydata_cleaned %>%
+  mutate(across(c("Varroa_Mites", "Other_Parasites_1/", "Disease_2/", "Pesticides", "Other_3/", "Unknown"), as.numeric))
+```
+
+```{r}
+# View the updated data frame
+hcnydata_cleaned<-hcnydata_cleaned[-1,]
+```
+
+```{r}
+head(hcnydata_cleaned)
+```
+
+```{r}
+# Verify the changes
+str(hcnydata_cleaned)
+```
+
+```{r}
+head(hcnydata_cleaned)
+```
+
+```{r}
+view(hcnydata_cleaned)
+```
+
+```{r}
+ggplot(hcnydata_cleaned, aes(Varroa_Mites)) + geom_histogram()
+```
+
+```{r}
+ggplot(hcnydata_cleaned, aes(x = Varroa_Mites)) +
+  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
+  labs(title = "Histogram of Varroa Mites",
+       x = "Varroa Mites",
+       y = "Frequency")
+```
+
+```{r}
+# Example scatter plot of two variables
+ggplot(hcnydata_cleaned, aes(x = Varroa_Mites, y = "Desease_2/")) +
+  geom_point() +
+  labs(title = "Scatter Plot of Varroa Mites vs Disease",
+       x = "Varroa Mites",
+       y = "2 American and European Foul Brood")
+```
+
+```{r}
+view(hcnydata_cleaned)
 ```
@@
 write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcnydata_cleaned.csv", row.names = FALSE)
-
-# View the updated data frame
-view(hcnydata_cleaned)
 ```
 
-```{r}
-ggplot(hcnydata_cleaned, aes(varroa_mites)) + geom_histogram()
-```
+\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*Conclusion- all code below is test code right now\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*
 

### 2024-06-27 Data-Exercise/data-exercise.qmd
Eddies Data Exercise for synthetic data 06.27.24

---
title: "Data Exercise"
---

Placeholder file for the future data exercise.
## Running Code
```{r}
library(MASS)
library(synthpop)
library(tidyverse)
library(dplyr)
library(writexl)
library(ggplot2)
library(broom)
```

```{r}
id <- paste0("Applicant #", 1:320)
```

```{r}
# Assign Management Experience Randomly
set.seed(0717)
management_random <- rbinom(320, 1, 0.5)
management <- ifelse(management_random == 0, "manager", "non-manager")
```

```{r}
set.seed(0717)
coding_viz <- rnorm(320, 85, 5)
coding_viz <- pmin(coding_viz, 100)  # Caps scores at 100
coding_viz <- round(coding_viz, 2)  # Rounds scores to two decimal places
```

```{r}
cor_var_means <- c(6.8, 7.2, 8.4, 77, 84, 80)
cor_var_matrix <- matrix(
  c(
    0.87, 0.6, 0.7, 0.36, 1.55, 0.57,
    0.6, 1.2, 0.52, 0.5, 1.2, 2.34,
    0.7, 0.52, 0.68, 0.45, 0.89, 0.75,
    0.36, 0.5, 0.45, 15.2, 1.09, 1.64,
    1.55, 1.2, 0.89, 1.09, 17.2, 1.88,
    0.57, 2.34, 0.75, 1.64, 1.88, 9.3
  ), byrow = T, nrow = 6
)
set.seed(0717)
correlated_vars_df <- as.data.frame(mvrnorm(n = 320, mu = cor_var_means, Sigma = cor_var_matrix))
```

```{r}
# Define the cap_score function
cap_score <- function(x, max_value) {
  pmin(x, max_value)
}
```

```{r}
# Assuming correlated_vars_df is already defined
correlated_vars_df_cols <- c("interview_p1", "interview_p2", "interview_p3", "coding_cleaning", "coding_ml", "performance")
colnames(correlated_vars_df) <- correlated_vars_df_cols

correlated_vars_df <- correlated_vars_df %>%
  mutate(
    interview_p1 = round(cap_score(interview_p1, 10), 1),
    interview_p2 = round(cap_score(interview_p2, 10), 1),
    interview_p3 = round(cap_score(interview_p3, 10), 1),
    coding_cleaning = round(cap_score(coding_cleaning, 100), 2),
    coding_ml = round(cap_score(coding_ml, 100), 2),
    performance = round(cap_score(performance, 100))
  )
```

```{r}

# Define the absences variable
set.seed(0717)
absences <- rpois(320, lambda = 1)

# Define the coding_viz variable
set.seed(0717)
coding_viz <- rnorm(320, 85, 5)
coding_viz <- pmin(coding_viz, 100)
coding_viz <- round(coding_viz, 2)

```

```{r}
# Define the situational variable
set.seed(0717)
situational <- round(runif(320, 60, 100))
```

```{r}
# Assuming correlated_vars_df is already defined and processed
# If not, please define it similarly to other variables

# Example definition of correlated_vars_df
# Make sure correlated_vars_df has 320 rows
correlated_vars_df <- data.frame(
  interview_p1 = rnorm(320, 7, 2),
  interview_p2 = rnorm(320, 7, 2),
  interview_p3 = rnorm(320, 7, 2),
  coding_cleaning = rnorm(320, 90, 5),
  coding_ml = rnorm(320, 90, 5),
  performance = rnorm(320, 85, 10)
)
```

```{r}
# Cap the scores as in previous steps
cap_score <- function(x, max_value) {
  pmin(x, max_value)
}

correlated_vars_df <- correlated_vars_df %>%
  mutate(
    interview_p1 = round(cap_score(interview_p1, 10), 1),
    interview_p2 = round(cap_score(interview_p2, 10), 1),
    interview_p3 = round(cap_score(interview_p3, 10), 1),
    coding_cleaning = round(cap_score(coding_cleaning, 100), 2),
    coding_ml = round(cap_score(coding_ml, 100), 2),
    performance = round(cap_score(performance, 100))
  )

# Assuming id and management are defined
# Example definition
id <- 1:320
management <- sample(70:100, 320, replace = TRUE)

# Combine into applicant_scores data frame
applicant_scores <- cbind(
  id, management, situational, coding_viz, correlated_vars_df, absences
)

# Subset the first 300 rows
applicant_final <- applicant_scores[1:300, ]

```

```{r}
# Write to Excel file
write_xlsx(as.data.frame(applicant_final), "Employee_Selection_Scores.xlsx")
```

```{r}
# Histogram of situational scores
ggplot(applicant_final, aes(x = situational)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Histogram of Situational Scores", x = "Situational Score", y = "Frequency")
```
```{r}
# Generalized linear model plot
ggplot(applicant_final, aes(x = situational, y = performance)) +
  geom_point() +
  geom_smooth(method = "glm", se = FALSE, color = "green") +
  labs(title = "Generalized Linear Model: Situational vs Performance", x = "Situational Score", y = "Performance Score")

```

```{r}
# Linear model plot
ggplot(applicant_final, aes(x = situational, y = performance)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Linear Model: Situational vs Performance", x = "Situational Score", y = "Performance Score")
```
```{r}
# Generalized linear model plot
ggplot(applicant_final, aes(x = situational, y = performance)) +
  geom_point() +
  geom_smooth(method = "glm", se = FALSE, color = "green") +
  labs(title = "Generalized Linear Model: Situational vs Performance", x = "Situational Score", y = "Performance Score")
```

```{r}
# Scatter plot
ggplot(applicant_final, aes(x = situational, y = performance)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot: Situational vs Performance", x = "Situational Score", y = "Performance Score")
```

*******Conclusion**************************************************

### 2024-06-27 docs/coding-exercise/coding-exercise.qmd
Eddies Data Exercise for synthetic data 06.27.24

---
title: "R Coding Exercise"
author: ""
date: "today"
execute: 
  warning: false
  error: true
format:
  html:
    theme: ""
---

## Loading and Checking Data
### Import libraries
```{r}
library(dslabs) # to import dataset
update.packages('tidyverse')
library(tidyverse) # data aggregation
library(dplyr) # data aggregation? 
data(gapminder)
```

### Help Function
```{r}
help("gapminder")
```

### Data Structure Overview
```{r}
str(gapminder)
```
### Data Summary
```{r}
summary(gapminder)
```
### Class Functions - Object Determination
```{r}
class(gapminder)
```

## Processing Data
### Assign only African countries to a new object/variable.
```{r}
africadata = filter(gapminder, continent == 'Africa')
head(africadata)
```

### Africa Data Structure Overview
```{r}
str(africadata)
```
### Africa Summary
```{r}
summary(africadata)
```
### African Infant Mortality Object
```{r}
infant_survival_rate = africadata %>%
  select(infant_mortality, life_expectancy)
head(infant_survival_rate)
```

### African Infant Mortality Data Structure
```{r}
str(infant_survival_rate)
```

### African Infant Mortality Data Summary
```{r}
summary(infant_survival_rate)
```
### African Mortality Object
```{r}
african_mortality = africadata %>%
  select(population, life_expectancy)
head(african_mortality)
```

### African Mortality Data Structure
```{r}
str(african_mortality)
```

### African Mortality Data Summary
```{r}
summary(african_mortality)
```
## Plotting
### Life Expectancy as a Function of Infant Mortality
```{r}
regression = lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality)
rsq <- function(x, y) summary(lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],3)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  infant_survival_rate$infant_mortality,
  infant_survival_rate$life_expectancy,
  main='African Life Expectancy \n Infant Mortality Effect',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
   col='blue', 
   bg='lightblue', 
   pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(170, 70, paste0(rsq_str, r_square), cex=0.75)
abline(
  lm(
    infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality),
  col='black',
  lwd=1.75)
```

### Life Expectancy as a Function of Population
```{r}
regression = lm(african_mortality$life_expectancy ~ african_mortality$population)
rsq <- function(x, y) summary(lm(african_mortality$life_expectancy ~ african_mortality$population))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],9)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  african_mortality$population,
  african_mortality$life_expectancy,
  main='African Life Expectancy \n Population Effect',
  xlab='Population',
  ylab='Life Expectancy',
  col='blue', 
  bg='lightblue', 
  pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(155000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(155000000, 70, paste0(rsq_str, r_square), cex=0.75)
abline(
lm(
  african_mortality$life_expectancy ~ african_mortality$population),
col='black',
lwd=1.75)
```

** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**

## More Data Processing
### Missing Years For Infant Mortality Data
```{r}
miss_inf_mort_rows = africadata %>%
  filter(is.na(infant_mortality))
head(miss_inf_mort_rows)
```

### Perform Year Groupby to Identify Distinct Missing Years
```{r}
miss_inf_mort_yrs = miss_inf_mort_rows %>%
  group_by(year) %>%
  count(infant_mortality)
miss_inf_mort_yrs
```

### Extract Only Year 2000 From africadata As New Object
```{r}
africadata2000 = filter(africadata, year == 2000)
africadata2000
```
### africadata2000 Data Structure
```{r}
str(africadata2000)
```
### africadata2000 Data Summary
```{r}
summary(africadata2000)
```

## More Plotting
### Create The Same Plots For africadata2000 Object
#### Filter New Object to Create Graph Datasets (1/2)
```{r}
african_mortality2000 = africadata2000 %>%
  select(population, life_expectancy)
head(african_mortality2000)
```
#### Filter New Object to Create Graph Datasets (2/2)
```{r}
infant_survival_rate2000 = africadata2000 %>%
  select(infant_mortality, life_expectancy)
head(african_mortality2000)
```

#### Infant Mortality Effect
```{r}
regression = lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality)
rsq <- function(x, y) summary(lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],3)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  infant_survival_rate2000$infant_mortality,
  infant_survival_rate2000$life_expectancy,
  main='African Life Expectancy \n Infant Mortality Effect',
  xlab='Infant Mortality',
  ylab='Life Expectancy',
   col='blue', 
   bg='lightblue', 
   pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(130, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(130, 73, paste0(rsq_str, r_square), cex=0.75)
abline(
  lm(
    infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality),
  col='black',
  lwd=1.75)
```

#### Population Effect
```{r}
regression = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
rsq <- function(x, y) summary(lm(african_mortality2000$life_expectancy ~ african_mortality2000$population))$r.squared
r_square = round(rsq(obs, mod), 3)
# summary(regression)$coefficients - keep to sanity check
coeff = round(summary(regression)$coefficients[2],9)
y_int = round(summary(regression)$coefficients[1],3)
plot(
  african_mortality2000$population,
  african_mortality2000$life_expectancy,
  main='African Life Expectancy \n Population Effect',
  xlab='Population',
  ylab='Life Expectancy',
  col='blue', 
  bg='lightblue', 
  pch = 21,
)
# rsq_str = expression(paste("r"^"2:"))
rsq_str = "r2: "
text(110000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
text(110000000, 73, paste0(rsq_str, r_square), cex=0.75)
abline(
lm(
  african_mortality2000$life_expectancy ~ african_mortality2000$population),
col='black',
lwd=1.75)
```

## Simple Model Fits
### Fit Life Expectancy to Infant Mortality
```{r}
fit1 = lm(infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality)
summary(fit1)
```
### Fit Life Expectancy to Population
```{r}
fit2 = lm(african_mortality2000$life_expectancy ~ african_mortality2000$population)
summary(fit2)
```
#### Conclusion

### 2024-06-27 data-exercise/data-exercise.qmd
3rd Time is a charm, I have removed and corrected misspelling of my  .qmd file and rendered it looks fine.

--- 2024-06-27 Data-Exercise/data-exercise.qmd
@@
 Placeholder file for the future data exercise.
-## Running Code
+
+## Loading and Checking Data
+## Import Libraries
 ```{r}
@@
 ```
+
 ```{r}

### 2024-06-27 data-exercise/coding-exercise.qmd
4th Attempt to push folder to update personal html 06.27.24 time10:51am

--- 2024-06-27 data-exercise/data-exercise.qmd
@@
 ## Loading and Checking Data
-## Import Libraries
+### Import libraries
 ```{r}
@@
 
-*******Conclusion**************************************************
+\*\*\*\*\*\*\*Conclusion\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

### 2024-06-27 Misc/data-exercise_Ecruz.qmd
Personal Website should have been updated 06.27.24 ECruz Time: 14.28pm

--- 2024-06-27 data-exercise/coding-exercise.qmd
@@
 library(ggplot2)
-library(broom)
 ```
@@
 
-\*\*\*\*\*\*\*Conclusion\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*
+\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\Conclusion*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

### 2024-06-27 data-exercise/data-exercise.qmd
Personal Website should have been updated 06.27.24 ECruz Time: 14.28pm

--- 2024-06-27 Data-Exercise/data-exercise.qmd
@@
 ---
-title: "Data Exercise"
+title: "Date Exercise"
 ---
@@
 Placeholder file for the future data exercise.
-## Running Code
 ```{r}
-library(MASS)
-library(synthpop)
-library(tidyverse)
-library(dplyr)
-library(writexl)
-library(ggplot2)
-library(broom)
+#update.packages('dslabs')
+library(dslabs) # to import dataset
+library(tidyverse) # data aggregation
+library(dplyr) # data aggregation? 
+data(gapminder)
 ```
 
+### Help Function
 ```{r}
-id <- paste0("Applicant #", 1:320)
+help("gapminder")
 ```
 
+### Data Structure Overview
 ```{r}
-# Assign Management Experience Randomly
-set.seed(0717)
-management_random <- rbinom(320, 1, 0.5)
-management <- ifelse(management_random == 0, "manager", "non-manager")
+str(gapminder)
+```
+### Data Summary
+```{r}
+summary(gapminder)
+```
+### Class Functions - Object Determination
+```{r}
+class(gapminder)
 ```
 
+## Processing Data
+### Assign only African countries to a new object/variable.
 ```{r}
-set.seed(0717)
-coding_viz <- rnorm(320, 85, 5)
-coding_viz <- pmin(coding_viz, 100)  # Caps scores at 100
-coding_viz <- round(coding_viz, 2)  # Rounds scores to two decimal places
+africadata = filter(gapminder, continent == 'Africa')
+head(africadata)
 ```
 
+### Africa Data Structure Overview
 ```{r}
-cor_var_means <- c(6.8, 7.2, 8.4, 77, 84, 80)
-cor_var_matrix <- matrix(
-  c(
-    0.87, 0.6, 0.7, 0.36, 1.55, 0.57,
-    0.6, 1.2, 0.52, 0.5, 1.2, 2.34,
-    0.7, 0.52, 0.68, 0.45, 0.89, 0.75,
-    0.36, 0.5, 0.45, 15.2, 1.09, 1.64,
-    1.55, 1.2, 0.89, 1.09, 17.2, 1.88,
-    0.57, 2.34, 0.75, 1.64, 1.88, 9.3
-  ), byrow = T, nrow = 6
-)
-set.seed(0717)
-correlated_vars_df <- as.data.frame(mvrnorm(n = 320, mu = cor_var_means, Sigma = cor_var_matrix))
+str(africadata)
+```
+### Africa Summary
+```{r}
+summary(africadata)
+```
+### African Infant Mortality Object
+```{r}
+infant_survival_rate = africadata %>%
+  select(infant_mortality, life_expectancy)
+head(infant_survival_rate)
 ```
 
+### African Infant Mortality Data Structure
 ```{r}
-# Define the cap_score function
-cap_score <- function(x, max_value) {
-  pmin(x, max_value)
-}
+str(infant_survival_rate)
 ```
 
+### African Infant Mortality Data Summary
 ```{r}
-# Assuming correlated_vars_df is already defined
-correlated_vars_df_cols <- c("interview_p1", "interview_p2", "interview_p3", "coding_cleaning", "coding_ml", "performance")
-colnames(correlated_vars_df) <- correlated_vars_df_cols
-
-correlated_vars_df <- correlated_vars_df %>%
-  mutate(
-    interview_p1 = round(cap_score(interview_p1, 10), 1),
-    interview_p2 = round(cap_score(interview_p2, 10), 1),
-    interview_p3 = round(cap_score(interview_p3, 10), 1),
-    coding_cleaning = round(cap_score(coding_cleaning, 100), 2),
-    coding_ml = round(cap_score(coding_ml, 100), 2),
-    performance = round(cap_score(performance, 100))
-  )
+summary(infant_survival_rate)
+```
+### African Mortality Object
+```{r}
+african_mortality = africadata %>%
+  select(population, life_expectancy)
+head(african_mortality)
 ```
 
+### African Mortality Data Structure
 ```{r}
-
-# Define the absences variable
-set.seed(0717)
-absences <- rpois(320, lambda = 1)
-
-# Define the coding_viz variable
-set.seed(0717)
-coding_viz <- rnorm(320, 85, 5)
-coding_viz <- pmin(coding_viz, 100)
-coding_viz <- round(coding_viz, 2)
-
+str(african_mortality)
 ```
 
+### African Mortality Data Summary
 ```{r}
-# Define the situational variable
-set.seed(0717)
-situational <- round(runif(320, 60, 100))
+summary(african_mortality)
+```
+## Plotting
+### Life Expectancy as a Function of Infant Mortality
+```{r}
+regression = lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality)
+rsq <- function(x, y) summary(lm(infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  infant_survival_rate$infant_mortality,
+  infant_survival_rate$life_expectancy,
+  main='African Life Expectancy \n Infant Mortality Effect',
+  xlab='Infant Mortality',
+  ylab='Life Expectancy',
+   col='aquamarine', 
+   bg='cornflowerblue', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality),
+  col='darkblue',
+  lwd=1.75)
 ```
 
+### Life Expectancy as a Function of Population
 ```{r}
-# Assuming correlated_vars_df is already defined and processed
-# If not, please define it similarly to other variables
-
-# Example definition of correlated_vars_df
-# Make sure correlated_vars_df has 320 rows
-correlated_vars_df <- data.frame(
-  interview_p1 = rnorm(320, 7, 2),
-  interview_p2 = rnorm(320, 7, 2),
-  interview_p3 = rnorm(320, 7, 2),
-  coding_cleaning = rnorm(320, 90, 5),
-  coding_ml = rnorm(320, 90, 5),
-  performance = rnorm(320, 85, 10)
+regression = lm(african_mortality$life_expectancy ~ african_mortality$population)
+rsq <- function(x, y) summary(lm(african_mortality$life_expectancy ~ african_mortality$population))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],9)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  african_mortality$population,
+  african_mortality$life_expectancy,
+  main='African Life Expectancy \n Population Effect',
+  xlab='Population',
+  ylab='Life Expectancy',
+  col='aquamarine', 
+  bg='cornflowerblue', 
+  pch = 21,
 )
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(155000000, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(155000000, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+lm(
+  african_mortality$life_expectancy ~ african_mortality$population),
+col='darkblue',
+lwd=1.75)
 ```
-
-```{r}
-# Cap the scores as in previous steps
-cap_score <- function(x, max_value) {
-  pmin(x, max_value)
-}
-
-correlated_vars_df <- correlated_vars_df %>%
-  mutate(
-    interview_p1 = round(cap_score(interview_p1, 10), 1),
-    interview_p2 = round(cap_score(interview_p2, 10), 1),
-    interview_p3 = round(cap_score(interview_p3, 10), 1),
-    coding_cleaning = round(cap_score(coding_cleaning, 100), 2),
-    coding_ml = round(cap_score(coding_ml, 100), 2),
-    performance = round(cap_score(performance, 100))
-  )
-
-# Assuming id and management are defined
-# Example definition
-id <- 1:320
-management <- sample(70:100, 320, replace = TRUE)
-
-# Combine into applicant_scores data frame
-applicant_scores <- cbind(
-  id, management, situational, coding_viz, correlated_vars_df, absences
-)
-
-# Subset the first 300 rows
-applicant_final <- applicant_scores[1:300, ]
-
-```
-
-```{r}
-# Write to Excel file
-write_xlsx(as.data.frame(applicant_final), "Employee_Selection_Scores.xlsx")
-```
-
-```{r}
-# Histogram of situational scores
-ggplot(applicant_final, aes(x = situational)) +
-  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
-  labs(title = "Histogram of Situational Scores", x = "Situational Score", y = "Frequency")
-```
-```{r}
-# Generalized linear model plot
-ggplot(applicant_final, aes(x = situational, y = performance)) +
-  geom_point() +
-  geom_smooth(method = "glm", se = FALSE, color = "green") +
-  labs(title = "Generalized Linear Model: Situational vs Performance", x = "Situational Score", y = "Performance Score")
-
-```
-
-```{r}
-# Linear model plot
-ggplot(applicant_final, aes(x = situational, y = performance)) +
-  geom_point() +
-  geom_smooth(method = "lm", se = FALSE, color = "red") +
-  labs(title = "Linear Model: Situational vs Performance", x = "Situational Score", y = "Performance Score")
-```
-```{r}
-# Generalized linear model plot
-ggplot(applicant_final, aes(x = situational, y = performance)) +
-  geom_point() +
-  geom_smooth(method = "glm", se = FALSE, color = "green") +
-  labs(title = "Generalized Linear Model: Situational vs Performance", x = "Situational Score", y = "Performance Score")
-```
-
-```{r}
-# Scatter plot
-ggplot(applicant_final, aes(x = situational, y = performance)) +
-  geom_point(color = "blue") +
-  labs(title = "Scatter Plot: Situational vs Performance", x = "Situational Score", y = "Performance Score")
-```
-
-*******Conclusion**************************************************

### 2024-06-27 data-exercise/data-exercise.qmd
Zane is helping me

--- 2024-06-27 data-exercise/data-exercise.qmd
@@
 
-## Loading and Checking Data
-## Import Libraries
 ```{r}
@@
 # Write to Excel file
-write_xlsx(as.data.frame(applicant_final), "Employee_Selection_Scores.xlsx")
+#write_xlsx(as.data.frame(applicant_final), "Employee_Selection_Scores.xlsx")
 ```
@@
 ```
-
-*******Conclusion**************************************************

### 2024-06-27 docs/coding-exercise/coding-exercise.qmd
online exactly the same as my local

@@
 title: "R Coding Exercise"
-author: ""
-date: "today"
+author: "Edward Cruz kml614"
+date: "June 18, 2024"
 execute: 
@@
 ## Loading and Checking Data
+
 ### Import libraries
+
 ```{r}
+update.packages('dslabs')
 library(dslabs) # to import dataset
-update.packages('tidyverse')
 library(tidyverse) # data aggregation
@@
 ### Help Function
+
 ```{r}
@@
 ### Data Structure Overview
+
 ```{r}
@@
 ```
+
 ### Data Summary
+
 ```{r}
@@
 ```
+
 ### Class Functions - Object Determination
+
 ```{r}
@@
 ## Processing Data
+
 ### Assign only African countries to a new object/variable.
+
 ```{r}
@@
 ### Africa Data Structure Overview
+
 ```{r}
@@
 ```
+
 ### Africa Summary
+
 ```{r}
@@
 ```
+
 ### African Infant Mortality Object
+
 ```{r}
@@
 ### African Infant Mortality Data Structure
+
 ```{r}
@@
 ### African Infant Mortality Data Summary
+
 ```{r}
@@
 ```
+
 ### African Mortality Object
+
 ```{r}
@@
 ### African Mortality Data Structure
+
 ```{r}
@@
 ### African Mortality Data Summary
+
 ```{r}
@@
 ```
+
 ## Plotting
+
 ### Life Expectancy as a Function of Infant Mortality
+
 ```{r}
@@
   ylab='Life Expectancy',
-   col='blue', 
-   bg='lightblue', 
+   col='aquamarine', 
+   bg='cornflowerblue', 
    pch = 21,
@@
     infant_survival_rate$life_expectancy ~ infant_survival_rate$infant_mortality),
-  col='black',
+  col='darkblue',
   lwd=1.75)
@@
 ### Life Expectancy as a Function of Population
+
 ```{r}
@@
   ylab='Life Expectancy',
-  col='blue', 
-  bg='lightblue', 
+  col='aquamarine', 
+  bg='cornflowerblue', 
   pch = 21,
@@
   african_mortality$life_expectancy ~ african_mortality$population),
-col='black',
+col='darkblue',
 lwd=1.75)
 ```
+** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**
+ The graph attempts to demonstrate that relationship between Population increase and life expectancy overcoming life expectancy with a low R^2 is not dependent on life expectancy but possibly resulting from the stongest will survive and improving healthcare accissbility, GDP-gross domestic product, eduction and food source successfully allowing those Africans to survive and continue.
+## More Data Processing
 
-** STILL NEED TO ANSWER QUESTION AT THE END OF THE PLOTTING SECTION**
+### Missing Years For Infant Mortality Data
 
-## More Data Processing
-### Missing Years For Infant Mortality Data
 ```{r}
@@
 ### Perform Year Groupby to Identify Distinct Missing Years
+
 ```{r}
@@
 ### Extract Only Year 2000 From africadata As New Object
+
 ```{r}
@@
 ```
+
 ### africadata2000 Data Structure
+
 ```{r}
@@
 ```
+
 ### africadata2000 Data Summary
+
 ```{r}
@@
 ## More Plotting
+
 ### Create The Same Plots For africadata2000 Object
+
 #### Filter New Object to Create Graph Datasets (1/2)
+
 ```{r}
@@
 ```
+
 #### Filter New Object to Create Graph Datasets (2/2)
+
 ```{r}
@@
 #### Infant Mortality Effect
+
 ```{r}
@@
     infant_survival_rate2000$life_expectancy ~ infant_survival_rate2000$infant_mortality),
-  col='black',
+  col='darkblue',
   lwd=1.75)
@@
 #### Population Effect
+
 ```{r}
@@
   african_mortality2000$life_expectancy ~ african_mortality2000$population),
-col='black',
+col='darkblue',
 lwd=1.75)
@@
 ## Simple Model Fits
-### Fit Life Expectancy to Infant Mortality
+
+********************This code added by Edward Cruz************************************************
 ```{r}
@@
 ```
+
+```{r}
+fit1 = lm(life_expectancy ~ log(infant_mortality), data=infant_survival_rate2000)
+summary(fit1)
+```
+
+```{r}
+# Linear regression
+fit1 = lm(life_expectancy ~ infant_mortality, data=infant_survival_rate2000)
+summary(fit1)
+
+# Polynomial regression of degree 2
+fit2 = lm(life_expectancy ~ poly(infant_mortality, 2), data=infant_survival_rate2000)
+summary(fit2)
+
+# Plotting
+plot(
+  infant_survival_rate2000$infant_mortality,
+  infant_survival_rate2000$life_expectancy,
+  main='Infant Mortality vs Life Expectancy',
+  xlab='Infant Mortality',
+  ylab='Life Expectancy',
+  col='aquamarine',
+  pch=19
+)
+
+# Add linear regression line
+abline(fit1, col='darkblue')
+
+# Add polynomial regression line
+new_data <- data.frame(infant_mortality = seq(min(infant_survival_rate2000$infant_mortality), max(infant_survival_rate2000$infant_mortality), length.out=100))
+lines(new_data$infant_mortality, predict(fit2, new_data), col='cornflowerblue')
+
+```
+
 ### Fit Life Expectancy to Population
+
 ```{r}
@@
 ```
+
+```{r}
+# Logarithmic regression
+fit2 = lm(life_expectancy ~ log(population), data=african_mortality2000)
+summary(fit2)
+
+```
+
+```{r}
+# Linear regression
+fit1 = lm(life_expectancy ~ population, data=african_mortality2000)
+summary(fit1)
+
+# Polynomial regression of degree 2
+fit2 = lm(life_expectancy ~ poly(population, 2), data=african_mortality2000)
+summary(fit2)
+
+# Plotting
+plot(
+  african_mortality2000$population,
+  african_mortality2000$life_expectancy,
+  main='Population vs Life Expectancy',
+  xlab='Population',
+  ylab='Life Expectancy',
+  col='aquamarine',
+  pch=19
+)
+
+# Add linear regression line
+abline(fit1, col='darkblue')
+
+# Add polynomial regression line
+new_data <- data.frame(population = seq(min(african_mortality2000$population), 
+                                        max(african_mortality2000$population), length.out=100))
+lines(new_data$population, predict(fit2, new_data), col='cornflowerblue')
+
+# Add a legend to differentiate the lines
+legend("topright", legend=c("Linear Fit", "Polynomial Fit (degree 2)"), col=c("darkblue", "cornflowerblue"), lwd=2)
+```
+
 #### Conclusion
+The African population and life expectancy analysis shows no significant relationship between population size and life expectancy in African countries according to this linear model.
+The very low R-squared value indicates that population size explains only a tiny fraction of the variation in life expectancy.
+This suggests that other factors, such as healthcare access, economic conditions, education, and other social determinants, are likely much more important in explaining differences in life expectancy across African countries.
+TO reiterate, there is tremendous suffering and mortality with the children of the populations however the resiliency of this country has been successfully documented indicating the economics, education and healthcare is improving to continue allowing the population with access to these areas to successfully prosper and survive.
+
+***  This part contributed by Leonel Salazar, IGE436 ***
+
+## Loading and Checking Data
+### Import libraries
+
+```{r}
+library(dslabs) # to import dataset
+library(tidyverse) # data aggregation
+library(dplyr) # data aggregation? 
+library(renv)
+library(caret)
+data("stars")
+```
+
+### Help Function
+```{r}
+help("stars")
+```
+
+### Data Structure Overview
+```{r}
+str(stars)
+```
+### Data Summary
+```{r}
+summary(stars)
+```
+### Class Functions - Object Determination
+```{r}
+class(stars)
+```
+
+### Star Type based on Magnitude
+```{r}
+Type_Magnitude = stars %>%
+  select(type, magnitude )
+head(Type_Magnitude)
+```
+
+### Stars Type and Magnitude Structure
+```{r}
+str(Type_Magnitude)
+```
+
+```{r}
+# Convert to factor (default order is alphabetical)
+Type_Magnitude$type <- as.factor(Type_Magnitude$type)
+
+str(Type_Magnitude)
+
+```
+
+### Stars Type and Magnitude data summary
+```{r}
+summary(Type_Magnitude)
+```
+### Stars by Type and Temp
+```{r}
+Type_Temp = stars %>%
+  select(type, temp)
+head(Type_Temp)
+```
+
+### Type Temp Data Structure
+```{r}
+Type_Temp$type <- as.factor(Type_Temp$type)
+
+# Convert to numeric
+Type_Temp$temp <- as.numeric(Type_Temp$temp)
+
+str(Type_Temp)
+
+```
+
+### Type Temp Data Summary
+```{r}
+summary(Type_Temp)
+```
+```{r}
+
+Mag_Temp = stars %>%
+  select(magnitude, temp)
+head(Mag_Temp)
+
+```
+
+### Histogram of variables
+
+```{r}
+
+hist(Type_Magnitude$magnitude)
+hist(Type_Temp$temp)
+
+```
+
+### Plotting
+### Star Magnitude as a function of Star temperature
+
+```{r}
+regression = lm(Mag_Temp$magnitude ~ Mag_Temp$temp)
+rsq <- function(x, y) summary(lm(Mag_Temp$magnitude ~ Mag_Temp$temp))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  Mag_Temp$magnitude,
+  Mag_Temp$temp,
+  main='Magnitude and temperature of star',
+  xlab='Magnitude',
+  ylab='Temperature',
+   col='lightblue', 
+   bg='black', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    Mag_Temp$magnitude ~ Mag_Temp$temp),
+  col='red',
+  lwd=1.75)
+```
+
+```{r}
+
+fit1 = lm(Mag_Temp$magnitude ~ Mag_Temp$temp)
+summary(fit1)
+
+```
+*** Based on this analysis, there's a statistically significant negative linear relationship between "temp" and "magnitude" in the "Mag_Temp" data.
+The model explains about 40% of the variance in "magnitude.
+***
+
+```{r}
+regression2 = lm(Type_Magnitude$magnitude ~ Type_Magnitude$type)
+rsq <- function(x, y) summary(lm(Type_Magnitude$magnitude ~ Type_Magnitude$type))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  Type_Magnitude$magnitude,
+  Type_Magnitude$type,
+  main='Magnitude by type of star',
+  xlab='Magnitude',
+  ylab='Type',
+   col='lightblue', 
+   bg='black', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    Type_Magnitude$magnitude ~ Type_Magnitude$type),
+  col='red',
+  lwd=1.75)
+
+```
+
+```{r}
+
+fit2 = lm(Type_Magnitude$magnitude ~ Type_Magnitude$type)
+summary(fit2)
+
+```
+*** Based on this analysis, there's a statistically significant relationship between "type" and "magnitude" in the "Type_Magnitude" data. The model explains about 73% of the variance in "magnitude ***
+
+```{r}
+
+regression3 = lm(Type_Temp$temp ~ Type_Temp$type)
+rsq <- function(x, y) summary(lm(Type_Temp$temp ~ Type_Temp$type))$r.squared
+r_square = round(rsq(obs, mod), 3)
+# summary(regression)$coefficients - keep to sanity check
+coeff = round(summary(regression)$coefficients[2],3)
+y_int = round(summary(regression)$coefficients[1],3)
+plot(
+  Type_Temp$temp,
+  Type_Temp$type,
+  main='Temp by type of star',
+  xlab='Temp',
+  ylab='Type',
+   col='lightblue', 
+   bg='black', 
+   pch = 21,
+)
+# rsq_str = expression(paste("r"^"2:"))
+rsq_str = "r2: "
+text(175, 75, paste0('y = ', coeff, 'x + ', y_int), cex=0.75)
+text(170, 70, paste0(rsq_str, r_square), cex=0.75)
+abline(
+  lm(
+    Type_Temp$temp ~ Type_Temp$type),
+  col='red',
+  lwd=1.75)
+
+```
+
+```{r}
+
+fit3 = lm(Type_Temp$temp ~ Type_Temp$type)
+summary(fit3)
+
+```
+
+*** The very high R-squared value (0.8924) suggests a strong relationship between "type" and "temp." However, it's important to be cautious with such a high value, as it could indicate overfitting ***
+
+#### Conclusion
+
+*** It might be possible to accurately predict what type of star a star may be by knowing the temperature and magnitude of a star.
+We would need better data because we can see in the histograms that the data is not linear or very normal.
+We would have to transform the data to get more accurate and reliable results.
+***

### 2024-06-27 hcny_cleaned.qmd
html update 6:49pm

---
title: "loading/cleaning/wrangling/exploring"
author: "Edward Cruz"
format: docx
---

```{r}
library(readxl)
#update.packages("tidyverse")
library(tidyverse)
#update.packages("dplyr")
library(dplyr)
#update.packages('mgcv')
library(mgcv)
```

```{r}
# Output cleaned data file to a csv file.
hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)
```

```{r}
head(hcny_data)
```
```{r}
str(hcny_data)
```
```{r}
summary(hcny_data)
```
```{r}
view(hcny_data)
```

```{r}
# Merge the first two rows to create a proper header
header <- hcny_data[1:4, ]
header <- sapply(header, function(x) paste(na.omit(x), collapse = " "))
colnames(hcny_data) <- header
head(header)
```

```{r}
# Remove the first two rows (header rows) and any completely blank rows
hcnydata_cleaned <- hcny_data[-c(1,2), ]
head(hcnydata_cleaned)
```
```{r}
# Assuming hcny_data_cleaned is your cleaned data frame
colnames(hcnydata_cleaned) <- c("State", "Varroa_Mites", "Other_Parasites_1/", "Disease_2/", "Pesticides", "Other_3/", "Unknown")
```

```{r}
head(hcnydata_cleaned)
```

```{r}
# Convert specified columns to numeric
hcnydata_cleaned <- hcnydata_cleaned %>%
  mutate(across(c("Varroa_Mites", "Other_Parasites_1/", "Disease_2/", "Pesticides", "Other_3/", "Unknown"), as.numeric))
```

```{r}
# View the updated data frame
hcnydata_cleaned<-hcnydata_cleaned[-1,]
```

```{r}
head(hcnydata_cleaned)
```
```{r}
# Verify the changes
str(hcnydata_cleaned)
```

```{r}
head(hcnydata_cleaned)
```

```{r}
view(hcnydata_cleaned)
```

```{r}
ggplot(hcnydata_cleaned, aes(Varroa_Mites)) + geom_histogram()
```

```{r}
ggplot(hcnydata_cleaned, aes(x = Varroa_Mites)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Varroa Mites",
       x = "Varroa Mites",
       y = "Frequency")
```
```{r}
# Example scatter plot of two variables
ggplot(hcnydata_cleaned, aes(x = Varroa_Mites, y = "Desease_2/")) +
  geom_point() +
  labs(title = "Scatter Plot of Varroa Mites vs Disease",
       x = "Varroa Mites",
       y = "2 American and European Foul Brood")
```

```{r}
view(hcnydata_cleaned)
```

```{r}
# Write the cleaned data to a new CSV file
write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcnydata_cleaned.csv", row.names = FALSE)
```

********************Conclusion- all code below is test code right now********************************

```{r}
# Assuming hcnydata_cleaned is your data frame
hcnydata_cleaned <- hcnydata_cleaned[apply(hcnydata_cleaned, 1, function(x) !all(is.na(x))), ]

# Print the modified data frame to verify the changes
head(hcnydata_cleaned)
view(hcnydata_cleaned)

```

```{r}
str(hcnydata_cleaned)
```

```{r, message=FALSE}
#update.packages('mgcv')
library(mgcv)
```

```{r}
# Convert the 'state' column to a factor
hcnydata_cleaned$state <- as.factor(hcnydata_cleaned$state)

# Display the resulting data frame
head(hcnydata_cleaned)
```

```{r}
# Step 3: Reindex the rows starting from 1
rownames(hcnydata_cleaned) <- NULL
```

```{r}
# Identify the rows to exclude
rows_to_exclude <- c(46)

# Remove rows 45 to 52 while excluding row 46
data_hcny <- hcnydata_cleaned[-c(45:52)[-which(c(45:52) %in% rows_to_exclude)], ]
```

```{r}
library(ggplot2)

# Create a bar graph to show state and varroa_mites
ggplot(data = data_hcny, aes(x = state, y = varroa_mites)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Varroa Mites by State",
       x = "State",
       y = "Varroa Mites Count") +
  theme_minimal() +
  coord_flip()
```

# Remove all rows with any NA values

```{r}

library(tidyr)

my_data <- data_hcny %>% drop_na()

# Verify that NAs have been removed
glimpse(my_data)
```
# Bivariate Analysis

```{r, message=FALSE, warning=FALSE}
## Example of a bivariate plot between 'state' and 'varroa_mites'
ggplot(my_data, aes(x = disease, y = varroa_mites)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Bivariate Relationship between State and Varroa Mites",
       x = "State",
       y = "Varroa Mites")
```
# Multivariable GLM

```{r}
install.packages("tidymodels")
library(tidymodels)
library(broom)

# Specify the GLM model
glm_spec <- linear_reg() %>%
  set_engine("glm")

# Create a recipe for preprocessing the data
glm_recipe <- recipe(varroa_mites ~ other_pests + disease + pesticides + other + unknown, data = my_data) %>%
  step_normalize(all_predictors())

# Create a workflow
glm_workflow <- workflow() %>%
  add_model(glm_spec) %>%
  add_recipe(glm_recipe)

# Fit the model
glm_fit <- fit(glm_workflow, data = my_data)

# Print the model summary using tidy()
model_summary <- tidy(glm_fit)
print(model_summary)
```

### 2024-06-27 hcny_cleaned_updatedMoji.qmd
html update 6:49pm

--- 2024-06-27 hcny_cleaned.qmd
@@
 title: "loading/cleaning/wrangling/exploring"
-author: "Edward Cruz"
+author: "Mojisola Popoola, Leonel Salazar, ED Cruz, Amelia"
 format: docx
@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 library(readxl)
-#update.packages("tidyverse")
 library(tidyverse)
-#update.packages("dplyr")
 library(dplyr)
-#update.packages('mgcv')
-library(mgcv)
+library(readr)
+library(tidymodels)
 ```
@@
 ```
+
 ```{r}
-view(hcny_data)
+# Step 2: Merge the first two rows to create a new header
+new_header <- paste(hcny_data[1, ], hcny_data[2, ], sep = " ")
+
+# Step 3: Set the new header
+colnames(hcny_data) <- new_header
+
+# Step 4: Remove the first three rows
+hcny_data <- hcny_data[-c(1:3), ]
+
+# Step 5: Remove completely blank rows
+hcny_data <- hcny_data[rowSums(hcny_data != "") > 0, ]
+
+# Display the resulting data frame
+head(hcny_data)
 ```
@@
 ```{r}
-# Merge the first two rows to create a proper header
-header <- hcny_data[1:4, ]
-header <- sapply(header, function(x) paste(na.omit(x), collapse = " "))
-colnames(hcny_data) <- header
-head(header)
+# Assuming hcny_data is your cleaned data frame
+colnames(hcny_data) <- c("state", "varroa_mites", "other_pests", "disease", "pesticides", "other", "unknown")
 ```
@@
 ```{r}
-# Remove the first two rows (header rows) and any completely blank rows
-hcnydata_cleaned <- hcny_data[-c(1,2), ]
+head(hcny_data)
+```
+
+```{r}
+# Convert specified columns to numeric with warning suppression
+hcnydata_cleaned <- hcny_data %>%
+  mutate(across(c("varroa_mites", "other_pests", "disease", "pesticides", "other", "unknown"), ~suppressWarnings(as.numeric(.))))
+
+# Display the resulting data frame
 head(hcnydata_cleaned)
-```
-```{r}
-# Assuming hcny_data_cleaned is your cleaned data frame
-colnames(hcnydata_cleaned) <- c("State", "Varroa_Mites", "Other_Parasites_1/", "Disease_2/", "Pesticides", "Other_3/", "Unknown")
 ```
@@
 
-```{r}
-# Convert specified columns to numeric
-hcnydata_cleaned <- hcnydata_cleaned %>%
-  mutate(across(c("Varroa_Mites", "Other_Parasites_1/", "Disease_2/", "Pesticides", "Other_3/", "Unknown"), as.numeric))
-```
-
-```{r}
-# View the updated data frame
-hcnydata_cleaned<-hcnydata_cleaned[-1,]
-```
-
-```{r}
-head(hcnydata_cleaned)
-```
 ```{r}
@@
 
-```{r}
-view(hcnydata_cleaned)
+```{r,message=FALSE, warning=FALSE}
+ggplot(hcnydata_cleaned, aes(varroa_mites)) + geom_histogram()
 ```
@@
 ```{r}
-ggplot(hcnydata_cleaned, aes(Varroa_Mites)) + geom_histogram()
-```
-
-```{r}
-ggplot(hcnydata_cleaned, aes(x = Varroa_Mites)) +
-  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
-  labs(title = "Histogram of Varroa Mites",
-       x = "Varroa Mites",
-       y = "Frequency")
-```
-```{r}
-# Example scatter plot of two variables
-ggplot(hcnydata_cleaned, aes(x = Varroa_Mites, y = "Desease_2/")) +
-  geom_point() +
-  labs(title = "Scatter Plot of Varroa Mites vs Disease",
-       x = "Varroa Mites",
-       y = "2 American and European Foul Brood")
-```
-
-```{r}
-view(hcnydata_cleaned)
+view((hcnydata_cleaned))
 ```
@@
 # Write the cleaned data to a new CSV file
-write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcnydata_cleaned.csv", row.names = FALSE)
+write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", row.names = FALSE)
 ```

### 2024-06-28 docs/Data-Exercise/data-exercise.qmd
All key files should be in docs/data exercise/data-exercise

--- 2024-06-27 Misc/data-exercise_Ecruz.qmd
@@
 
-Placeholder file for the future data exercise.
+Placeholder file for the future R coding exercise.
 
@@
 ```
-
-\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\Conclusion*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

### 2024-07-04 cdcdata-exercise/deathcount_cleaned.qmd
Updated R Code to create synthetic data 07.04.24

---
title: "loading/cleaning/wrangling/exploring"
author: "Edward Cruz"
format: docx
---

```{r}
library(readxl)
#update.packages("tidyverse")
library(tidyverse)
#update.packages("dplyr")
library(dplyr)
#update.packages('mgcv')
library(mgcv)
```

```{r}
# Output cleaned data file to a csv file.
DthCount_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/Weekly_DeathCounts.csv", header = FALSE, stringsAsFactors = FALSE)
```

```{r}
head(DthCount_data)
```

```{r}
str(DthCount_data)
```

```{r}
summary(DthCount_data)
```

```{r}
#view(DthCount_data)
```

```{r}
# Merge the first two rows to create a proper header
header <- DthCount_data[1:2, ]
header <- sapply(header, function(x) paste(na.omit(x), collapse = " "))
colnames(DthCount_data) <- header
head(header)
```

```{r}
# Remove the first row (header rows) and any completely blank rows
DthCount_cleaned <- DthCount_data[-c(1,1), ]
head(DthCount_cleaned)
```

```{r}
#view(DthCount_cleaned)
```

```{r}
# Assuming DthCount_cleaned is your cleaned data frame
colnames(DthCount_cleaned) <- c("Jurisdiction", "Week_Ending_Date", "State", "Year", "Week","Cause","Death_Count","Cause_Subgroup","Period","Suppress","Note","Avg_No_Deaths","Variance_2015_2019_2020", "Perc_Diff_2015_2019_2020","Types")
```

```{r}
head(DthCount_cleaned)
```

```{r}
# Convert specified columns to numeric
DthCount2_cleaned <- DthCount_cleaned %>%
  mutate(across(c("Year", "Week", "Death_Count", "Avg_No_Deaths", "Variance_2015_2019_2020", "Perc_Diff_2015_2019_2020"), as.numeric))
```

```{r}
str(DthCount2_cleaned)
```

```{r}
head(DthCount2_cleaned)
```

```{r}
# Write the cleaned data to a new CSV file
write.csv(DthCount2_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DthCount2_cleaned.csv", row.names = FALSE)
```

```{r}
DthCount3_cleaned <- DthCount2_cleaned %>% select(-Suppress,-Note,-Types)

```

```{r}
#view(DthCount3_cleaned)
```

```{r}
ggplot(DthCount3_cleaned, aes(Death_Count)) + geom_histogram()
```

```{r}
ggplot(DthCount3_cleaned, aes(x = Avg_No_Deaths)) +
  geom_histogram(binwidth = 950, fill = "blue", color = "black") +
  labs(title = "Histogram of Death Count",
       x = "Average Number of Deaths",
       y = "Frequency")
```

```{r}
# Example scatter plot of two variables
ggplot(DthCount3_cleaned, aes(x = Death_Count, y = Avg_No_Deaths)) +
  geom_point() +
  labs(title = "Scatter Plot",
       x = "Deaths",
       y = "Average Number of Deaths")
```

```{r}
# Enhanced scatter plot of two variables
ggplot(DthCount3_cleaned, aes(x = Death_Count, y = Avg_No_Deaths)) +
  geom_point(color = "blue", size = 5, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Scatter Plot of Death Count vs Average Number of Deaths",
       x = "Death Count",
       y = "Average Number of Deaths") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

```{r}
# Bar chart of Death Count per Year
ggplot(DthCount3_cleaned, aes(x = Year, y = Death_Count)) +
  geom_bar(stat = "identity", outline = "black", color = "darkorange", alpha = .8) +
  labs(title = "Bar Chart of Death Count Per Year",
       x = "Year",
       y = "Death Count") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

```{r}
# Write the cleaned data to a new CSV file
write.csv(DthCount3_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DthCount3_cleaned.csv", row.names = FALSE)
```

```{r}
# Assuming dat_cleaned is your data frame
DthCount3A_cleaned <- DthCount3_cleaned[apply(DthCount3_cleaned, 1, function(x) !all(is.na(x))), ]

# Print the modified data frame to verify the changes
head(DthCount3_cleaned)

```

```{r}
#view(DthCount3_cleaned)
```

```{r}
str(DthCount3A_cleaned)
```

```{r, message=FALSE}
#update.packages('mgcv')
library(mgcv)
```

```{r}
# Convert the 'state' column to a factor
DthCount3A_cleaned$State <- as.factor(DthCount3_cleaned$State)

# Display the resulting data frame
head(DthCount3_cleaned)
```

```{r}
# Step 3: Reindex the rows starting from 1
rownames(DthCount3A_cleaned) <- NULL
```

```{r}
str(DthCount3A_cleaned)
```

```{r}
library(ggplot2)

# Bar chart of Death Count by State with skyblue fill and flipped coordinates
ggplot(data = DthCount3A_cleaned, aes(x = State, y = Death_Count)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Varroa Mites by State",
       x = "State",
       y = "Death Count") +
  theme_minimal() +
  coord_flip() +
  theme(
    axis.text.x = element_text(size = 12),  # Adjust the size as needed
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )

```

# Remove all rows with any NA values

```{r}

library(tidyr)

DeathCnt_data <- DthCount3A_cleaned %>% drop_na()

# Verify that NAs have been removed
glimpse(DeathCnt_data)
```

```{r, message=FALSE, warning=FALSE}
## Example of a bivariate plot between 'state' and 'varroa_mites'
ggplot(DeathCnt_data, aes(x = State, y = Death_Count)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Death Count by State",
       x = "State",
       y = "Death Count")
```

```{r}
DeathCntFinal_data<-DeathCnt_data
```

```{r}
library(ggplot2)

# Bar chart of Death Count by State with skyblue fill and flipped coordinates
ggplot(data = DeathCntFinal_data, aes(x = State, y = Death_Count)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Varroa Mites by State",
       x = "State",
       y = "Death Count") +
  theme_minimal() +
  coord_flip() +
  theme(
    axis.text.x = element_text(size = 12),  # Adjust the size as needed
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )

```

```{r}
# Write the cleaned data to a new CSV file
write.csv(DeathCntFinal_data, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DeathCntFinal_data.csv", row.names = FALSE)
```

################################Eddie Cruz #################################################################

```{r}
# Install and load necessary packages
library(synthpop)
```

```{r}
# Output cleaned data file to a csv file.
DthCount_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/Storage CDCdata-exercise/DeathCntFinal_data.csv", header = FALSE, stringsAsFactors = FALSE)
```

```{r}
# Load your data (assuming it's saved as a CSV file)
data <- read.csv("DeathCntFinal_data.csv")

```

```{r}
# Subset the data to keep only the first 500 observations
data <- data[1:500, ]
```

```{r}
# Handle missing values (for simplicity, we'll remove rows with any NAs)
data <- na.omit(data)

```

```{r}
# Identify and remove constant columns
constant_columns <- sapply(data, function(col) length(unique(col)) == 1)
data <- data[, !constant_columns]
```

```{r}
# Convert Week_Ending_Date to Date type and create a new variable for month
data$Week_Ending_Date <- as.Date(data$Week_Ending_Date)
```

```{r}
# Remove the original Week_Ending_Date column to avoid issues with high levels
data$Week_Ending_Date <- NULL

```

```{r}
# Convert numeric variables with few levels to factors
data$Year <- as.factor(data$Year)
data$Period <- as.factor(data$Period)

# Create a predictor matrix to handle collinear variables
predictor_matrix <- syn(data)$predictor.matrix
# Set collinear variables to 0 for others' synthesis
predictor_matrix["Period", "Year"] <- 0
predictor_matrix["Year", "Period"] <- 0
predictor_matrix["Avg_No_Deaths", "Death_Count"] <- 0
predictor_matrix["Death_Count", "Avg_No_Deaths"] <- 0

# Generate synthetic data
synth_data <- syn(data, predictor.matrix = predictor_matrix, maxfaclevels = 200, minnumlevels = 5)

# Access the synthetic dataset
synthetic_data <- synth_data$syn

# Save the synthetic dataset to a CSV file
write.csv(synthetic_data, "synthetic_data.csv", row.names = FALSE)
```

### 2024-07-05 cdcdata-exercise/deathcount_cleaned.qmd
updated R code 07.05.24

@@
 library(readxl)
-#update.packages("tidyverse")
 library(tidyverse)
-#update.packages("dplyr")
 library(dplyr)
-#update.packages('mgcv')
 library(mgcv)
@@
 ```{r}
+#Review variables
 head(DthCount_data)
@@
 ```{r}
+# Type of variable
 str(DthCount_data)
+
 ```
@@
 ```{r}
-# Assuming DthCount_cleaned is your cleaned data frame
+# Clean up variable labels
 colnames(DthCount_cleaned) <- c("Jurisdiction", "Week_Ending_Date", "State", "Year", "Week","Cause","Death_Count","Cause_Subgroup","Period","Suppress","Note","Avg_No_Deaths","Variance_2015_2019_2020", "Perc_Diff_2015_2019_2020","Types")
@@
 # Write the cleaned data to a new CSV file
-write.csv(DthCount2_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DthCount2_cleaned.csv", row.names = FALSE)
+#write.csv(DthCount2_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DthCount2_cleaned.csv", row.names = FALSE)
 ```
@@
 ```{r}
+#suppress three variables
 DthCount3_cleaned <- DthCount2_cleaned %>% select(-Suppress,-Note,-Types)
-
 ```
@@
 # Write the cleaned data to a new CSV file
-write.csv(DthCount3_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DthCount3_cleaned.csv", row.names = FALSE)
+#write.csv(DthCount3_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DthCount3_cleaned.csv", row.names = FALSE)
 ```
@@
 DthCount3A_cleaned <- DthCount3_cleaned[apply(DthCount3_cleaned, 1, function(x) !all(is.na(x))), ]
+```
 
+```{r}
 # Print the modified data frame to verify the changes
@@
 DthCount3A_cleaned$State <- as.factor(DthCount3_cleaned$State)
+```
 
+```{r}
 # Display the resulting data frame
@@
 ```
-
+##########################################################Edward Cruz###########################################################################
 ```{r}
@@
 ```{r}
+#Remove NA values
+DeathCnt_data <- DthCount3A_cleaned %>% drop_na()
+```
 
-library(tidyr)
-
-DeathCnt_data <- DthCount3A_cleaned %>% drop_na()
-
+```{r}
 # Verify that NAs have been removed
@@
 ```
-
+# View a cleaner plot
 ```{r, message=FALSE, warning=FALSE}
@@
 ```{r}
-# Write the cleaned data to a new CSV file
+# Write the cleaned data to a new CSV file which will be used to create synthetic data
 write.csv(DeathCntFinal_data, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/cdcdata-exercise/DeathCntFinal_data.csv", row.names = FALSE)
@@
 
-################################Eddie Cruz #################################################################
+################################Eddie Cruz & Leonels Salazar #################################################################
 
@@
 data <- na.omit(data)
-
 ```
@@
 data$Week_Ending_Date <- NULL
-
 ```
@@
 ```
+#####################################################Leonel Salazar#############################################################################

### 2024-07-10 data-exercise/data-exercise.qmd
online exactly the same as my local
update website for data exercise

@@
 ---
-title: "Date Exercise"
+title: "R Coding Exercise"
 ---
 
-Placeholder file for the future data exercise.
+Placeholder file for data coding exercise.
 
+## Loading and Checking Data
+### Import libraries
 ```{r}

### 2024-07-10 cdcdata-exercise/cdcdata-exercise.qmd
page for CDCdata cleaning exercise
Updating CDC data wrong label found

--- 2024-07-05 cdcdata-exercise/deathcount_cleaned.qmd
@@
 ---
-title: "loading/cleaning/wrangling/exploring"
-author: "Edward Cruz"
-format: docx
+title: "CDC data Exercise"
 ---
+
+Placeholder file for cdc data coding exercise.
+
+## Loading and Checking Data
+### Import libraries
 
@@
 
-# Bar chart of Death Count by State with skyblue fill and flipped coordinates
+# Bar chart of Death Count by State with darkblue fill and flipped coordinates
 ggplot(data = DthCount3A_cleaned, aes(x = State, y = Death_Count)) +
   geom_bar(stat = "identity", fill = "darkblue") +
-  labs(title = "Varroa Mites by State",
+  labs(title = "Death Count by State",
        x = "State",
@@
   geom_bar(stat = "identity", fill = "darkblue") +
-  labs(title = "Varroa Mites by State",
+  labs(title = "Death Count by State",
        x = "State",

### 2024-07-11 presentation-exercise/presentation-exercise.qmd
Presentation Exercise

---
title: "Presentation Exercise"
---

Placeholder file for the future data/results presentation exercise.
***#New York Times - Some Collegs have more students from the top 1 percent than the Botto 60. Find Yours. Jan 18, 2017***

```{r}
# Load necessary libraries
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggthemes)

# Create synthetic data
income_groups <- c("Bottom 20%", "2nd quintile", "3rd quintile", "4th quintile", "Top 20%", "Top 10%", "Top 5%", "Top 1%", "Top 0.1%")
elite_university <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9) # Example percentages
any_college <- c(0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 1, 1) # Example percentages

# Combine into a data frame
data <- data.frame(
  income_group = factor(income_groups, levels = income_groups),
  elite_university = elite_university,
  any_college = any_college
)

# Convert to long format for ggplot2
data_long <- data %>%
  pivot_longer(cols = c(elite_university, any_college), names_to = "type", values_to = "percentage")

# Plot the data
ggplot(data_long, aes(x = income_group, y = percentage, fill = type)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("elite_university" = "red", "any_college" = "blue"), 
                    labels = c("Elite University", "Any College")) +
  labs(
    title = "Where today’s 25-year-olds went to college, grouped by their parents’ income",
    x = "Parental Income Group",
    y = "Percentage",
    fill = "College Type"
  ) +
  theme_fivethirtyeight() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12)  # Adjust title size here (default is 18)
  )

```

***NBC News Inflation in America: Where are prices rising and falling?***
```{r}
#NBC News Inflation in America: Where are prices rising and falling?
# Load necessary libraries
library(ggplot2)

# Example data (replace with actual data)
categories <- c("Overall inflation", "Airfare", "Cars", "Clothing", "Electricity", "Gas", "Housing costs", "Meat and eggs")
percentage_change <- c(2.5, -15.3, 6.7, 1.2, 3.8, -4.1, 4.5, 5.9)  # Example percentages

# Create a data frame
data <- data.frame(
  category = categories,
  percentage_change = percentage_change
)

# Plot the data
ggplot(data, aes(x = category, y = percentage_change, fill = factor(sign(percentage_change)))) +
  geom_bar(stat = "identity", color = "black", show.legend = FALSE) +
  scale_fill_manual(values = c("red", "green"), labels = c("Decrease", "Increase")) +
  geom_text(aes(label = paste0(percentage_change, "%")), vjust = ifelse(percentage_change >= 0, -0.5, 1.5), color = "black", size = 4) +
  coord_flip() +  # Flip the coordinates to make horizontal bars
  labs(
    title = "Year-over-Year Price Changes for Major Categories of Goods",
    subtitle = "Data as of June 12, 2024\nAll items are percentage change from a year earlier",
    x = "Category",
    y = "Percentage Change (%)"
  ) +
  theme_economist(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.title.y = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),  # Adjusted axis text size
    axis.text.y = element_text(hjust = 1)
  )

```
***NBC News Inflation in America: Where are prices rising and falling?***
```{r}
#NBC News Inflation in America: Where are prices rising and falling?
# Load necessary libraries
library(ggplot2)

# Example data (replace with actual data)
years <- c(2019, 2020, 2021, 2022, 2023)
cpi_change <- c(1.8, 1.4, 4.2, 2.9, 5.1)  # Example CPI percentage change

# Create a data frame
data <- data.frame(
  year = years,
  cpi_change = cpi_change
)

# Plot the data
ggplot(data, aes(x = year, y = cpi_change)) +
  geom_line(color = "#0072B2", size = 1.5) +  # Line plot
  geom_point(color = "#0072B2", size = 3) +   # Points at data points
  geom_text(aes(label = paste0(cpi_change, "%")), vjust = -0.5, hjust = 0.5, color = "#0072B2", size = 4) +  # Labels on points
  scale_x_continuous(breaks = years, labels = years) +  # X-axis labels
  scale_y_continuous(limits = c(0, max(cpi_change) * 1.1), breaks = seq(0, max(cpi_change) * 1.1, by = 1), labels = paste0(seq(0, max(cpi_change) * 1.1, by = 1), "%")) +  # Y-axis labels
  labs(
    title = "Consumer Price Index Change from One Year Earlier",
    x = "Year",
    y = "CPI Change (%)"
  ) +
  theme_wsj(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    panel.grid.major.x = element_line(color = "grey90"),  # Gridlines
    panel.grid.minor.x = element_blank(),  # Minor gridlines
    panel.grid.major.y = element_line(color = "grey90")
  )

```
***NBC News Inflation in America: Where are prices rising and falling?***
```{r}
#NBC News Inflation in America: Where are prices rising and falling?
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Example data (replace with actual data)
categories <- c("Overall inflation", "Airfare", "Cars", "Clothing", "Electricity", "Gas", "Housing costs", "Meat and eggs")
months <- c("July '23", "Aug. '23", "Sept. '23", "Oct. '23", "Nov. '23", "Dec. '23", "Jan. '24", "Feb. '24", "Mar. '24", "Apr. '24", "May '24")
percentage_changes <- matrix(c(
  3.2, -5.1, 2.9, 1.5, 3.8, -2.3, 4.7, 5.2, 4.3, 3.9, 3.6,
  -15.2, -10.5, -8.1, -9.4, -6.3, -12.1, -7.8, -5.9, -6.5, -7.2, -8.3,
  6.5, 4.8, 5.2, 7.1, 6.3, 5.9, 5.7, 6.2, 6.8, 7.5, 7.9,
  1.2, 1.8, 2.5, 2.9, 2.4, 1.6, 2.8, 3.1, 3.4, 3.6, 3.9,
  4.1, 3.5, 3.9, 4.2, 3.8, 4.5, 4.3, 4.6, 4.8, 4.2, 4.1,
  -3.2, -4.1, -3.5, -3.9, -3.6, -4.3, -3.8, -3.4, -3.7, -3.9, -4.2,
  5.5, 5.2, 5.9, 6.3, 5.7, 5.8, 5.6, 5.9, 6.1, 5.7, 5.9,
  6.8, 7.1, 6.5, 7.2, 6.9, 6.7, 6.5, 6.3, 6.9, 7.1, 7.3
), nrow = length(categories), byrow = TRUE)

# Create a data frame
data <- data.frame(
  category = rep(categories, each = length(months)),
  month = rep(months, length(categories)),
  percentage_change = as.vector(percentage_changes)
)

# Plot the heatmap
ggplot(data, aes(x = month, y = category, fill = percentage_change)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#D7191C", mid = "white", high = "#1A9641", midpoint = 0, limits = c(-10, 10), name = "Percentage Change (%)") +
  geom_text(aes(label = paste0(percentage_change, "%")), size = 3, color = "black") +
  labs(
    title = "Year-over-Year Price Changes for Major Categories of Goods",
    subtitle = "Data as of June 12, 2024\nAll items are percentage change from a year earlier",
    x = "Month",
    y = "Category"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 16, face = "bold"),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12)
  )

```

### 2024-07-11 presentation-exercise/presentation-exercise.qmd
Secondary Update of Presentation Exercise

@@
   labs(
-    title = "Consumer Price Index Change from One Year Earlier",
+    title = "Consumer Price Index Change from One Year Prior",
     x = "Year",
@@
   ) +
-  theme_wsj(base_size = 14) +
+  theme_wsj(base_size =1) +
   theme(
@@
   scale_fill_gradient2(low = "#D7191C", mid = "white", high = "#1A9641", midpoint = 0, limits = c(-10, 10), name = "Percentage Change (%)") +
-  geom_text(aes(label = paste0(percentage_change, "%")), size = 3, color = "black") +
+  geom_text(aes(label = paste0(percentage_change, "%")), size = 1, color = "black") +
   labs(
-    title = "Year-over-Year Price Changes for Major Categories of Goods",
+    title = "Year-over-Year Price Changes for Major Category of Goods",
     subtitle = "Data as of June 12, 2024\nAll items are percentage change from a year earlier",
@@
   ) +
-  theme_classic(base_size = 14) +
+  theme_classic(base_size = 10) +
   theme(

### 2024-07-18 Honey breakdown.qmd
This is the update for part 4 project Group10 manuscript date is not 07.17.24

---
title: "Honey Bee Code"
author: "Edward Cruz, Jr."
editor: visual
---

```{r}
# Install necessary packages

# Load the libraries
library(ggplot2)
library(dplyr)
library(scales)
library(showtext)
library(ggimage)
```

```{r}
# Create a data frame with the composition of honey
honey_data <- data.frame(
  component = c("Fructose", "Glucose", "Water", "Maltose", "Trisaccharides, Carbohydrates", "Sucrose", "Minerals, Vitamins, Enzymes"),
  percentage = c(38.5, 31.0, 17.1, 7.2, 4.2, 1.5, 0.5)
)

# Define colors for the segments
honey_data$color <- c("#FFA726", "#FB8C00", "#FFD54F", "#FFB74D", "#90CAF9", "#F06292", "#BA68C8")

# Path to the bee icon
bee_jpg <- "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/DA-6833-02T-202430 Practicum II/bee_2.jpg" # Update with your bee icon path

```

```{r}
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +  # Adjust width here
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  ) +
  # Add the bee icon
  ggimage::geom_image(aes(x = 1.5, y = 0, image = bee_jpg), size = 0.1)  # Adjust x and y to position the bee outside the pie chart
```

### 2024-07-27 tidytuesday-exercise/tidytuesday-exercise.qmd
Tidy Tuesday

---
title: "Assignment # 7 Analysis Workflow"
author: Edward Cruz
date: "`r Sys.Date()`"
format: html
editor: visual
theme: cosmo
---

```{r}
rm(list = ls())
```

# Load libraries

```{r, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(dplyr)
```

# Load the data

```{r}
finalist <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/tidytuesday-exercise/finalists.csv")
```

```{r}
# Display the first few rows of the dataset
head(finalist)
```

```{r}
# Summary of the dataset
summary(finalist)
```

```{r}
# Structure of the dataset
str(finalist)
```

## Hypothesis: "Does the average age of finalists vary significantly by Season?"

## Wrangle the Data

Data wrangling involves cleaning and transforming the data to make it suitable for analysis.

# Remove Unnecessary Columns

```{r}
finalist <- finalist %>%
  select(-c(Hometown, Description))
```

# Convert Birthday Columns to standard Date format

```{r, message=FALSE, warning=FALSE}
finalist <- finalist %>%
  mutate(Birthday = as.Date(Birthday, format = "%d-%b-%y"))
```

# Separate Birthplace into City and State

```{r}
library(tidyr)
finalist <- finalist %>%
  separate(Birthplace, into = c("City", "State"), sep = ", ")

finalist$State <- as.factor(finalist$State)

str(finalist)
```

# Handle Missing Values

```{r}
finalist <- na.omit(finalist)
```

# Calculate Age and create the Age column

```{r}

finalist <- finalist %>%
  mutate(Age = as.numeric(difftime(Sys.Date(), as.Date(Birthday), units = "weeks")) %/% 52)
```

## Exploratory Data Analysis (EDA)

EDA involves summarizing the main characteristics of the data often using visual methods.

# Summary Statistics

```{r}
summary(finalist)
```

```{r}
# Summary statistics for Age by State
summary_stats <- finalist %>%
  group_by(Season) %>%
  summarise(
    Mean_Age = round(mean(Age, na.rm = TRUE), 1),
    Median_Age = round(median(Age, na.rm = TRUE), 1),
    SD_Age = round(sd(Age, na.rm = TRUE), 1),
    Count = n()
  )

# Print the summary statistics
print(summary_stats)
```

# Distribution of Age
```{r}
# Histogram of Age with white background
ggplot(finalist, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "#0000FF", color = "#424242") +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

# Count by State

```{r}
# Set the color to light honey dew
light_honeydew <- "#F0FFF0"

# Create the bar plot with light honey dew background
ggplot(finalist, aes(x = State)) +
  geom_bar(fill = "#00BFFF", color = "#1E90FF") +
  labs(title = "Count by State", x = "State", y = "Frequency") +
  coord_flip() +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = light_honeydew, color = NA),
    plot.background = element_rect(fill = light_honeydew, color = NA)
  )
```

# Scatter Plot (if applicable)

```{r}
ggplot(finalist, aes(x = Season, y = Age)) +
   geom_point(color = "#00008B") +  # Dark blue color for dots
   labs(title = "Scatter Plot of Season vs Age", x = "Season", y = "Age") +
   theme_minimal(base_size = 15) +  # Use a minimal theme for a cleaner look
   theme(
     plot.background = element_rect(fill = "#F0FFF0"),  # Light honey dew background color
     panel.background = element_rect(fill = "#F0FFF0"), # Light honey dew panel background
     panel.grid.major = element_line(color = "gray80"), # Optional: Adjust grid lines if desired
     panel.grid.minor = element_line(color = "gray90")  # Optional: Adjust grid lines if desired
   )
```

## Split the Data into Train/Test Sets using createDataPartition

```{r, message=FALSE, warning=FALSE}
library(caret)
library(tidymodels)
library(rsample)

set.seed(123)

# Create a partition for training (80%) and testing (20%)
trainIndex <- createDataPartition(finalist$Age, p = 0.8, list = FALSE)

# Subset the data into training and testing sets
train_data <- finalist[trainIndex, ]
test_data <- finalist[-trainIndex, ]

# Define train control
cv_folds <- vfold_cv(train_data, v = 10, repeats = 5)
```

We will fit three different types of models to our data using the tidymodels framework:

Linear Regression Random Forest Support Vector Machine (SVM) We'll use cross-validation (CV) for model training and fitting, and evaluate the models based on performance metrics such as RMSE, R-squared, and residuals.

# Define a recipe for preprocessing

```{r}
age_recipe <- recipe(Age ~ State,Season, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

```

# Linear Regression Model

```{r}
# Define the linear regression model
lm_model <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")
```

# Random Forest Model

```{r}
# Define the random forest model
rf_model <- rand_forest() %>%
  set_engine("ranger") %>%
  set_mode("regression")
```

# Support Vector Machine (SVM) Model

```{r}
# Define the SVM model
svm_model <- svm_rbf() %>%
  set_engine("kernlab") %>%
  set_mode("regression")
```

# Create Workflows

We will create workflows that combine the recipe with each model.

```{r}
# Create workflows for each model
lm_workflow <- workflow() %>%
  add_recipe(age_recipe) %>%
  add_model(lm_model)

rf_workflow <- workflow() %>%
  add_recipe(age_recipe) %>%
  add_model(rf_model)

svm_workflow <- workflow() %>%
  add_recipe(age_recipe) %>%
  add_model(svm_model)
```

# Train Models with Cross-Validation

```{r,message=FALSE, warning=FALSE}
# Control object to save predictions
ctrl <- control_grid(save_pred = TRUE)

# Train linear regression model with cross-validation
lm_res <- lm_workflow %>%
  tune_grid(resamples = cv_folds, control = ctrl)

# Train random forest model with cross-validation
rf_res <- rf_workflow %>%
  tune_grid(resamples = cv_folds, control = ctrl)

# Train SVM model with cross-validation
svm_res <- svm_workflow %>%
  tune_grid(resamples = cv_folds, control = ctrl)
```

# Fit Resamples

Fit the models using cross-validation:

```{r,message=FALSE,warning=FALSE}
# Fit resamples for each workflow
lm_res <- fit_resamples(lm_workflow, resamples = cv_folds, control = ctrl)
rf_res <- fit_resamples(rf_workflow, resamples = cv_folds, control = ctrl)
svm_res <- fit_resamples(svm_workflow, resamples = cv_folds, control = ctrl)
```

# Collect Predictions

```{r}
# Collect predictions for each model
lm_predictions <- collect_predictions(lm_res)
rf_predictions <- collect_predictions(rf_res)
svm_predictions <- collect_predictions(svm_res)
```

# Calculate RMSE and R-squared to Evaluate Model Performance

```{r}
library(yardstick)

# Calculate metrics for Linear Regression Model
lm_metrics <- lm_predictions %>%
  group_by(id) %>%
  summarize(
    rmse = rmse_vec(truth = Age, estimate = .pred),
    rsq = rsq_vec(truth = Age, estimate = .pred)
  )

print("Linear Regression Metrics:")
print(lm_metrics)

# Calculate metrics for Random Forest Model
rf_metrics <- rf_predictions %>%
  group_by(id) %>%
  summarize(
    rmse = rmse_vec(truth = Age, estimate = .pred),
    rsq = rsq_vec(truth = Age, estimate = .pred)
  )

print("Random Forest Metrics:")
print(rf_metrics)

# Calculate metrics for SVM Model
svm_metrics <- svm_predictions %>%
  group_by(id) %>%
  summarize(
    rmse = rmse_vec(truth = Age, estimate = .pred),
    rsq = rsq_vec(truth = Age, estimate = .pred)
  )

print("SVM Metrics:")
print(svm_metrics)
```

# Analyze Residuals and Uncertainty

We can also analyze residuals and uncertainty for each model.

```{r}
# Residuals for linear regression model
lm_residuals <- lm_predictions$.pred - lm_predictions$Age
hist(lm_residuals, main = "Linear Regression Residuals", xlab = "Residuals")

# Residuals for random forest model
rf_residuals <- rf_predictions$.pred - rf_predictions$Age
hist(rf_residuals, main = "Random Forest Residuals", xlab = "Residuals")

# Residuals for SVM model
svm_residuals <- svm_predictions$.pred - svm_predictions$Age
hist(svm_residuals, main = "SVM Residuals", xlab = "Residuals")
```

***Model Evaluation Summary***
Performance Metrics (Cross-validation results)
***Linear Regression:***

RMSE: 7.533, 7.469, 7.636, 7.322, 7.534
R²: 0.000735, 0.000419, 0.003392, 0.000386, 0.002878

***Random Forest:***

RMSE: 7.227, 7.084, 7.202, 7.018, 7.128
R²: 0.001538, 0.000015, 0.004366, 0.000328, 0.000998

***SVM:***

RMSE: 6.954, 6.974, 7.017, 6.884, 6.993
R²: 0.010999, 0.013058, 0.030090, 0.004642, 0.020371

```{r}
# Create data frames for each model
linear_regression <- data.frame(
  Model = "Linear Regression",
  RMSE = c(7.533, 7.469, 7.636, 7.322, 7.534),
  R_squared = c(0.000735, 0.000419, 0.003392, 0.000386, 0.002878)
)

random_forest <- data.frame(
  Model = "Random Forest",
  RMSE = c(7.227, 7.084, 7.202, 7.018, 7.128),
  R_squared = c(0.001538, 0.000015, 0.004366, 0.000328, 0.000998)
)

svm <- data.frame(
  Model = "SVM",
  RMSE = c(6.954, 6.974, 7.017, 6.884, 6.993),
  R_squared = c(0.010999, 0.013058, 0.030090, 0.004642, 0.020371)
)

# Combine all data frames
all_models <- rbind(linear_regression, random_forest, svm)

# Calculate summary statistics
summary_stats <- aggregate(all_models[, c("RMSE", "R_squared")], 
                           by = list(Model = all_models$Model), 
                           FUN = function(x) c(mean = mean(x), sd = sd(x)))

# Flatten the summary statistics
summary_stats_flat <- data.frame(
  Model = summary_stats$Model,
  RMSE_mean = summary_stats$RMSE[,1],
  RMSE_sd = summary_stats$RMSE[,2],
  R_squared_mean = summary_stats$R_squared[,1],
  R_squared_sd = summary_stats$R_squared[,2]
)

# Round the results to 4 decimal places
summary_stats_flat[,2:5] <- round(summary_stats_flat[,2:5], 4)

# Print the summary
print(summary_stats_flat)

# If you want a nicer looking table, you can use the knitr package
if (!require(knitr)) install.packages("knitr")
library(knitr)

# Create a nicely formatted table
kable(summary_stats_flat, 
      caption = "Model Evaluation Summary",
      col.names = c("Model", "RMSE (Mean)", "RMSE (SD)", "R² (Mean)", "R² (SD)"),
      align = c("l", "c", "c", "c", "c"),
      format = "markdown")
```

```{r}

library(knitr)
# Create a data frame for decision criteria
decision_criteria <- data.frame(
  Criteria = c("Interpretability", "Computational Efficiency", "Robustness to Overfitting"),
  Linear_Regression = c("High", "High", "Low"),
  Random_Forest = c("Moderate", "Moderate", "High"),
  SVM = c("Low", "Low", "Moderate")
)
# Create a nicely formatted table with a title
table_with_title <- kable(decision_criteria, 
                          format = "markdown", 
                          align = c("l", "c", "c", "c"),
                          caption = "Decision Criteria Beyond Performance Metrics")

# Print the table with title
cat("# Decision Criteria Beyond Performance Metrics\n\n")
print(table_with_title)
# Print the data frame
print(decision_criteria)

```

***Alignment with Scientific Question/Hypothesis***

Interpretability is crucial if the goal is to understand how average age of finalists varies by Season.

Choice of the Best Model: Support Vector Machine (SVM)
Reasons:

Performance: Lowest RMSE and highest R² across all repeats.
Robustness: Handles complex relationships well when properly tuned.
Scientific Hypothesis: Superior choice if predictive accuracy is prioritized over interpretability.

Conclusion
Despite lower interpretability, the SVM model's superior predictive performance makes it the best choice for predicting age based on state data in this specific task.
Final Assessment
The Support Vector Machine (SVM) model will be used for the final assessment based on its overall performance and suitability for the task at hand.

# Final Model Performance on Test Data

```{r}
set.seed(123)

# Define the recipe
age_recipe <- recipe(Age ~ State, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

# Define the SVM model
svm_model <- svm_rbf() %>%
  set_engine("kernlab") %>%
  set_mode("regression")

# Create the workflow
svm_workflow <- workflow() %>%
  add_recipe(age_recipe) %>%
  add_model(svm_model)

# Fit the model on the entire training data
final_svm_fit <- svm_workflow %>%
  fit(data = train_data)

# Make predictions on the test data
test_predictions <- predict(final_svm_fit, new_data = test_data) %>%
  bind_cols(test_data)
# Calculate performance metrics
svm_rmse <- rmse(test_predictions, truth = Age, estimate = .pred)
svm_rsq <- rsq(test_predictions, truth = Age, estimate = .pred)

```

```{r}
# Assuming your code has been run and svm_rmse and svm_rsq are calculated

# Create a data frame for the results
final_results <- data.frame(
  Metric = c("RMSE", "R-squared"),
  Value = c(svm_rmse$.estimate, svm_rsq$.estimate)
)

# Round the values to 4 decimal places
final_results$Value <- round(final_results$Value, 4)

# Print the basic table
print("Final Model Performance on Test Data:")
print(final_results)

# Create a nicely formatted table
cat("\nFormatted Table:\n")
kable(final_results, 
      caption = "Final SVM Model Performance on Test Data",
      col.names = c("Metric", "Value"),
      align = c("l", "c"),
      format = "markdown")
```

```{r}
library(ggplot2)

ggplot(test_predictions, aes(x = .pred, y = Age)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "blue") +  # Set line color to blue
  labs(title = "Predicted vs Actual Ages", x = "Predicted Age", y = "Actual Age") +
  theme(
    panel.background = element_rect(fill = "#F0FFF0"),  # Set background color to light melon
    plot.background = element_rect(fill = "#F0FFF0")
  )

```

```{r}
library(ggplot2)

ggplot(test_predictions, aes(x = .pred, y = .pred - Age)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "blue") +  # Set horizontal line color to blue
  labs(title = "Residuals Plot", x = "Predicted Age", y = "Residuals") +
  theme(
    panel.background = element_rect(fill = "#F0FFF0"),  # Set background color to light melon
    plot.background = element_rect(fill = "#F0FFF0")
  )
```
# 7

```{r}
# Create a data frame for the summary
summary_table <- data.frame(
  Section = c("Objective", "Methodology", "Findings", "Conclusion"),
  Description = c(
    "Build and evaluate predictive models to estimate if average age of finalists varies by Season.",
    "Data preparation, model training with cross-validation, model evaluation, and final assessment of chosen model (SVM).",
    "SVM outperformed Linear Regression and Random Forest in cross-validation. SVM showed good performance on test data with lower RMSE and higher R².",
    "SVM chosen as best model due to superior performance. Balance between performance, interpretability, and efficiency is crucial."
  )
)

# Create a data frame for the cross-validation results
cv_results <- data.frame(
  Model = c("Linear Regression", "Random Forest", "SVM"),
  RMSE_Mean = c(7.499, 7.132, 6.964),
  R_Squared_Mean = c(0.0012, 0.0018, 0.0150)
)

# Print the summary table
cat("Summary of Analysis:\n")
print(knitr::kable(summary_table, format = "pipe"))

# Print the cross-validation results
cat("\nCross-Validation Results:\n")
print(knitr::kable(cv_results, format = "pipe", digits = 4))
```

```{r}
# Create a data frame for the summary results
cv_summary <- data.frame(
  Model = c("Linear Regression", "Random Forest", "SVM"),
  RMSE_Mean = c(7.499, 7.132, 6.964),
  R_Squared_Mean = c(0.0012, 0.0018, 0.0150)
)

# Round the values to 4 decimal places
cv_summary$RMSE_Mean <- round(cv_summary$RMSE_Mean, 4)
cv_summary$R_Squared_Mean <- round(cv_summary$R_Squared_Mean, 4)

# Print the basic table
print("Summary Table of Cross-Validation Results:")
print(cv_summary)

# If you want a nicer looking table, you can use the knitr package
if (!require(knitr)) install.packages("knitr")
library(knitr)

# Create a nicely formatted table
cat("\nFormatted Table:\n")
kable(cv_summary, 
      caption = "Summary Table of Cross-Validation Results",
      col.names = c("Model", "RMSE (Mean)", "R² (Mean)"),
      align = c("l", "c", "c"),
      format = "markdown")
```

This summary provides a concise overview of your analysis, including the objective, methodology, key findings, and conclusion. The cross-validation results table offers a quick comparison of the performance of the three models you evaluated.

# Residual Plot for SVM Model on Test Data

```{r}
library(ggplot2)

ggplot(test_predictions, aes(x = .pred, y = Age)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "blue") +  # Set line color to blue
  labs(title = "Residual Plot for SVM Model",
       x = "Predicted Age",
       y = "Actual Age") +
  theme(
    panel.background = element_rect(fill = "#F0FFF0"),  # Set background color to light melon
    plot.background = element_rect(fill = "#F0FFF0")
  )
```
This study exemplifies a thorough methodology for assessing and choosing predictive models, highlighting the critical role of cross-validation and in-depth model evaluation. Our results indicate that while there is a discernible relationship between season and age, it is relatively modest, as evidenced by the low R² values across all models tested.
The systematic comparison of linear regression, random forest, and support vector machine (SVM) models revealed varying levels of predictive accuracy. The SVM model emerged as the top performer, suggesting that the relationship between season and age may involve complex, non-linear patterns that simpler models struggle to capture.
Despite the SVM model's superior performance, the overall low R² values across all models indicate that season alone is a weak predictor of age. This underscores the complexity of the relationship and suggests that other factors, not included in this analysis, may play significant roles in determining the age of participants.
These findings emphasize the importance of careful model selection and the need to consider both statistical performance and practical interpretability when choosing a predictive model. While more sophisticated models like SVM can offer improved predictive power, the trade-off between complexity and interpretability should be carefully weighed in the context of the specific research or business objectives.
In conclusion, this analysis not only provides insights into the relationship between season and age but also serves as a template for rigorous model evaluation in predictive analytics. It underscores the value of employing diverse modeling techniques and thorough validation processes to ensure robust and reliable predictions.

### 2024-07-29 ManuscriptGrp10C07.28.2024.qmd
final draft manuscript

--- 2024-06-24 docs/ManuscriptGrp10.qmd
@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/apa.csl
 ---
@@
 library(knitr)
+library(ggplot2)
+library(dplyr)
+library(scales)
+library(showtext)
+library(ggimage)
 ```
@@
 # Introduction
+
+```{r}
+# Create a data frame with the composition of honey
+honey_data <- data.frame(
+  component = c("Fructose", "Glucose", "Water", "Maltose", "Trisaccharides, Carbohydrates", "Sucrose", "Minerals, Vitamins, Enzymes"),
+  percentage = c(38.5, 31.0, 17.1, 7.2, 4.2, 1.5, 0.5)
+)
+
+# Define colors for the segments
+honey_data$color <- c("#FFA726", "#FB8C00", "#FFD54F", "#FFB74D", "#90CAF9", "#F06292", "#BA68C8")
+
+# Plot with ggplot2
+ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
+  geom_bar(width = 0.8, stat = "identity", color = "white") +
+  coord_polar("y", start = 0) +
+  scale_fill_manual(values = honey_data$color) +
+  geom_text(aes(x= 1.8,label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
+  labs(
+    title = "COMPOSITION OF HONEY",
+    subtitle = "Illustration of honey components by percentage",
+    fill = NULL
+  ) +
+  theme_void() +
+  theme(
+    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
+    plot.subtitle = element_text(size = 14, hjust = 0.5),
+    legend.position = "bottom",
+    legend.text = element_text(size = 12)
+  )
+```
 
@@
 
-## General Background Information
+You can add options to executable code like this
+
+```{r}
+library(ggplot2)
+library(sf)
+library(maps)
+```
+
+```{r}
+states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
+
+```
+
+```{r}
+varroa_data <- data.frame(
+  state = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
+            "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
+            "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
+            "maine", "maryland", "massachusetts", "michigan", "minnesota", 
+            "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
+            "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
+            "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", 
+            "south carolina", "south dakota", "tennessee", "texas", "utah", 
+            "vermont", "virginia", "washington", "west virginia", "wisconsin", "wyoming"),
+  year = c("1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
+           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995")
+)
+
+```
+
+```{r}
+states <- merge(states, varroa_data, by.x = "ID", by.y = "state", all.x = TRUE)
+
+```
+
+```{r}
+ggplot(data = states) +
+  geom_sf(aes(fill = year), color = "#292929", size = 0.2) +
+  scale_fill_manual(
+    values = c(
+      "1986 - 1987" = "#CD2626", 
+      "1988 - 1989" = "#CD9B1D", 
+      "1990 - 1991" = "#FFD700", 
+      "1992 - 1993" = "#00CD00", 
+      "1994 - 1995" = "#104E8B"
+    ),
+    na.value = "#757575"
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Spread of Varroa Mites",
+    fill = "Year of First Appearance"
+  )
+
+```
+
+Written By:sdns6mchl4.
+(2016, February 24).
+Varroa mite spread in the United States.
+Beesource Beekeeping Forums.
+https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/ \## General Background Information
+
+[**Uncapped Honey Floresville,Texas Hive**]{.underline}
+
+![](images/Bee%20Frame%202024-01.jpg){width="674"}
+
+[**Capped Honey one Month Later same hive frame- Italian bees Floresville, Texas**]{.underline}
+
+![](images/Bee%20Frame%20Capped%20Honey.jpg){fig-align="center" width="697"}
 
@@
 Hypothesis: "The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.
+
+[**Bacterium Infection Foul Brood.**]{.underline}
+
+![](images/American-Foul-Brood-Comb-1024x768-1.jpg)
+
+[**Dead bees resulting from extreme heat found in hive**]{.underline}.
+
+![](images/dead-bees-hive-colony-collapse-600nw-2289939105.webp){width="679"}
 
@@
 #| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
+#| fig-cap: "A figure that is manually genera#ted and shows some overview/schematic. This has nothing to do with the data, it's just a random one #from one of our projects I found and placed here."
 #| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
+#knitr::include_graphics(here("assets","antigen-recognition.png"))
 ```
@@
 library(tidyverse)
-
+library(ggplot2)
+library(knitr)
 ```
@@
 
-data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/DatabyState.csv")
+data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/DatabyState.csv")
+```
 
+```{r}
 View(data)
@@
 ```{r}
-
 # Select all columns except 3, 6, and 9
 Data_Clean <- dplyr::select(data, -c(3, 6, 9))
-
 ```
@@
 ```{r}
-
 # Output cleaned data file to a csv file.
-write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")
-
-view(Data_Clean)
-
+write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/Databystate_Clean.csv")
 ```
@@
 ```{r}
-library(readxl)
-library(tidyverse)
-
+view(Data_Clean)
 ```
@@
 ```{r}
-
-data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/DatabyState.csv")
-
+data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/DatabyState.csv")
 ```
@@
 ```{r}
-
 # Output cleaned data file to a csv file.
-write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")
-
+write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/Databystate_Clean.csv")
 ```
@@
 ```{r}
-
 # Assuming your data frame is named "data"
-
 # Filter for rows where "Loss" or "Deadout" is present in any column (case-insensitive)
@@
 data_filtered <- Data_Clean[rowSums(sapply(data, grepl, pattern = c("Loss"), ignore.case = TRUE)) > 0, ]
-
 ```
@@
 ```{r}
+# Output cleaned data file to a csv file.
+write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/Databystate_Filtered.csv")
+```
 
-# Output cleaned data file to a csv file.
-write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Filtered.csv")
-
+```{r}
+view(Data_Clean)
 ```
@@
 ```
+STATE ANSI IS THE CODE FOR STATE BY PERIOD AND YEAR. VALUE IS COUNT.
 
+The above data set is organized by year, state, period of the year, type of loss, and count.
+There are only three observed years, with observations from 2015 to 2017.
+Loss by collapse disorder describes a colony that losses a majority of its worker bees.
+Loss by deadout describes a loss of the entire colony: workers, drones, larvae, pupa, and queen.
+The bar chart shown below depicts loss count by type and period.
+The periods roughly follow North American seasons; April through June is spring, January through March is winter, July through September is summer, and October through December is fall.
+We can observe a pattern in the losses increasing in the first half of the year (January through June), while losses decrease the second half.
+This could suggest that as the weather warms, colonies are affected by the increasing heat.
 ```{r}
@@
 ```
-
+The above data set is organized by year, state, period of the year, type of loss, and count.
+There are only three observed years, with observations from 2015 to 2017.
+Loss by collapse disorder describes a colony that losses a majority of its worker bees.
+Loss by deadout describes a loss of the entire colony: workers, drones, larvae, pupa, and queen.
+The bar chart shown below depicts loss count by type and period.
+The periods roughly follow North American seasons; April through June is spring, January through March is winter, July through September is summer, and October through December is fall.
+We can observe a pattern in the losses increasing in the first half of the year (January through June), while losses decrease the second half.
+This could suggest that as the weather warms, colonies are affected by the increasing heat
 ```{r}
@@
 library(readxl)
-update.packages("tidyverse")
 library(tidyverse)
-update.packages("dplyr")
 library(dplyr)
-update.packages('mgcv')
 library(mgcv)
@@
 # Output cleaned data file to a csv file.
-hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)
+hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)
 ```
@@
 ```
-
+This gives us the properties of out data set
 ```{r}
@@
 
-```{r}
+```{r warning=FALSE}
 # Convert specified columns to numeric
@@
 ```
-
+The above indicated the count of Varroa Mites in our data
+From our project, we found that varroa mites are the top cause for colony loss.
+Four states had the highest levels of varroa cases and were chosen for comparison, in the United States: Wisconsin, Texas, Ohio, and Pennsylvania.
+The box plot below shows one outlier in this top 5 areas, with the average being just above 50.
 ```{r}
@@
 ```
-
+The above plot illustrates the comparison between American and European Foul Brood and Varroa Mites
 ```{r}
@@
 # Write the cleaned data to a new CSV file
-write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/hcnydata_cleaned.csv", row.names = FALSE)
+write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcnydata_cleaned.csv", row.names = FALSE)
 ```
 
-\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*Conclusion- all code below is test code right now\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*
+```{r}
+# Output cleaned data file to a csv file.
+hcny_data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", header = FALSE, stringsAsFactors = FALSE)
+```
 
+```{r}
+head(hcny_data)
+```
+
+```{r}
+str(hcny_data)
+```
+
+```{r}
+summary(hcny_data)
+```
+
+```{r}
+# Step 2: Merge the first two rows to create a new header
+new_header <- paste(hcny_data[1, ], hcny_data[2, ], sep = " ")
+
+# Step 3: Set the new header
+colnames(hcny_data) <- new_header
+
+# Step 4: Remove the first three rows
+hcny_data <- hcny_data[-c(1:3), ]
+
+# Step 5: Remove completely blank rows
+hcny_data <- hcny_data[rowSums(hcny_data != "") > 0, ]
+
+# Display the resulting data frame
+head(hcny_data)
+```
+
+```{r}
+# Assuming hcny_data is your cleaned data frame
+colnames(hcny_data) <- c("state", "varroa_mites", "other_pests", "disease", "pesticides", "other", "unknown")
+```
+
+```{r}
+head(hcny_data)
+```
+
+```{r}
+# Convert specified columns to numeric with warning suppression
+hcnydata_cleaned <- hcny_data %>%
+  mutate(across(c("varroa_mites", "other_pests", "disease", "pesticides", "other", "unknown"), ~suppressWarnings(as.numeric(.))))
+
+# Display the resulting data frame
+head(hcnydata_cleaned)
+```
+
+```{r}
+head(hcnydata_cleaned)
+```
+
+```{r}
+# Verify the changes
+str(hcnydata_cleaned)
+```
+
+```{r}
+head(hcnydata_cleaned)
+```
+
+```{r,message=FALSE, warning=FALSE}
+ggplot(hcnydata_cleaned, aes(varroa_mites)) + geom_histogram()
+```
+
+![](images/_20220406_on_varroadestructor_minden_00463080.jpg){fig-align="center" width="541"}
+
+```{r}
+view((hcnydata_cleaned))
+```
+
+```{r}
+# Write the cleaned data to a new CSV file
+write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", row.names = FALSE)
+```
+
+```{r}
+# Assuming hcnydata_cleaned is your data frame
+hcnydata_cleaned <- hcnydata_cleaned[apply(hcnydata_cleaned, 1, function(x) !all(is.na(x))), ]
+
+# Print the modified data frame to verify the changes
+head(hcnydata_cleaned)
+view(hcnydata_cleaned)
+```
+
+```{r}
+str(hcnydata_cleaned)
+```
+
+```{r, message=FALSE}
+#update.packages('mgcv')
+library(mgcv)
+```
+
+```{r}
+# Convert the 'state' column to a factor
+hcnydata_cleaned$state <- as.factor(hcnydata_cleaned$state)
+
+# Display the resulting data frame
+head(hcnydata_cleaned)
+```
+
+```{r}
+# Step 3: Reindex the rows starting from 1
+rownames(hcnydata_cleaned) <- NULL
+```
+
+```{r}
+# Identify the rows to exclude
+rows_to_exclude <- c(46)
+
+# Remove rows 45 to 52 while excluding row 46
+data_hcny <- hcnydata_cleaned[-c(45:52)[-which(c(45:52) %in% rows_to_exclude)], ]
+```
+
+```{r}
+library(ggplot2)
+
+# Create a bar graph to show state and varroa_mites
+ggplot(data = data_hcny, aes(x = state, y = varroa_mites)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  labs(title = "Varroa Mites by State",
+       x = "State",
+       y = "Varroa Mites Count") +
+  theme_minimal() +
+  coord_flip()
+```
+T
+Varroa Mite
+
+![](images/Varrpa%201.webp){fig-align="center" width="307"}
+
+![](images/varroa-mite-bee.webp){fig-align="center" width="286"}
+The above image shows the count of Varroa Mites by States.Colorado and North Dakota comes highest in the plot with Oklahoma and New Mexico the lowest in count.
+This indicated that we have more of Varroa Mites in North Dakota and Colorado compared to Oklahoma and New Mexico
 ## Statistical analysis
@@
 
+[**`Texas Bee gathering pollen from natural Texas foliage`**]{.underline}
+
+![](images/Bee%20Feeding.jpg){fig-align="center" width="609"}
+
 ## Exploratory/Descriptive analysis
@@
 Even more can be in the R and Quarto files that are part of your project.*
+
+# Remove all rows with any NA values
+
+```{r}
+
+library(tidyr)
+library(dplyr)
+
+my_data <- data_hcny %>% drop_na()
+
+# Verify that NAs have been removed
+glimpse(my_data)
+```
+
+# Bivariate Analysis
+
+```{r, message=FALSE, warning=FALSE}
+## Example of a bivariate plot between 'state' and 'varroa_mites'
+ggplot(my_data, aes(x = disease, y = varroa_mites)) +
+  geom_point() +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(title = "Bivariate Relationship between State and Varroa Mites",
+       x = "State",
+       y = "Varroa Mites")
+```
+From the above graph we see that Varroa Mites concentrated more in certain states than others
+# Multivariable GLM
+
+```{r}
+#install.packages("tidymodels")
+library(tidymodels)
+library(broom)
+library(stats)
+library(MASS)
+# Specify the GLM model
+glm_spec <- linear_reg() %>%
+  set_engine("glm")
+
+# Create a recipe for preprocessing the data
+glm_recipe <- recipe(varroa_mites ~ other_pests + disease + pesticides + other + unknown, data = my_data) %>%
+  step_normalize(all_predictors())
+
+# Create a workflow
+glm_workflow <- workflow() %>%
+  add_model(glm_spec) %>%
+  add_recipe(glm_recipe)
+
+# Fit the model
+glm_fit <- fit(glm_workflow, data = my_data)
+
+# Print the model summary using tidy()
+model_summary <- tidy(glm_fit)
+print(model_summary)
+```
 
@@
 
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
-
 ## Basic statistical analysis
@@
 @fig-result shows a scatterplot figure produced by one of the R scripts.
-
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
+plot <- ggplot(my_data, aes(x = varroa_mites , y =  other_pests + disease + pesticides + other + unknown )) +
+  geom_point(color = "blue") +              # Scatter plot
+  geom_smooth(method = "lm", color = "red") + # Regression line
+  ggtitle("Scatter Plot with Linear Regression Line") +
+  xlab("varroa_mites") +
+  ylab("other_pests + disease + pesticides + other + unknown")
+print(plot)
 ```
+The above diagram indicates that varroa mites are proportionally related to other psests, diseases, pesticides and other unknown parasite
 
@@
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
+saveRDS(my_data, file = "my_data.rds")
 ```
+
+```{r}
+file.exists("my_data.rds")
+```
+
+```{r}
+getwd()
+```
+
+```{r}
+loaded_data <- readRDS("my_data.rds")
+print(loaded_data)
+```
+
+```{r}
+my_data <- readRDS("my_data.rds")
+
+```
+
+```{r}
+model <- lm(varroa_mites ~ other_pests + disease + pesticides + other + unknown, data = my_data)
+```
+
+```{r}
+summary(model)
+```
+From the linear regression  we see that only 49% of variation in the data can be explained by varroa mites, The p value is 0.01 which is statistically significant
+```{r}
+library(broom)
+library(knitr)
+```
+
+```{r}
+model_tidy <- tidy(model)
+```
+
+```{r}
+kable(model_tidy, caption = "Linear Model Fit Table")
+```
+The p value for varroa Mites is statistically significant at 0.009 when compared with other disease or pesticides.
 

### 2024-07-31 manuscript-exercise/manuscript_final-2.qmd
Manuscript Project Draft

---
title: "Honey Bee Colonies Impacted by Varroa, American Fouldbrood and Global Warming"
author: Edward Cruz, Jr, Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
format: html
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load a few R packages
library(here)
library(knitr)
library(ggplot2)
library(dplyr)
library(scales)
library(showtext)
library(ggimage)
```

# Summary/Abstract

Group Ten is conducting a comprehensive analysis of historical data from
various public agencies to evaluate the impact of Varroa mites, American
Foulbrood, and global warming on hive losses in the United States.This
research utilizes extensive datasets from the National Agricultural
Statistics Service, the Agricultural Statistics Board, and the United
States Department of Agriculture (USDA), encompassing multiple years of
data. The data highlights the topic on hive losses attributed to mites,
bacterial infections, and environmental factors related to global
warming. Through advanced data visualization techniques in R, we aim to
demonstrate and validate the detrimental effects of these factors on
honeybee colonies, highlighting the consequent implications for honey
production and broader food security.

{{< pagebreak >}}

# Introduction

Honeybees have served as nature's pollinators for centuries. With their
relationship with humans documented as far back as ancient Egyptian and
Hindu cultures. Historically, humans have maintained beehives and
utilized honey for its medicinal properties in various civilizations,
including the Egyptians, Assyrians, Chinese, Greeks, and Romans. Honey's
natural antibacterial qualities made it a valuable treatment for wounds.
This was a practice continued by Romans and Russians during World War I.
Honeybees and other pollinators are essential for food production and
nutritional security. But even though they face numerous survival
challenges Varroa mites pose a significant threat to bee colonies. These
tiny red-brown parasites live on adult honeybees and reproduce on larvae
and pupae. Another major threat is American Foulbrood Disease (AFB).
This is caused by the bacterium Paenibacillus larvae, which is fatal to
honeybee larvae and requires incineration of infected hives to prevent
its spread. The impact of climate change on honeybee colony losses is a
growing area of research. While correlations have been found between
higher winter temperatures and increased colony losses, the effects of
warmer autumn and winter temperatures on colony dynamics and survival
require further investigation.

### Composition of Honey

The first image is a pie chart that illustrates the composition of honey
by percentage. The main components are:

-   **Fructose**: 38.5%
-   **Glucose**: 31%
-   **Water**: 17.1%
-   **Maltose**: 7.2%
-   **Other components**:
    -   Sucrose: 1.5%
    -   Minerals, Vitamins, Enzymes: 0.5%
    -   Trisaccharides, Carbohydrates: 4.2%

This composition indicates that honey is primarily made up of sugars,
specifically fructose and glucose, with water being the third major
component. The presence of minerals, vitamins, and enzymes, though in
smaller amounts, adds nutritional value to honey. This detailed
breakdown underscores honey's role as a natural sweetener with
additional health benefits beyond its primary carbohydrate content.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create a data frame with the composition of honey
honey_data <- data.frame(
  component = c("Fructose", "Glucose", "Water", "Maltose", "Trisaccharides, Carbohydrates", "Sucrose", "Minerals, Vitamins, Enzymes"),
  percentage = c(38.5, 31.0, 17.1, 7.2, 4.2, 1.5, 0.5)
)

# Define colors for the segments
honey_data$color <- c("#FFA726", "#FB8C00", "#FFD54F", "#FFB74D", "#90CAF9", "#F06292", "#BA68C8")

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)
library(maps)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
varroa_data <- data.frame(
  state = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
            "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
            "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
            "maine", "maryland", "massachusetts", "michigan", "minnesota", 
            "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
            "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
            "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", 
            "south carolina", "south dakota", "tennessee", "texas", "utah", 
            "vermont", "virginia", "washington", "west virginia", "wisconsin", "wyoming"),
  year = c("1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
states <- merge(states, varroa_data, by.x = "ID", by.y = "state", all.x = TRUE)
```

### Spread of Varroa Mites

The second image is a map illustrating the spread of Varroa mites in the
United States from 1986 to 1995. The map uses different colors to
represent the year of the first appearance of Varroa mites in each
state:

-   **1986 - 1987**: Red
-   **1988 - 1989**: Yellow
-   **1990 - 1991**: Green
-   **1992 - 1993**: Blue
-   **1994 - 1995**: Grey
-   **NA**: States where data is not available

The map reveals the progressive spread of Varroa mites from the
mid-1980s to the mid-1990s, affecting bee colonies across the country.
The earliest appearances were concentrated in specific regions and
gradually spread to more states over time. This visual representation
highlights the increasing geographical distribution of Varroa mites and
underscores the growing challenge they pose to beekeeping and
agricultural industries in the United States.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = states) +
  geom_sf(aes(fill = year), color = "#292929", size = 0.2) +
  scale_fill_manual(
    values = c(
      "1986 - 1987" = "#CD2626", 
      "1988 - 1989" = "#CD9B1D", 
      "1990 - 1991" = "#FFD700", 
      "1992 - 1993" = "#00CD00", 
      "1994 - 1995" = "#104E8B"
    ),
    na.value = "#757575"
  ) +
  theme_minimal() +
  labs(
    title = "Spread of Varroa Mites",
    fill = "Year of First Appearance"
  )

```

```{r}
# Load necessary libraries
library(tidyverse)

# Create a data frame with the provided data
data <- tibble(
  Year = 1901:2023,
  `Earth's surface` = c(-0.15, -0.43, -1.4, -0.86, -1.02, -0.29, -0.54, 0.06, -0.59, 0.4, 0.01, -1.79, -0.48, -0.18, -0.57, -1.17, -1.96, -0.15, -0.47, -0.95, 1.78, 0.01, -0.38, -1.43, 0.5, -0.07, 0.13, -0.1, -1.17, -0.04, 1.52, -0.29, 0.97, 2.08, -0.12, 0.13, -0.47, 1.16, 1.24, -0.13, 0.64, -0.18, 0.05, -0.19, -0.27, 0.93, -0.1, -0.41, 0, -0.63, -0.9, 0.25, 1.35, 1.31, -0.33, 0.32, 0.02, -0.09, 0.09, -0.58, -0.15, -0.12, 0.24, -0.35, -0.33, -0.53, -0.26, -0.7, -0.52, -0.41, -0.36, -0.65, 0.27, 0.24, -0.52, -0.55, 0.53, -0.97, -1.14, 0.37, 1.1, -0.67, -0.14, -0.04, -0.72, 1.3, 1.31, 0.61, -0.18, 1.49, 1.14, 0.58, -0.76, 0.85, 0.63, -0.13, 0.18, 2.21, 1.86, 1.25, 1.68, 1.19, 1.24, 1.08, 1.62, 2.23, 1.63, 0.27, 0.37, 0.96, 1.16, 3.26, 0.41, 0.52, 2.38, 2.9, 2.53, 1.5, 0.66, 2.35, 2.49, 1.37, 2.39),
  `Lower troposphere (UAH)` = c(rep(NA, 78), 0.049093434, 0.731593434, 1.102093434, 0.008593434, 0.034093434, 0.292093434, 0.110593434, 1.220593434, 1.054093434, 0.698593434, 0.548593434, 1.451593434, 1.031593434, 0.484093434, -0.174406566, 1.043593434, 0.797593434, 0.502093434, 0.457093434, 1.691593434, 1.796593434, 1.351093434, 1.255093434, 0.994093434, 1.223593434, 0.769093434, 1.303093434, 1.444093434, 1.765093434, 0.467593434, 0.524593434, 0.929593434, 1.013593434, 1.955593434, 0.655093434, 0.763093434, 2.234593434, 2.030593434, 2.252593434, 1.645093434, 1.133593434, 1.970593434, 2.059093434, 1.567093434, NA),
  `Lower troposphere (RSS)` = c(rep(NA, 78), -0.808971111, 0.178178889, 0.658778889, -0.579771111, -0.383871111, -0.217371111, -1.053021111, 0.737978889, 1.111328889, 0.557978889, 0.221828889, 1.280228889, 1.043228889, 0.585278889, -0.261321111, 1.078178889, 0.772628889, 0.278078889, 0.418178889, 1.911578889, 1.882928889, 1.471328889, 1.536578889, 1.144628889, 1.465628889, 1.058528889, 1.583078889, 1.749278889, 1.858928889, 0.503378889, 0.595928889, 0.997028889, 1.109378889, 2.677328889, 0.696278889, 0.781478889, 2.547878889, 2.476328889, 2.350478889, 1.793978889, 1.221428889, 2.290628889, 2.485328889, 1.596878889, 1.886228889)
)
# Pivot the data for plotting
data_long <- data %>%
  pivot_longer(cols = c(`Earth's surface`, `Lower troposphere (UAH)`, `Lower troposphere (RSS)`), 
               names_to = "Measurement", values_to = "Temperature")

# Create the plot
ggplot(data_long, aes(x = Year, y = Temperature, color = Measurement)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  theme_minimal() +
  labs(
    title = "Temperature Changes Over the Years",
    x = "Year",
    y = "Temperature Anomaly (°C)",
    color = "Measurement"
  ) +
  scale_color_manual(values = c("Earth's surface" = "#254976", "Lower troposphere (UAH)" = "#4A8C1D", "Lower troposphere (RSS)" = "#B5153A"))

```

Written By:sdns6mchl4. (2016, February 24). Varroa mite spread in the
United States. Beesource Beekeeping Forums.
https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/
\## General Background Information

### **Uncapped Honey Floresville,Texas Hive**

![](images/Bee%20Frame%202024-01.jpg){width="674"}

### **Capped Honey one Month Later same hive frame- Italian bees Floresville, Texas**

![](images/Bee%20Frame%20Capped%20Honey.jpg){fig-align="center"
width="697"}

## Description of data and data source

Bee colonies maintained by beekeepers are considered livestock by the
USDA due to their ability to produce honey, a consumable food item, and
their essential role in assisting farmers with pollination crop seasons.
Given the importance of bee colonies in agriculture, it was logical to
source data from the following two authoritative websites: 1. USDA
National Agricultural Statistics Service (NASS): This site provides
comprehensive agricultural data, including statistics on honey
production and colony health. 2. Bee Informed Partnership: This site
offers detailed insights and research on bee colony management and
health, contributing valuable information on the status and trends of
bee populations. Index Catalog // USDA Economics, Statistics and Market
Information System. (n.d.-a).
https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
USDA - National Agricultural Statistics Service - Surveys - honey bee
surveys and reports. (n.d.).
https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/

## Questions/Hypotheses to be addressed

Hypothesis: "The negative impacts of mites, bacterium, and global
warming have detrimental effects on honeybee colonies in the United
States and Texas, which in turn will lead to a decline in honey
production and negatively impact food production." This hypothesis can
be tested and validated through a visualization of outcomes using R,
demonstrating the relationship between these factors and their effects
on honeybee colonies.

### **Bacterium Infection Foul Brood**

![](images/American-Foul-Brood-Comb-1024x768-1.jpg)

### **Dead bees resulting from extreme heat found in hive**

![](images/dead-bees-hive-colony-collapse-600nw-2289939105.webp){width="679"}

To cite other work (important everywhere, but likely happens first in
introduction), make sure your references are in the bibtex file
specified in the YAML header above and have the right bibtex key. Then
you can include like this:

Examples of reproducible research projects can for instance be found in
[@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning
processes, and the analysis approaches. You might want to provide a
shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not
created with code (if you can do it with code, do it). @fig-schematic is
an example of some - completely random/unrelated - schematic that was
generated with Biorender. We store those figures in the `assets` folder.

```{r}

#knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

We got our data from the United States Department of Agriculture (USDA).

## Data import and cleaning

We decided to clean out our data from a few different datasets. We had
to remove blank spaces and columns that were not pertinent to our
analysis. We then filtered out other observations that did not directly
deal with the data we are exploring. We are looking for cause of death
to bee colonies and how they are affected by mites and climate change so
we wanted to single out data that represented the losses so we can
explore the different states by year and determine how the colonies were
affected.

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(knitr)
```

```{r}

library(here)

# Install and load the here package
#install.packages("here")
library(here)

here()
# Create a 'data' directory within the project
dir.create(here("data"))

# Read the CSV file
data <- read.csv(here("Databystate.csv"))

#data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/DatabyState.csv")
```

```{r}
# Select all columns except 3, 6, and 9
Data_Clean <- dplyr::select(data, -c(3, 6, 9))
```

```{r}
# Output cleaned data file to a csv file.

write.csv(Data_Clean, "Data_Clean.csv", row.names = FALSE)

#write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")
```

```{r}
# Assuming your data frame is named "data"

# Filter for rows where "Loss" or "Deadout" is present in any column (case-insensitive)

library(stringr)  # Load stringr package for regular expressions
data_filtered <- Data_Clean[rowSums(sapply(data, grepl, pattern = c("Loss"), ignore.case = TRUE)) > 0, ]
```

```{r}
# Output cleaned data file to a csv file.

write.csv(data_filtered, "Data_Filtered.csv", row.names = FALSE)

#write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Filtered.csv")
```

```{r}
# View data str to see what class they are.

str(data_filtered)
```

### Boxplot of Value by Year and Data Item

The third image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" from 2015 to 2017. The boxplot
provides insights into: - The central tendency and spread of losses due
to both disorders. - Presence of outliers indicates that there are
occasional, significantly higher losses. - Comparative analysis shows
that both types of losses have varying distributions each year.

This indicates that while both types of losses are present each year,
the magnitude and frequency of extreme loss events can differ,
suggesting variability in the factors contributing to colony losses.

```{r}

# Load necessary libraries
library(ggplot2)
library(here)

str(data_filtered)

# Load the dataset
data_filtered <- read.csv(here::here("Data_Filtered.csv"), stringsAsFactors = FALSE)

# Convert 'Value' to numeric
data_filtered$Value <- as.numeric(gsub(",", "", data_filtered$Value))

# Handle missing values by replacing NAs with the median value
data_filtered$Value[is.na(data_filtered$Value)] <- median(data_filtered$Value, na.rm = TRUE)

# Boxplot of Value by Year and Data.Item
ggplot(data_filtered, aes(x = factor(Year), y = Value, fill = Data.Item)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boxplot of Value by Year and Data.Item", x = "Year", y = "Value") +
  theme_minimal()
```

### Scatterplot of Value by State ANSI

The fourth image is a scatterplot showing values for different states
(ANSI coded) across the years 2015 to 2017. Key points include: - Each
color represents a different year (2015: Red, 2016: Green, 2017:
Blue). - The scatterplot reveals the distribution of values across
states and highlights any trends or patterns over the three years.

This scatterplot indicates that while some states experience
consistently high or low values, others show significant variability
year over year. This could suggest regional differences in factors
affecting bee colony health.

```{r}
# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()

```

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

```{r}
# Load necessary library
library(ggplot2)
library(scales)

# Assuming data_filtered is already loaded and has the required columns

# Create the plot
p <- ggplot(data_filtered, aes(x = Period, y = Value, fill = Data.Item)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boxplot of Value by Period and Data.Item", x = "Period", y = "Value")

# Print the plot
print(p)

```

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

### Overall Analysis

-   The composition of honey indicates its high sugar content, essential
    for energy but also highlighting the presence of beneficial
    components like enzymes and vitamins.
-   The spread of Varroa mites map highlights the widespread and growing
    impact of these pests over time, emphasizing the need for ongoing
    management and control efforts.
-   The boxplots and scatterplots provide a detailed view of the
    variability and trends in colony losses due to different factors,
    both over time and across different regions. These plots suggest
    that while losses are a consistent issue, their magnitude and causes
    can vary widely, pointing to the need for tailored strategies to
    address colony health.

These visualizations collectively provide a comprehensive overview of
the challenges faced by bee colonies, from composition and health to
external threats like Varroa mites and seasonal variations in losses.

### **VARROA MITE EXPOSED**

![](images/varroa-mite-bee-01.webp){fig-align="center" width="244"}

### **VARROA MITE**

![](images/Varrpa%201-02.webp){fig-align="center" width="246"}

\*\*\*\*\*\*\*\*\*\*\*\*\*\* Leo provided additional code below and
works perfectly\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(cluster)
library(car)
library(factoextra)
library(rpart)
library(rpart.plot)
library(randomForest)
library(e1071)

```

```{r}
# Load your dataset
# Assuming your data frame is named 'data_hcny'

data_hcny <- read.csv(here("hcny_CleanDraft.csv"))

#data_hcny <- read.csv("C:/Users/Leonel/Desktop/MSDA/MS Data Analytics/Current Class/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/Group Project/hcnydata_cleaned.csv")

data_clean <- na.omit(data_hcny)

```

```{r}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)
str(data_clean)
```

```{r}
# Visualizations

# Bar Plot for varroa_mites by State
ggplot(data_clean, aes(x = state, y = varroa_mites)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Varroa Mites by State", y = "Percentage", x = "State")
```

```{r}
str(data_clean)
```

```{r}
# Stacked Bar Plot
data_long <- data_clean %>%
  gather(key = "factor", value = "percentage", -state)

ggplot(data_long, aes(x = state, y = percentage, fill = factor)) +
  geom_bar(stat = "identity") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Factors Affecting Bees by State", y = "Percentage", x = "State")
```

```{r}
# Heatmap
data_melt <- melt(data_clean, id.vars = "state")

ggplot(data_melt, aes(x = variable, y = state, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Factors Affecting Bees", y = "State", x = "Factor")
```

```{r}
# Pie Chart for the United States
us_data <- data_clean %>%
  filter(state == "United States") %>%
  dplyr::select(-state) %>%
  gather(key = "factor", value = "percentage")

ggplot(us_data, aes(x = "", y = percentage, fill = factor)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  theme_void() +
  labs(title = "Distribution of Factors Affecting Bees in the United States")
```

```{r}
# Correlation Analysis
cor_data <- cor(data_clean[,-1])
print("Correlation Matrix:")
print(cor_data)
```

```{r}
# Assumption Checks for Descriptive Analysis
print("Assumption Checks:")
```

```{r}
# Normality check
shapiro_results <- lapply(data_clean[-1], function(column) shapiro.test(column))
print("Shapiro-Wilk Test for Normality:")
print(shapiro_results)
```

```{r}
# Homogeneity of variances

levene_results <- leveneTest(varroa_mites ~ state, data = data_clean)
print("Levene's Test for Homogeneity of Variances:")
print(levene_results)
```

```{r}
# Cluster Analysis
data_cluster <- data_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state)

hc <- hclust(dist(data_cluster), method = "complete")
plot(hc, labels = data_clean$state[data_clean$state != "United States"], main = "Hierarchical Clustering of States")
```

```{r}
# Principal Component Analysis (PCA)
data_pca <- data_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state) %>%
  scale()
```

```{r}
pca_result <- prcomp(data_pca, center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r}
fviz_pca_var(pca_result, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

```

```{r}
str(data_clean)
```

```{r}
# Support Vector Machine (SVM)
svm_model <- svm(varroa_mites ~ other_pests + disease + pesticides + other + unknown, data = data_clean)
summary(svm_model)
```

```{r}
# Assuming predictions and data_clean$varroa_mites are already created

# Example data
data_clean <- data.frame(
  Varroa_Mites = c(5, 10, 15, 20),
  other_variable = c(2, 4, 6, 8)
)

# Fit a linear model
model <- lm(Varroa_Mites ~ other_variable, data = data_clean)

# Generate predictions
predictions <- predict(model, newdata = data_clean)

# Plot the data
plot(predictions, data_clean$varroa_mites, 
     xlab = "Predictions", 
     ylab = "Varroa Mites", 
     main = "Predictions vs Varroa Mites",
     las = 1, # makes y-axis labels horizontal
     cex.lab = 1.2, # increases axis labels size
     cex.axis = 1.2) # increases tick labels size

```

### **Honey Bee Collecting Pollen**

![](images/Bee%20Feeding-01.jpg){fig-align="center" width="523"}

### Honey Bee with exposed bloated Varroa Mite

![](images/_20220406_on_varroadestructor_minden_00463080.jpg){fig-align="center"
width="541"}

```{r}
# Write the cleaned data to a new CSV file

# Assuming your dataset is named data_hcny
data_hcny_clean <- data_hcny %>% drop_na()

write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)

#write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", row.names = FALSE)
```

## Full analysis

### 2024-08-01 manuscript-exercise/manuscript_final-2.qmd
2nd HTML

@@
 ![](images/dead-bees-hive-colony-collapse-600nw-2289939105.webp){width="679"}
-
+#Citation section
 To cite other work (important everywhere, but likely happens first in
@@
 [@mckay2020; @mckay2020a].
+
+#Here are the BibTex entries for citations:
+
+Using this in RMarkdown:
+#Make sure to save these entries in a file names'reference.bib' and include it in your R Markdown document like this:
 
@@
 ## Full analysis
+
+#Citations
+```{r}
+library(knitr)
+```
+
+```{r}

### 2024-08-01 manuscript_final-2/manuscript_final-2.qmd
HTML 3

--- 2024-08-01 manuscript-exercise/manuscript_final-2.qmd
@@
 
-![](images/Bee%20Frame%202024-01.jpg){width="674"}
+![](/images/Bee%20Frame%202024.jpg){fig-align="center"}
 
@@
 
-![](images/Bee%20Frame%20Capped%20Honey.jpg){fig-align="center"
-width="697"}
+![](/images/Bee%20Frame%20Capped%20Honey.jpg){fig-align="center"}
 
@@
 
-![](images/American-Foul-Brood-Comb-1024x768-1.jpg)
+![](/images/American-Foul-Brood-Comb-1024x768-1.jpg){fig-align="center"}
 
@@
 
-![](images/dead-bees-hive-colony-collapse-600nw-2289939105.webp){width="679"}
-#Citation section
-To cite other work (important everywhere, but likely happens first in
-introduction), make sure your references are in the bibtex file
-specified in the YAML header above and have the right bibtex key. Then
-you can include like this:
+![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)
+
+#Citation section To cite other work (important everywhere, but likely
+happens first in introduction), make sure your references are in the
+bibtex file specified in the YAML header above and have the right bibtex
+key. Then you can include like this:
 
@@
 
-Using this in RMarkdown:
-#Make sure to save these entries in a file names'reference.bib' and include it in your R Markdown document like this:
+Using this in RMarkdown: #Make sure to save these entries in a file
+names'reference.bib' and include it in your R Markdown document like
+this:
 
@@
 
-![](images/varroa-mite-bee-01.webp){fig-align="center" width="244"}
+![](/images/_20220406_on_varroadestructor_minden_00463080.jpg){fig-align="center"}
 
@@
 
-![](images/Varrpa%201-02.webp){fig-align="center" width="246"}
+![](/images/Varrpa%201.webp){fig-align="center" width="527"}
 
@@
 
-![](images/Bee%20Feeding-01.jpg){fig-align="center" width="523"}
+![](/images/Bee%20Feeding.jpg){fig-align="center"}
 
@@
 
-![](images/_20220406_on_varroadestructor_minden_00463080.jpg){fig-align="center"
-width="541"}
+![](/images/varroa-mite-bee-01.webp){fig-align="center" width="565"}
 
@@
 #Citations
-```{r}
-library(knitr)
-```
-
-```{r}

### 2024-08-01 starter-analysis-exercise/products/report/starter-analysis-report.qmd
HTML4

--- 2024-08-01 manuscript_final-2/manuscript_final-2.qmd
@@
 ---
-title: "Honey Bee Colonies Impacted by Varroa, American Fouldbrood and Global Warming"
-author: Edward Cruz, Jr, Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
-format: html
-editor: 
-  markdown: 
-    wrap: 72
+title: "Manuscript/Report Template for a Data Analysis Project"
+subtitle: ""
+author: Your Name
+date: today
+format:
+  html:
+    toc: false
+    number-sections: true
+    highlight-style: github
+bibliography: ../dataanalysis-template-references.bib
+csl: ../apa.csl
 ---
@@
 #Citations
+
+{{< pagebreak >}}
+
+# References

### 2024-08-01 starter-analysis-exercise/products/report/starter-analysis-report.qmd
HTML5

--- 2024-08-01 manuscript_final-2/manuscript_final-2.qmd
@@
 title: "Honey Bee Colonies Impacted by Varroa, American Fouldbrood and Global Warming"
+subtitle: ""
 author: Edward Cruz, Jr, Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
-format: html
+date: today
+format:
+  html:
+    toc: false
+    number-sections: true
+    highlight-style: github
+bibliography: ../dataanalysis-template-references.bib
+csl: ../apa.csl
 editor: 

### 2024-08-01 starter-analysis-exercise/products/report/starter-analysis-report.qmd
HTML 6

@@
     highlight-style: github
-bibliography: ../dataanalysis-template-references.bib
+bibliography: C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/starter-analysis-exercise/products/report/citations.bib
 csl: ../apa.csl
@@
 honeybee colonies, highlighting the consequent implications for honey
-production and broader food security.
+production and broader food security. BeeHive peered reviewed article
+\[\@hivecollapse\]
 

### 2024-08-01 starter-analysis-exercise/products/report/manuscript_final-2.qmd
HTML 7 - Noter: There is a template built for this report changed the label of the file in index to reflect the group 08/01/24

--- 2024-08-01 starter-analysis-exercise/products/report/starter-analysis-report.qmd
@@
     highlight-style: github
-bibliography: ../dataanalysis-template-references.bib
+bibliography: ../citations.bib
 csl: ../apa.csl
@@
 honeybee colonies, highlighting the consequent implications for honey
-production and broader food security.
+production and broader food security. \[\@hivecollapse\]
 
@@
 
-\*\*\*\*\*\*\*\*\*\*\*\*\*\* Leo provided additional code below and
-works perfectly\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*
-
 ```{r}

### 2024-08-01 starter-analysis-exercise/products/report/manuscript_final-2.qmd
Updated HTML, photos, and references included with citations

@@
 ---
-title: "Honey Bee Colonies Impacted by Varroa, American Fouldbrood and Global Warming"
+title: "Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming"
 subtitle: ""
-author: Edward Cruz, Jr, Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
+author: Edward Cruz, Jr., Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
 date: today
@@
 library(ggimage)
+
 ```
@@
 honeybee colonies, highlighting the consequent implications for honey
-production and broader food security. \[\@hivecollapse\]
+production and broader food security. ***Dennis and Kemp's study on
+honey bee hive collapse [@hivecollapse] provides important insights into
+Allee effects and ecological resilience.***
 
@@
 warmer autumn and winter temperatures on colony dynamics and survival
-require further investigation.
+require further investigation. ***Dennis and Kemp's study on honey bee
+hive collapse [@hivecollapse] provides important insights into Allee
+effects and ecological resilience.***[@usda_honey_bees; @usda_nass]
 
@@
 
+***[@varroamap] Varroa Map of infestation periods***
+
 ```{r}
@@
 
+[@epa_climate_indicator] ***Global warming trends from 1900-2024***
+
+***Spread of Varroa Mite by Year of First Apperance [@varroamap]***
 Written By:sdns6mchl4. (2016, February 24). Varroa mite spread in the
@@
 https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/
+
 \## General Background Information
@@
 
-![](/images/varroa-mite-bee-01.webp){fig-align="center" width="565"}
+![](images/varroa-mite-bee.jpg){fig-align="center"}
 
@@
 
-#Citations
+##The full analysis is still pending###
 

### 2024-08-01 starter-analysis-exercise/products/report/manuscript_final-2.qmd
Another update to include a new reference citation

@@
 Allee effects and ecological resilience.***
+***Warmer climate may impact bees[@2024warmer]***
 
@@
 hive collapse [@hivecollapse] provides important insights into Allee
-effects and ecological resilience.***[@usda_honey_bees; @usda_nass]
+effects and ecological resilience.***[@usda_honey_bees; @usda_nass]***
+***Warmer climate may impact bees[@2024warmer]***
 

### 2024-08-01 starter-analysis-exercise/products/report/manuscript_final-2.qmd
HTML7 Picture adjustments

@@
 honey bee hive collapse [@hivecollapse] provides important insights into
-Allee effects and ecological resilience.***
-***Warmer climate may impact bees[@2024warmer]***
+Allee effects and ecological resilience.*** ***Warmer climate may impact
+bees[@2024warmer]***
 
@@
 hive collapse [@hivecollapse] provides important insights into Allee
-effects and ecological resilience.***[@usda_honey_bees; @usda_nass]***
+effects and ecological resilience.*** [@usda_honey_bees;
+@usda_nass]\*\*\*
+
 ***Warmer climate may impact bees[@2024warmer]***
@@
 
-***[@varroamap] Varroa Map of infestation periods***
+[@varroamap] ***Varroa Map of infestation periods***
 
@@
 health, contributing valuable information on the status and trends of
-bee populations. Index Catalog // USDA Economics, Statistics and Market
-Information System. (n.d.-a).
+bee populations. [@usda_honey_bees]
+
+Index Catalog // USDA Economics, Statistics and Market Information
+System. (n.d.-a).
 https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
@@
 
-![](/images/Bee%20Feeding.jpg){fig-align="center"}
+![](/images/Bee%20Feeding.jpg){fig-align="center" width="313"}
 

### 2024-08-01 starter-analysis-exercise/products/report/manuscript_final-2.qmd
additional updates

@@
 
-[@epa_climate_indicator] ***Global warming trends from 1900-2024***
+[@epa_climate] ***Global warming trends from 1900-2024***
 
@@
 ![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)
-
-#Citation section To cite other work (important everywhere, but likely
-happens first in introduction), make sure your references are in the
-bibtex file specified in the YAML header above and have the right bibtex
-key. Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in
-[@mckay2020; @mckay2020a].
-
-#Here are the BibTex entries for citations:
-
-Using this in RMarkdown: #Make sure to save these entries in a file
-names'reference.bib' and include it in your R Markdown document like
-this:
-
 {{< pagebreak >}}

### 2024-08-02 starter-analysis-exercise/products/report/manuscript_final-2.qmd
update to personal website 08.02.24

@@
 ```{r}
+# Install necessary packages
+#install.packages("DiagrammeR")
 
-#knitr::include_graphics(here("assets","antigen-recognition.png"))
+# Load the DiagrammeR package
+library(DiagrammeR)
+
+# Create the flowchart using DiagrammeR
+grViz("
+  digraph Data_Analysis_Flowchart {
+    graph [layout = dot, rankdir = TB]
+
+    node [shape = box, style = filled, color = lightblue]
+
+    Define_Objectives [label = 'Define Objectives']
+    Collect_Data [label = 'Collect Data']
+    Clean_Data [label = 'Clean Data']
+    Analyze_Data [label = 'Analyze Data']
+    Interpret_Results [label = 'Interpret Results']
+    Visualize_Data [label = 'Visualize Data']
+    Compile_Report [label = 'Compile Report']
+    Review_Revise [label = 'Review & Revise']
+    Present_Report [label = 'Present Report']
+
+    Define_Objectives -> Collect_Data
+    Collect_Data -> Clean_Data
+    Clean_Data -> Analyze_Data
+    Analyze_Data -> Interpret_Results
+    Interpret_Results -> Visualize_Data
+    Visualize_Data -> Compile_Report
+    Compile_Report -> Review_Revise
+    Review_Revise -> Present_Report
+  }
+")
 ```
+
+![](/images/FlowChart.jpg){fig-align="center"}
 
@@
 summary(data_clean)
+
+```
+
+```{r}
 str(data_clean)

### 2024-08-02 starter-analysis-exercise/products/report/manuscript_final-2.qmd
Update to Flow Chart Removed snippit 08.02.24
updated workflow schematic 08.02.24

--- 2024-08-01 starter-analysis-exercise/products/report/manuscript_final-2.qmd
@@
 ```{r}
+# Install necessary packages
+if (!require("DiagrammeR")) install.packages("DiagrammeR")
 
-#knitr::include_graphics(here("assets","antigen-recognition.png"))
+# Load the DiagrammeR package
+library(DiagrammeR)
+
+# Create the flowchart using DiagrammeR
+grViz("
+  digraph Data_Analysis_Flowchart {
+    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
+
+    node [style = filled, fontsize = 20, width = 2, height = 2]
+
+    Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
+    Collect_Data [label = 'Collect Data', shape = box, color = lightgreen]
+    Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]
+    Analyze_Data [label = 'Analyze Data', shape = diamond, color = lightcoral]
+    Interpret_Results [label = 'Interpret Results', shape = trapezium, color = lightgrey]
+    Visualize_Data [label = 'Visualize Data', shape = parallelogram, color = lightpink]
+    Compile_Report [label = 'Compile Report', shape = house, color = lightcyan]
+    Review_Revise [label = 'Review & Revise', shape = invhouse, color = lightgoldenrodyellow]
+    Present_Report [label = 'Present Report', shape = hexagon, color = lightseagreen]
+
+    Define_Objectives -> Collect_Data
+    Collect_Data -> Clean_Data
+    Clean_Data -> Analyze_Data
+    Analyze_Data -> Interpret_Results
+    Interpret_Results -> Visualize_Data
+    Visualize_Data -> Compile_Report
+    Compile_Report -> Review_Revise
+    Review_Revise -> Present_Report
+  }
+")
+
 ```

### 2024-08-02 starter-analysis-exercise/products/report/manuscript_final-2.qmd
fixed the schematic layout

@@
 
-    node [style = filled, fontsize = 20, width = 2, height = 2]
+    node [style = filled, fontsize = 30, width = 2, height = 2]
 
     Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
-    Collect_Data [label = 'Collect Data', shape = box, color = lightgreen]
+    Collect_Data [label = 'Collect Data', shape = box, color = lightgrey]
     Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]

### 2024-08-02 starter-analysis-exercise/products/report/manuscript_final-2.qmd
removed all r -from view

@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load necessary libraries
@@
 
-Sometimes you might want to show a schematic diagram/figure that was not
-created with code (if you can do it with code, do it). @fig-schematic is
-an example of some - completely random/unrelated - schematic that was
-generated with Biorender. We store those figures in the `assets` folder.
-
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Install necessary packages
@@
 
-We got our data from the United States Department of Agriculture (USDA).
+Data was derived from the United States Department of Agriculture (USDA) website, which permits data sets to be downloaded in CSV(comma-separated values) format.
 
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 library(readxl)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Select all columns except 3, 6, and 9
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Output cleaned data file to a csv file.
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Assuming your data frame is named "data"
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # View data str to see what class they are.
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Scatterplot of Value by State.ANSI
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load necessary library
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load your dataset
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Descriptive Statistics
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 str(data_clean)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Visualizations
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Stacked Bar Plot
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Heatmap
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Pie Chart for the United States
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Correlation Analysis
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Assumption Checks for Descriptive Analysis
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Normality check
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Homogeneity of variances
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Cluster Analysis
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Principal Component Analysis (PCA)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 pca_result <- prcomp(data_pca, center = TRUE, scale. = TRUE)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 fviz_pca_var(pca_result, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Support Vector Machine (SVM)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Assuming predictions and data_clean$varroa_mites are already created
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Write the cleaned data to a new CSV file

### 2024-08-03 starter-analysis-exercise/products/report/manuscript_final-2a.qmd
Updated statements and in-depth

--- 2024-08-02 starter-analysis-exercise/products/report/manuscript_final-2.qmd
@@
 ---
-title: "Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming"
-subtitle: ""
-author: Edward Cruz, Jr., Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
-date: today
-format:
-  html:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../citations.bib 
-csl: ../apa.csl
-editor: 
-  markdown: 
-    wrap: 72
+title: "Group 10 Final Project"
+output: html_document
 ---
@@
 honey bee hive collapse [@hivecollapse] provides important insights into
-Allee effects and ecological resilience.*** ***Warmer climate may impact
-bees[@2024warmer]***
+Allee effects and ecological resilience.***
 
@@
 hive collapse [@hivecollapse] provides important insights into Allee
-effects and ecological resilience.*** [@usda_honey_bees;
-@usda_nass]\*\*\*
-
-***Warmer climate may impact bees[@2024warmer]***
-
-### Composition of Honey
-
-The first image is a pie chart that illustrates the composition of honey
-by percentage. The main components are:
-
--   **Fructose**: 38.5%
--   **Glucose**: 31%
--   **Water**: 17.1%
--   **Maltose**: 7.2%
--   **Other components**:
-    -   Sucrose: 1.5%
-    -   Minerals, Vitamins, Enzymes: 0.5%
-    -   Trisaccharides, Carbohydrates: 4.2%
-
-This composition indicates that honey is primarily made up of sugars,
-specifically fructose and glucose, with water being the third major
-component. The presence of minerals, vitamins, and enzymes, though in
-smaller amounts, adds nutritional value to honey. This detailed
-breakdown underscores honey's role as a natural sweetener with
-additional health benefits beyond its primary carbohydrate content.
+effects and ecological resilience.***[@usda_honey_bees; @usda_nass]
 
@@
 ```
+### Composition of Honey
+
+Compositions of honey is a pie chart that illustrates the composition of honey
+by percentage. The main components are:
+
+-   **Fructose**: 38.5%
+-   **Glucose**: 31%
+-   **Water**: 17.1%
+-   **Maltose**: 7.2%
+-   **Other components**:
+    -   Sucrose: 1.5%
+    -   Minerals, Vitamins, Enzymes: 0.5%
+    -   Trisaccharides, Carbohydrates: 4.2%
+
+This composition indicates that honey is primarily made up of sugars,
+specifically fructose and glucose, with water being the third major
+component. The presence of minerals, vitamins, and enzymes, though in
+smaller amounts, adds nutritional value to honey. This detailed
+breakdown underscores honey's role as a natural sweetener with
+additional health benefits beyond its primary carbohydrate content.
 
@@
 
-### Spread of Varroa Mites
-
-The second image is a map illustrating the spread of Varroa mites in the
-United States from 1986 to 1995. The map uses different colors to
-represent the year of the first appearance of Varroa mites in each
-state:
-
--   **1986 - 1987**: Red
--   **1988 - 1989**: Yellow
--   **1990 - 1991**: Green
--   **1992 - 1993**: Blue
--   **1994 - 1995**: Grey
--   **NA**: States where data is not available
-
-The map reveals the progressive spread of Varroa mites from the
-mid-1980s to the mid-1990s, affecting bee colonies across the country.
-The earliest appearances were concentrated in specific regions and
-gradually spread to more states over time. This visual representation
-highlights the increasing geographical distribution of Varroa mites and
-underscores the growing challenge they pose to beekeeping and
-agricultural industries in the United States.
-
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
+### Spread of Varroa Mites
 
-[@varroamap] ***Varroa Map of infestation periods***
+the spread of varroa mites image is a map illustrating the spread of Varroa mites in the
+United States from 1986 to 1995. The map uses different colors to
+represent the year of the first appearance of Varroa mites in each
+state:
+
+-   **1986 - 1987**: Red
+-   **1988 - 1989**: Yellow
+-   **1990 - 1991**: Green
+-   **1992 - 1993**: Blue
+-   **1994 - 1995**: Grey
+-   **NA**: States where data is not available
+
+The map reveals the progressive spread of Varroa mites from the
+mid-1980s to the mid-1990s, affecting bee colonies across the country.
+The earliest appearances were concentrated in specific regions and
+gradually spread to more states over time. This visual representation
+highlights the increasing geographical distribution of Varroa mites and
+underscores the growing challenge they pose to beekeeping and
+agricultural industries in the United States.
+
+***[@varroamap] Varroa Map of infestation periods***
 
@@
 
-[@epa_climate] ***Global warming trends from 1900-2024***
+The chart provided illustrates the changes in temperature anomalies over the years, from 1900 to approximately 2025.
+The temperature anomaly, measured in degrees Celsius (°C), is plotted on the y-axis, while the years are plotted on the x-axis.
+The data is divided into three different measurements: Earth's surface, lower troposphere (RSS), and lower troposphere (UAH), each represented by different colors (blue, red, and green respectively).
+
+### Detailed Description:
+- **Trend Analysis**:
+  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
+  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
+- **Measurement Comparison**:
+  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
+  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
+  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
+- **Anomalies & Peaks**:
+  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
+  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.
+
+### Key Points:
+- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
+- Significant variability is seen in Earth's surface temperatures in the early years.
+- Post-1975, a clear warming trend is observed across all measurements.
+- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
+- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
+- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.
+
+### Impact on Bees, Varroa Mites, and Colony Loss:
+- **Bees**:
+  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
+  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
+  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.
+
+- **Varroa Mites**:
+  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
+  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.
+
+- **Colony Loss**:
+  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
+  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
+  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.
+
+### Conclusion:
+The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health.
+Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.
+
+[@epa_climate_indicator] ***Global warming trends from 1900-2024***
 
@@
 
-Bee colonies maintained by beekeepers are considered livestock by the
-USDA due to their ability to produce honey, a consumable food item, and
-their essential role in assisting farmers with pollination crop seasons.
-Given the importance of bee colonies in agriculture, it was logical to
-source data from the following two authoritative websites: 1. USDA
-National Agricultural Statistics Service (NASS): This site provides
-comprehensive agricultural data, including statistics on honey
-production and colony health. 2. Bee Informed Partnership: This site
-offers detailed insights and research on bee colony management and
-health, contributing valuable information on the status and trends of
-bee populations. [@usda_honey_bees]
+### Data Sources
 
-Index Catalog // USDA Economics, Statistics and Market Information
-System. (n.d.-a).
-https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
-USDA - National Agricultural Statistics Service - Surveys - honey bee
-surveys and reports. (n.d.).
-https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/
+Bee colonies maintained by beekeepers are classified as livestock by the USDA due to their production of honey, a consumable food item, and their critical role in pollination during crop seasons.
+Given the importance of bee colonies in agriculture, we sourced data from two authoritative websites:
+
+1. **USDA National Agricultural Statistics Service (NASS)**:
+   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
+   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).
+
+2. **Bee Informed Partnership**:
+   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
+   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).
+
+These sources provided reliable and extensive data necessary for analyzing the factors affecting bee colonies, honey production, and the broader implications for agriculture.
 
@@
 
-Hypothesis: "The negative impacts of mites, bacterium, and global
-warming have detrimental effects on honeybee colonies in the United
-States and Texas, which in turn will lead to a decline in honey
-production and negatively impact food production." This hypothesis can
-be tested and validated through a visualization of outcomes using R,
-demonstrating the relationship between these factors and their effects
-on honeybee colonies.
+### Hypotheses
+
+**Hypothesis**:
+"The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.
+
+**Null Hypothesis (H0)**:
+The impacts of mites, bacterium, and global warming do not have a significant detrimental effect on honeybee colonies in the United States and Texas, and there is no consequent decline in honey production or negative impact on food production.
+
+**Alternative Hypothesis (H1)**:
+The impacts of mites, bacterium, and global warming have a significant detrimental effect on honeybee colonies in the United States and Texas, leading to a decline in honey production and negatively impacting food production.
+
+These hypotheses will be tested using various data visualization techniques in R, allowing us to explore and validate the relationships between these factors and their effects on honeybee colonies.
 
@@
 ![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)
+
+#Citation section To cite other work (important everywhere, but likely
+happens first in introduction), make sure your references are in the
+bibtex file specified in the YAML header above and have the right bibtex
+key. Then you can include like this:
+
+Examples of reproducible research projects can for instance be found in
+[@mckay2020; @mckay2020a].
+
+#Here are the BibTex entries for citations:
+
+Using this in RMarkdown: #Make sure to save these entries in a file
+names'reference.bib' and include it in your R Markdown document like
+this:
+
 {{< pagebreak >}}
 
-# Methods
-
-*Describe your methods. That should describe the data, the cleaning
-processes, and the analysis approaches. You might want to provide a
-shorter description here and all the details in the supplement.*
+# Methodology
 
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# Install necessary packages
-if (!require("DiagrammeR")) install.packages("DiagrammeR")
-
-# Load the DiagrammeR package
 library(DiagrammeR)
 
-# Create the flowchart using DiagrammeR
-grViz("
-  digraph Data_Analysis_Flowchart {
-    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
+DiagrammeR::grViz("
+digraph G {
 
-    node [style = filled, fontsize = 30, width = 2, height = 2]
+  node [shape = box, style = filled, color = lightblue, fontname = Helvetica]
 
-    Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
-    Collect_Data [label = 'Collect Data', shape = box, color = lightgrey]
-    Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]
-    Analyze_Data [label = 'Analyze Data', shape = diamond, color = lightcoral]
-    Interpret_Results [label = 'Interpret Results', shape = trapezium, color = lightgrey]
-    Visualize_Data [label = 'Visualize Data', shape = parallelogram, color = lightpink]
-    Compile_Report [label = 'Compile Report', shape = house, color = lightcyan]
-    Review_Revise [label = 'Review & Revise', shape = invhouse, color = lightgoldenrodyellow]
-    Present_Report [label = 'Present Report', shape = hexagon, color = lightseagreen]
+  subgraph cluster_0 {
+    label = 'Analytical Report Workflow'
 
-    Define_Objectives -> Collect_Data
-    Collect_Data -> Clean_Data
-    Clean_Data -> Analyze_Data
-    Analyze_Data -> Interpret_Results
-    Interpret_Results -> Visualize_Data
-    Visualize_Data -> Compile_Report
-    Compile_Report -> Review_Revise
-    Review_Revise -> Present_Report
+    Introduction -> Data_Acquisition -> Data_Cleaning_Preparation -> Exploratory_Data_Analysis -> Data_Visualization -> Advanced_Statistical_Analysis -> Results_Discussion -> Conclusion -> Appendix -> References
+  }
+
+  Introduction [label = '1. Introduction']
+  Data_Acquisition [label = '2. Data Acquisition']
+  Data_Cleaning_Preparation [label = '3. Data Cleaning and Preparation']
+  Exploratory_Data_Analysis [label = '4. Exploratory Data Analysis (EDA)']
+  Data_Visualization [label = '5. Data Visualization']
+  Advanced_Statistical_Analysis [label = '6. Advanced Statistical Analysis']
+  Results_Discussion [label = '7. Results and Discussion']
+  Conclusion [label = '8. Conclusion']
+  Appendix [label = '9. Appendix']
+  References [label = '10. References']
 }
 ")
+```
 
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+
+#knitr::include_graphics(here("assets","antigen-recognition.png"))
 ```
@@
 
-Data was derived from the United States Department of Agriculture (USDA) website, which permits data sets to be downloaded in CSV(comma-separated values) format.
+### Data Acquisition
+
+We sourced our data from the United States Department of Agriculture (USDA), leveraging their comprehensive and authoritative datasets.
+This data provided crucial information for our analysis, including statistics on honey production, colony health, and factors affecting bee populations.
+Utilizing USDA data ensured the reliability and relevance of our analysis, as it encompasses extensive, well-documented records pertinent to agricultural and ecological studies.
 
@@
 
-We decided to clean out our data from a few different datasets. We had
-to remove blank spaces and columns that were not pertinent to our
-analysis. We then filtered out other observations that did not directly
-deal with the data we are exploring. We are looking for cause of death
-to bee colonies and how they are affected by mites and climate change so
-we wanted to single out data that represented the losses so we can
-explore the different states by year and determine how the colonies were
-affected.
+### Data Cleaning and Preparation
+
+We integrated data from multiple datasets to ensure a comprehensive analysis. The cleaning process involved several key steps:
+
+1. **Removal of Irrelevant Data**:
+   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
+   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.
+
+2. **Filtering Observations**:
+   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
+   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.
+
+3. **Exploratory Focus**:
+   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.
+
+These steps ensured that our dataset was clean, relevant, and well-prepared for subsequent analysis, enabling a focused exploration of the causes of bee colony losses and their relationship with mites and climate change.
 
@@
 
-here()
 # Create a 'data' directory within the project
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# View data str to see what class they are.
-
-str(data_filtered)
-```
-
-### Boxplot of Value by Year and Data Item
-
-The third image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" from 2015 to 2017. The boxplot
-provides insights into: - The central tendency and spread of losses due
-to both disorders. - Presence of outliers indicates that there are
-occasional, significantly higher losses. - Comparative analysis shows
-that both types of losses have varying distributions each year.
-
-This indicates that while both types of losses are present each year,
-the magnitude and frequency of extreme loss events can differ,
-suggesting variability in the factors contributing to colony losses.
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 library(here)
-
-str(data_filtered)
 
@@
 
-### Scatterplot of Value by State ANSI
+### Analysis of Boxplot
 
-The fourth image is a scatterplot showing values for different states
-(ANSI coded) across the years 2015 to 2017. Key points include: - Each
-color represents a different year (2015: Red, 2016: Green, 2017:
-Blue). - The scatterplot reveals the distribution of values across
-states and highlights any trends or patterns over the three years.
+The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across the years 2015 to 2017. A few key observations can be made from the data:
 
-This scatterplot indicates that while some states experience
-consistently high or low values, others show significant variability
-year over year. This could suggest regional differences in factors
-affecting bee colony health.
+**Central Tendency and Spread**:
+   - The median values for both types of losses (represented by the central line in each box) are relatively low across all years, indicating that the majority of losses are concentrated towards the lower end of the scale.
+
+**Yearly Comparison**:
+   - For the year 2015, the losses due to Colony Collapse Disorder and Deadout are higher compared to the subsequent years.
+This can be observed from the height of the boxes and the spread of the data points.
+   - There is a noticeable decrease in the losses for both categories in 2016 and 2017. The boxes for these years are smaller and closer to the x-axis, indicating lower values.
+
+**Outliers**:
+   - There are significant outliers present in all years for both types of losses.
+These outliers indicate that while most of the data points are clustered around the lower values, there are instances of very high losses that deviate from the norm.
+
+**Comparison between Loss Types**:
+   - The distribution of "LOSS, DEADOUT" appears to have more variability compared to "LOSS, COLONY COLLAPSE DISORDER", particularly in 2015 and 2016.
+This is indicated by the wider interquartile ranges and more spread out data points.
+
+**Overall Trends**:
+   - Overall, the boxplot suggests a decline in both types of losses from 2015 to 2017. However, the presence of outliers in each year indicates that there are still occasional severe loss events.
+
+This highlights the trends and variations in honeybee colony losses due to Colony Collapse Disorder and Deadout over a three-year period, providing insights into the distribution and severity of these losses.
 
@@
   geom_point() +
+  scale_y_continuous(labels = comma) +
   labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
@@
 
-### Boxplot of Value by Period and Data Item
+### Analysis of Scatterplot
 
-The fifth image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
-year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
-THRU DEC**
+The scatterplot visualizes the distribution of values across different states (represented by State ANSI codes) for the years 2015, 2016, and 2017. Several key observations can be drawn from the data:
 
-Key insights include: - Seasonal trends in colony losses, with some
-periods showing higher median losses or more variability. - Outliers
-indicate extreme loss events occurring in specific periods.
+**Value Distribution Across States**:
+   - The values vary significantly across different states. Most states have relatively low values, but there are notable exceptions with very high values.
 
-This seasonal analysis suggests that certain times of the year are more
-prone to higher losses, potentially due to environmental factors,
-resource availability, or management practices.
+**Yearly Trends**:
+   - In 2015, there are multiple high values, especially at the lower end of the State ANSI codes. These are represented by the red dots.
+   - In 2016, represented by green dots, there are fewer high values, but some extreme outliers are present, such as the one exceeding 200,000.
+   - In 2017, indicated by blue dots, the values are generally lower compared to the previous years, with fewer extreme outliers.
+
+**Outliers**:
+   - Each year has significant outliers that deviate from the general trend of low values across states. These outliers suggest occasional severe losses in specific states.
+
+**State-Specific Trends**:
+   - Certain states (e.g., those with lower ANSI codes) exhibit higher variability and more frequent high values compared to others.
+   - States towards the middle and higher end of the ANSI scale generally show more consistent, lower values.
+
+**Comparative Analysis**:
+   - The plot indicates that while overall losses may be high in certain states, the distribution and frequency of these high values vary across years.
+This suggests possible changes in factors affecting bee colony losses over time.
+
+This shows the distribution and variability of bee colony losses across different states over three years, providing insights into state-specific trends and the presence of extreme loss events.
 
@@
 
-### Boxplot of Value by Period and Data Item
+### Analysis of Boxplot
 
-The fifth image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
-year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
-THRU DEC**
+The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the year.
+The data is segmented into four periods: APR-THRU-JUN, JAN-THRU-MAR, JUL-THRU-SEP, and OCT-THRU-DEC.
+Several key observations can be made from the data:
 
-Key insights include: - Seasonal trends in colony losses, with some
-periods showing higher median losses or more variability. - Outliers
-indicate extreme loss events occurring in specific periods.
+**Central Tendency and Spread**:
+   - The median values for both types of losses are low across all periods, indicating that most losses are concentrated towards the lower end of the scale.
+   - The interquartile ranges (IQRs) for "LOSS, DEADOUT" are relatively consistent across periods, suggesting stable variability in losses throughout the year.
 
-This seasonal analysis suggests that certain times of the year are more
-prone to higher losses, potentially due to environmental factors,
-resource availability, or management practices.
+**Period Comparison**:
+   - The period JUL-THRU-SEP shows the highest median value for "LOSS, DEADOUT," indicating that this period might be particularly challenging for bee colonies.
+   - The period JAN-THRU-MAR has a few outliers but generally shows lower losses compared to other periods.
+
+**Outliers**:
+   - Significant outliers are present in all periods for both types of losses. These outliers suggest occasional severe loss events that deviate significantly from the norm.
+   - The period APR-THRU-JUN shows the highest outlier for "LOSS, COLONY COLLAPSE DISORDER," indicating some extreme events during this time.
+
+**Comparison between Loss Types**:
+   - "LOSS, DEADOUT" consistently shows a wider range of values compared to "LOSS, COLONY COLLAPSE DISORDER" across all periods, suggesting more variability in deadout losses.
+
+**Overall Trends**:
+   - The plot suggests that while losses occur throughout the year, there are specific periods (e.g., JUL-THRU-SEP) where losses might be higher.
+   - The consistent presence of outliers indicates that, despite general trends, extreme loss events are a recurring issue.
+
+This boxplot shows insights into the seasonal variability of bee colony losses due to Colony Collapse Disorder and Deadout, highlighting specific periods with higher losses and the presence of extreme loss events.
 
@@
 summary(data_clean)
-
-```
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
 str(data_clean)
@@
 
-![](/images/Bee%20Feeding.jpg){fig-align="center" width="313"}
+![](/images/Bee%20Feeding.jpg){fig-align="center"}
 
@@
 ```
-
-## Full analysis
-
-##The full analysis is still pending###
-
-{{< pagebreak >}}
-
-# References

### 2024-08-03 starter-analysis-exercise/products/report/manuscript_final-2.qmd
Update format

@@
 
-### Composition of Honey
-
-The first image is a pie chart that illustrates the composition of honey
-by percentage. The main components are:
-
--   **Fructose**: 38.5%
--   **Glucose**: 31%
--   **Water**: 17.1%
--   **Maltose**: 7.2%
--   **Other components**:
-    -   Sucrose: 1.5%
-    -   Minerals, Vitamins, Enzymes: 0.5%
-    -   Trisaccharides, Carbohydrates: 4.2%
-
-This composition indicates that honey is primarily made up of sugars,
-specifically fructose and glucose, with water being the third major
-component. The presence of minerals, vitamins, and enzymes, though in
-smaller amounts, adds nutritional value to honey. This detailed
-breakdown underscores honey's role as a natural sweetener with
-additional health benefits beyond its primary carbohydrate content.
-
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
+### Composition of Honey
+
+The first image is a pie chart that illustrates the composition of honey
+by percentage. The main components are:
+
+-   **Fructose**: 38.5%
+-   **Glucose**: 31%
+-   **Water**: 17.1%
+-   **Maltose**: 7.2%
+-   **Other components**:
+    -   Sucrose: 1.5%
+    -   Minerals, Vitamins, Enzymes: 0.5%
+    -   Trisaccharides, Carbohydrates: 4.2%
+
+This composition indicates that honey is primarily made up of sugars,
+specifically fructose and glucose, with water being the third major
+component. The presence of minerals, vitamins, and enzymes, though in
+smaller amounts, adds nutritional value to honey. This detailed
+breakdown underscores honey's role as a natural sweetener with
+additional health benefits beyond its primary carbohydrate content.
 
@@
 
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+ggplot(data = states) +
+  geom_sf(aes(fill = year), color = "#292929", size = 0.2) +
+  scale_fill_manual(
+    values = c(
+      "1986 - 1987" = "#CD2626", 
+      "1988 - 1989" = "#CD9B1D", 
+      "1990 - 1991" = "#FFD700", 
+      "1992 - 1993" = "#00CD00", 
+      "1994 - 1995" = "#104E8B"
+    ),
+    na.value = "#757575"
+  ) +
+  theme_minimal() +
+  labs(
+    title = "Spread of Varroa Mites",
+    fill = "Year of First Appearance"
+  )
+
+```
 ### Spread of Varroa Mites
@@
 agricultural industries in the United States.
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-ggplot(data = states) +
-  geom_sf(aes(fill = year), color = "#292929", size = 0.2) +
-  scale_fill_manual(
-    values = c(
-      "1986 - 1987" = "#CD2626", 
-      "1988 - 1989" = "#CD9B1D", 
-      "1990 - 1991" = "#FFD700", 
-      "1992 - 1993" = "#00CD00", 
-      "1994 - 1995" = "#104E8B"
-    ),
-    na.value = "#757575"
-  ) +
-  theme_minimal() +
-  labs(
-    title = "Spread of Varroa Mites",
-    fill = "Year of First Appearance"
-  )
-
-```
-
-[@varroamap] ***Varroa Map of infestation periods***
+***[@varroamap] Varroa Map of infestation periods***
 
@@
 ```
+The chart provided illustrates the changes in temperature anomalies over the years, from 1900 to approximately 2025.
+The temperature anomaly, measured in degrees Celsius (°C), is plotted on the y-axis, while the years are plotted on the x-axis.
+The data is divided into three different measurements: Earth's surface, lower troposphere (RSS), and lower troposphere (UAH), each represented by different colors (blue, red, and green respectively).
+
+### Detailed Description:
+- **Trend Analysis**:
+  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
+  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
+- **Measurement Comparison**:
+  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
+  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
+  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
+- **Anomalies & Peaks**:
+  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
+  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.
+
+### Key Points:
+- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
+- Significant variability is seen in Earth's surface temperatures in the early years.
+- Post-1975, a clear warming trend is observed across all measurements.
+- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
+- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
+- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.
+
+### Impact on Bees, Varroa Mites, and Colony Loss:
+- **Bees**:
+  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
+  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
+  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.
+
+- **Varroa Mites**:
+  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
+  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.
+
+- **Colony Loss**:
+  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
+  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
+  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.
+
+### Conclusion:
+The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health.
+Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.
 
@@
 
-Bee colonies maintained by beekeepers are considered livestock by the
-USDA due to their ability to produce honey, a consumable food item, and
-their essential role in assisting farmers with pollination crop seasons.
-Given the importance of bee colonies in agriculture, it was logical to
-source data from the following two authoritative websites: 1. USDA
-National Agricultural Statistics Service (NASS): This site provides
-comprehensive agricultural data, including statistics on honey
-production and colony health. 2. Bee Informed Partnership: This site
-offers detailed insights and research on bee colony management and
-health, contributing valuable information on the status and trends of
-bee populations. [@usda_honey_bees]
+### Data Sources
 
-Index Catalog // USDA Economics, Statistics and Market Information
-System. (n.d.-a).
-https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
-USDA - National Agricultural Statistics Service - Surveys - honey bee
-surveys and reports. (n.d.).
-https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/
+Bee colonies maintained by beekeepers are classified as livestock by the USDA due to their production of honey, a consumable food item, and their critical role in pollination during crop seasons.
+Given the importance of bee colonies in agriculture, we sourced data from two authoritative websites:
+
+1. **USDA National Agricultural Statistics Service (NASS)**:
+   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
+   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).
+
+2. **Bee Informed Partnership**:
+   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
+   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).
+
+These sources provided reliable and extensive data necessary for analyzing the factors affecting bee colonies, honey production, and the broader implications for agriculture.
 
 ## Questions/Hypotheses to be addressed
+**Hypothesis**:
+"The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.
 
-Hypothesis: "The negative impacts of mites, bacterium, and global
-warming have detrimental effects on honeybee colonies in the United
-States and Texas, which in turn will lead to a decline in honey
-production and negatively impact food production." This hypothesis can
-be tested and validated through a visualization of outcomes using R,
-demonstrating the relationship between these factors and their effects
-on honeybee colonies.
+**Null Hypothesis (H0)**:
+The impacts of mites, bacterium, and global warming do not have a significant detrimental effect on honeybee colonies in the United States and Texas, and there is no consequent decline in honey production or negative impact on food production.
+
+**Alternative Hypothesis (H1)**:
+The impacts of mites, bacterium, and global warming have a significant detrimental effect on honeybee colonies in the United States and Texas, leading to a decline in honey production and negatively impacting food production.
+
+These hypotheses will be tested using various data visualization techniques in R, allowing us to explore and validate the relationships between these factors and their effects on honeybee colonies.
 
@@
 ![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)
+
+#Here are the BibTex entries for citations:
+
+@Article{hivecollapse,
+  author = {Brian Dennis and William P. Kemp},
+  title = {How Hives Collapse: Allee Effects, Ecological Resilience, and the Honey Bee},
+  journal = {PLOS ONE},
+  year = {2016},
+  volume = {11},
+  number = {2},
+  pages = {e0150055},
+  doi = {10.1371/journal.pone.0150055},
+  url = {https://doi.org/10.1371/journal.pone.0150055}
+A study performed to determine various elements impacting bee hive collapse}
+
+@USDAWebsite{varroamap,
+  author = {sdns6mchl4},
+  title = {Varroa mite spread in the United States},
+  year = {2016},
+  month = {February 24},
+  url = {https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/},
+  note = {Beesource Beekeeping Forums}
+Map highlighting states of Varroa infestation from periods 1986-1995}
+
+@USDAWebsite{usda_honey_bees,
+  author = {{USDA Economics, Statistics and Market Information System}},
+  title = {Index Catalog},
+  year = {n.d.-a},
+  url = {https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en},
+  note = {Accessed: 2024-07-31}
+Data sets}
+
+@USDAWebsite{usda_nass,
+  author = {{USDA - National Agricultural Statistics Service}},
+  title = {Surveys - honey bee surveys and reports},
+  year = {n.d.},
+  url = {https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/},
+  note = {Accessed: 2024-07-31}
+Data sets}
+
+@EPAWebsite{epa_climate,
+  author = {{U.S. Environmental Protection Agency}},
+  title = {Climate Change Indicators: U.S. and Global Temperature},
+  year = {2024},
+  howpublished = {\url{https://www.epa.gov/climate-indicators/climate-change-indicators-us-and-global-temperature}},
+  note = {Accessed: 2024-08-01}
+Climate change from periods of 1900-2024}
+
+@article{2024warmer,
+  title={Warmer autumns and winters could reduce honey bee overwintering survival with potential risks for pollination services},
+  author={Rajagopalan, K. and DeGrandi-Hoffman, G. and Pruett, M. and others},
+  journal={Scientific Reports},
+  volume={14},
+  pages={5410},
+  year={2024},
+  publisher={Nature Publishing Group},
+  doi={10.1038/s41598-024-55327-8}
+Peer reviewed article concering global warming possibly impacting bee hive survivability and viability during winter periods}
+
 {{< pagebreak >}}
 
-# Methods
+# Methodology
 
-*Describe your methods. That should describe the data, the cleaning
-processes, and the analysis approaches. You might want to provide a
-shorter description here and all the details in the supplement.*
+## Data aquisition
+
+The United States Department of Agriculture (USDA) website served as our data source.
+We downloaded CSV files containing vital information on honey production, colony health, and factors impacting bee populations.
+By using USDA data, we ensured our analysis was based on reliable, comprehensive, and relevant records widely recognized in agricultural and ecological research.
+
+## Data import and cleaning
+
+### Data Cleaning and Preparation
+We integrated data from multiple datasets to ensure a comprehensive analysis. The cleaning process involved several key steps:
+
+1. **Removal of Irrelevant Data**:
+   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
+   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.
+
+2. **Filtering Observations**:
+   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
+   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.
+
+3. **Exploratory Focus**:
+   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.
+
+These steps ensured that our dataset was clean, relevant, and well-prepared for subsequent analysis, enabling a focused exploration of the causes of bee colony losses and their relationship with mites and climate change.
 
@@
   digraph Data_Analysis_Flowchart {
-    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
+    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 75, labelloc = 't']
 
-    node [style = filled, fontsize = 30, width = 2, height = 2]
+    node [style = filled, fontsize = 50, width = 6, height = 6, penwidth = 8]
 
@@
 ```
-
-## Data aquisition
-
-Data was derived from the United States Department of Agriculture (USDA) website, which permits data sets to be downloaded in CSV(comma-separated values) format.
-
-## Data import and cleaning
-
-We decided to clean out our data from a few different datasets. We had
-to remove blank spaces and columns that were not pertinent to our
-analysis. We then filtered out other observations that did not directly
-deal with the data we are exploring. We are looking for cause of death
-to bee colonies and how they are affected by mites and climate change so
-we wanted to single out data that represented the losses so we can
-explore the different states by year and determine how the colonies were
-affected.
 
@@
 ```
+### Analysis of Scatterplot
 
-### Boxplot of Value by Period and Data Item
+The scatterplot visualizes the distribution of values across different states (represented by State ANSI codes) for the years 2015, 2016, and 2017. Several key observations can be drawn from the data:
 
-The fifth image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
-year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
-THRU DEC**
+**Value Distribution Across States**:
+   - The values vary significantly across different states. Most states have relatively low values, but there are notable exceptions with very high values.
 
-Key insights include: - Seasonal trends in colony losses, with some
-periods showing higher median losses or more variability. - Outliers
-indicate extreme loss events occurring in specific periods.
+**Yearly Trends**:
+   - In 2015, there are multiple high values, especially at the lower end of the State ANSI codes. These are represented by the red dots.
+   - In 2016, represented by green dots, there are fewer high values, but some extreme outliers are present, such as the one exceeding 200,000.
+   - In 2017, indicated by blue dots, the values are generally lower compared to the previous years, with fewer extreme outliers.
 
-This seasonal analysis suggests that certain times of the year are more
-prone to higher losses, potentially due to environmental factors,
-resource availability, or management practices.
+**Outliers**:
+   - Each year has significant outliers that deviate from the general trend of low values across states. These outliers suggest occasional severe losses in specific states.
+
+**State-Specific Trends**:
+   - Certain states (e.g., those with lower ANSI codes) exhibit higher variability and more frequent high values compared to others.
+   - States towards the middle and higher end of the ANSI scale generally show more consistent, lower values.
+
+**Comparative Analysis**:
+   - The plot indicates that while overall losses may be high in certain states, the distribution and frequency of these high values vary across years.
+This suggests possible changes in factors affecting bee colony losses over time.
+
+This shows the distribution and variability of bee colony losses across different states over three years, providing insights into state-specific trends and the presence of extreme loss events.
 
@@
 
-##The full analysis is still pending###
+The full analysis is still pending
 
@@
 
-# References
+## References
